### 2019-06-13 paper/transfer_learning.tex
Add paper files

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amsmath, amssymb}
\usepackage{multirow}
\usepackage{booktabs}

% Fixup hyphenation
\hyphenation{cor-res-pon-din}
\hyphenation{cor-re-lat-tion}
\hyphenation{compa-rison}
\hyphenation{compa-risons}
\hyphenation{ini-tial-iza-tion}
\hyphenation{in-clud-ing}

\newcommand{\etal}{\textit{et al.}}

\begin{document}
%
\title{Embedding to Reference t-SNE Space Adresses Batch-Effects in Single-Cell Classification}
%
\titlerunning{t-SNE Embedding and Batch Effects}
%
\author{Pavlin G. Poli\v{c}ar\inst{1} \and
Martin Stra\v{z}ar\inst{1} \and
Bla\v{z} Zupan\inst{1,2}}
%
\authorrunning{P. G. Poli\v{c}ar, M. Stra\v{z}ar, and B. Zupan}
%
\institute{University of Ljubljana, SI-1000 Ljubljana, Slovenia 
\email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
\and
Baylor College of Medicine, Houston, TX 77030, USA
}

\maketitle

\begin{abstract}
Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data. When working with multiple data sets, the straightforward application of these methods often fails, and instead of revealing underlying classes, the resulting visualizations expose domain-specific clusters. To circumvent these batch effects, we here propose an embedding procedure that takes a t-SNE visualization constructed on reference data set and uses it as a scaffold for new embeddings. New, secondary data is embedded one data-instance at the time, thus disregarding any interactions between data items in the secondary data and implicitly mitigating batch effects. We demonstrate the utility of this procedure on analysis of six recently published single-cell gene expression data sets containing up to tens of thousands of cells and thousands of genes. In these data sets, the batch effects were particularly strong as the data comes from different institutions and was obtained using different experimental protocols. The visualizations constructed by the proposed approach are absent of batch effects, and the cells from secondary data sets correctly co-cluster with cells from the primary data sharing the same cell type.

\keywords{Batch effects \and Embedding \and t-SNE \and Visualisation \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
\end{abstract}

\section{Introduction}

Two-dimensional embeddings and their visualizations may assist in the analysis and interpretation of multi-dimensional data. Intuitively, two data instances should be co-located in the resulting visualization if their multi-dimensional profiles are similar. For this task, non-linear local embeddings such as t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform manifold approximation and projection~\cite{umap} have recently complemented traditional data transformation and embedding approaches such as principal component analysis (PCA) and multi-dimensional scaling~\cite{distill,umap_single_cell}. While useful for visualizing the data from a single coherent source, these methods may encounter problems if the data comes from different sources. Here, if executing the dimensionality reduction procedures on a merged data, the resulting visualizations would typically depict source-specific clusters instead of reveal clusters that are semantically enriched, say, with same-type of data instances across different sources. This source-specific confounding is often referred to as {\em domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or {\em dataset shift}~\cite{dataset_shift}. In the bioinformatics literature, the domain-specific differences are more commonly referred to as {\em batch effects}~\cite{cca,mnn,seurat}.

Massive, multi-variate biological data sets typical suffer from source-specific biases. Consider an example from single-cell genomics, a domain we will focus on in this manuscript and that was --- besides current scientific challenges --- selected also due to availability and the abundance of recently published data. Single-cell data sets may include thousands of cells whose profiles consist of tens of thousands of cell-specific gene expressions. Single cell studies typically start with the analysis of cell types, where we expect that cells of the same type would cluster in two-dimensional data visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells originating from visual cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also separates the data from the two brain regions. These two regions share the same cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two studies to overlap. Batch effects similarly prohibit the utility of t-SNE in the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from human cell atlas~\cite{baron2016} and similarly-typed cells from diabetic patients~\cite{xin2016}. Just like with data from brain cells, pancreas cells cluster first according to the data source, again resulting in an uninformative visualization that primarily exposes the batch effect.

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/batch_effect.pdf}
\caption{Batch effects are a driving factor of variation between the data sets. Depicted is a t-SNE visualisation of two pairs of data sets. In each pair, the data sets share cell types, so it would be expected that the cells from the reference data (blue) would mix with the cells in a secondary data sets (orange). Instead, t-SNE visualisation clusters data according to the data source.} \label{fig:batch_effect}
\end{figure}

Current solutions to embedding the data from various data sources address the batch effect problems up-front. The data is typically pre-processed and transformed, possibly to the latent space, so that the batch effects are removed. The recently proposed procedures for batch-effect removal include canonical correlation analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of batch effect removal is a good mixing of cells from different sources in a t-SNE visualization. Elimination of batch effects may require aggressive data preprocessing which may blur the boundaries between cell types. Another problem is also the inclusion of any new data, for which the entire data analysis pipeline has to be rerun, usually resulting in different layout and clusters that have little resemblance to original visualization and thus require reinterpretation.

We propose a direct solution of rendering t-SNE visualizations that addresses batch effects. Our approach considers one of the data sets as a {\em reference} and aims to visually classify the cells in the other, {\em secondary data set}. We construct the t-SNE embedding using the reference data set, and then use it as a scaffold for the embedding of data points from the secondary data. Embedding is performed one data point at a time. Independence of each new embedding of data instances from secondary data set causes clustering landscape to only depend on a reference scaffold, thus removing data source-driven variation. In other words, when including new data, the scaffold inferred from the reference data set is kept unchanged and defines the ``gravitational field'' to independently embedded each new data item. For example, in Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t-SNE visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are dispersed across entire embedding space and their cell type correctly matches the prevailing type in reference clusters.

At the core of the proposed solution is a mapping that implements embedding of new data to existing t-SNE visualization. In the following, we introduce t-SNE, then describe its recently proposed multi-scale extension, and outline our algorithm that embeds new data into a previously trained scaffold. While a utility of such an algorithm was already hinted to in recent publication~\cite{art_of_using_tsne}, we here provide its practical and theoretically-grounded implementation. Considering the abundance of recent publications in batch effect removal, we present surprising evidence that a computationally more direct and elegant embedding procedure solves the batch effects problem when constructing interpretable visualizations from different data sources.

\section{Methods}

We describe an end-to-end pipeline that uses t-SNE embeddings as a scaffold for new data samples, and enables visualisation of data from different sources while mitigating batch effects. Our proposed approach starts with t\nobreakdash -SNE to embedd a reference data set, with the aim of constructing a two-dimensional visualisation to facilitate interpretation and cluster classification. We regard the resulting two-dimensional embedding as a scaffold to which we add samples from the new, secondary data source. Each new sample is then placed into the reference embedding and optimized independently using the t\nobreakdash -SNE loss function. Independent embedding of each data item from a secondary data set disregards any interactions present in that data, and prevents forming clusters that would be data-souce specific. Below, we start with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing the notation and relevant parts upon which we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).

\subsection{Data embedding by t-SNE and its extensions\label{sec:tsne}}

The embedding by t-SNE is local and non-linear, tailored to visualisation of high dimensional data sets. Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of samples in the reference data set, t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used as a visualization tool, $d$ is typically set to two. The similarity between two data points in t-SNE is defined as:

\begin{equation}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
{\sum_{k \neq \i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
\label{eq:gaussian_kernel}
\end{equation}

\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to

\begin{equation}
p_{ij} = \frac{p_{j \mid i} + p_{i \mid j}}{2N}.
\label{eq:symmetrize}
\end{equation}

The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the perplexity of the distribution matches a user-specified parameter value

\begin{equation}
\text{Perplexity} = 2^{H(P_i)}
\end{equation}

\noindent where $H(P_i)$ is the Shannon entropy of $P_i$,

\begin{equation}
H(P_i) = -\sum_i p_{j \mid i} \log_2 (p_{j \mid i}).
\end{equation}

\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying density of the data in the multi-dimensional space.

The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the embedding space are defined using the $t$-distribution with one degree of freedom

\begin{equation}
q_{ij} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{y}_j ||^2 \right )^{-1}}
{\sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right )^{-1}},
\quad q_{ii} = 0.
\label{eq:cauchy_kernel}
\end{equation}

The data transformation by t-SNE finds an embedding $\mathbf{Y}$ that minimizes the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,

\begin{equation}
C = \text{KL}(\mathbf{P} \mid \mid \mathbf{Q}) = \sum_{ij} p_{ij} \log \frac{p_{ij}}{q_{ij}}.
\label{eq:kl_divergence}
\end{equation}

The time complexity needed to evaluate the similarities in Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of the algorithm impractical for any reasonably-sized data. To address larger data sets, we instead adapt a recent approach for low-rank approximation of similarity matrix based on polynomial interpolation that reduces the time complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the visualization of massive data sets possibly containing millions of data points~\cite{fi_tsne}.

The t-SNE embeddings substantially depend on the value of the perplexity parameter. The perplexity parameter can be interpreted as the number of neighbors for which the distances in the embedding space are preserved. Small values of perplexity result in tightly-packed clusters of points and lead to ignoring the long-range interactions between clusters. Larger values may result in a more globally consistent visualisations, preserving distances on a large scale and organizing clusters in a more meaningful way. Larger values of perplexity may often result in merging multiple small clusters, thus obscuring local aspects of the data~\cite{art_of_using_tsne}.

A trade-off between the local organization and global consistency may be achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}. Multi-scale kernels are defined as

\begin{equation}
p_{j \mid i} \propto \frac{1}{L} \sum_{l=1}^{L} \exp \left ( - \frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_{i,l}^2 \right ), \quad p_{i \mid i} = 0
\label{eq:multiscale}
\end{equation}

\noindent where $L$ is the number of mixture components. The bandwidths $\sigma_{i,l}$ are selected in the same manner as in Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each $l$. In our experiments, we used a mixture of two Gaussian kernels with perplexity values of 50 and 500. We note that a similar formulation of multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the resulting embeddings are visually very similar to those obtained with the approach described above (for brevity, data not shown).

\subsection{Adding new data points to reference embedding\label{sec:transfer}}

An algorithm to embedd new data point to a reference embedding consists of estimating similarities between the point and the reference data and optimizing the position of the data point in the embedding space. Unlike parametric models such as principal component analysis or autoencoders, t-SNE does not define an explicit mapping to embedding space, and embeddings need to be found through loss function optimization.

For each new data point, we first estimate its distance to each of the data instances in the reference data set. We initialize the position of the data point in embedding space to the median embedding position of its $k$ nearest neighbors from the reference data set. While we found the algorithm to be robust to choices of $k$, we use $k=25$ in our experiments.

We adapt the standard t-SNE formulation from Eqs.~\ref{eq:gaussian_kernel}~and~\ref{eq:cauchy_kernel} with

\begin{align}
p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} d(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
\end{align}

\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots, \mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots, \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the symmetrization step in Eq.~\ref{eq:symmetrize}. This enables new points to be inserted into the embedding independently of one another. The gradients of $\mathbf{w}_j$ with respect to the loss (Eq.~\ref{eq:kl_divergence}) are:

\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}
\label{eq:gradient}
\end{equation}

In the optimization step, we refine point positions using batch gradient descent. We use an adaptive learning rate scheme with momentum proposed by Jacobs to speed up the convergence~\cite{momentum}. We run gradient descent with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the optimization converged for all cases in our experiments.  The time complexity needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$, however, by adapting the same polynomial interpolation based approximation, this is reduced to $\mathcal{O}(N)$.

Special care must be taken to reduce the learning rate $\eta$ as the default value in most implementations ($\eta = 200$) may cause points to ``shoot off'' from the reference embedding. This phenomenon is caused due to the embedding to already defined t-SNE space, where the distances between data points and corresponding gradients of the optimization function may be quite large. When running s standard t-SNE, points are initialized and scaled to have variance 0.0001. The resulting gradients tend to be very small during the initial phase, resulting in stable convergence. When embedding new samples, the range of the embedding is much larger, resulting in much larger gradients, and the default learning rate causes points to move very far from the reference embedding. In our experiments, we found that lowering the learning rate to $\eta \sim 0.1$ produces stable solutions. This is especially important when using the interpolation-based approximation. Our approach places a grid of interpolation points over the embedding space, where the number of grid points is determined by the embedding range. Clearly, if even one point ``shoots off'' far from the embedding, the number of required grid points escalates, substantially increasing the runtime. The reduced learning rate suppresses this issue, and does not slow the convergence because of an adaptive learning rate scheme and running the optimization for a sufficient number of steps.

\section{Experiments and Discussion}

We apply the proposed transfer learning approach to t-SNE visualizations of single-cell data. In single-cell data sets, the data includes the variety of cells from specific tissue and characterizes the cells through the expression of its genes. In experiments, we considered several recently published data sets where cells were annotated with the cell type. Our aim was to construct t-SNE visualizations where similarly-typed cells would cluster together, despite the differences between data sources. Below, we enlist the data sets, describe single-cell specific data preprocessing procedures, and display the resulting data visualizations. Finally, we discuss the success of the proposed approach in alleviating the batch effects.

\subsection{Data}

We use three pairs of reference and secondary single-cell data sets originating from different organisms and tissues. The data in each pair were chosen so that the majority of cell types from the secondary data set were included in the reference set (Table~\ref{tab:datasets}).

\begin{table}[ht]
\begin{center}
\setlength\tabcolsep{4pt}
\begin{tabular}{l c c r c c}
\toprule
Study & Organism/Tissue & Protocol & Cells & Cell Types & Sparsity (\%) \\
\midrule
Hrvatin \etal & \multirow{2}{*}{mouse brain} & inDrop & 48,266 & 9 & 94 \\
Chen \etal & & Drop-seq & 14,437 & 6 & 93 \\[5pt]
Baron \etal & \multirow{2}{*}{human pancreas} & inDrop & 8,569 & 9 & 91 \\
Xin \etal & & SMARTer & 1,492 & 4 & 86 \\[5pt]
Macosko \etal & \multirow{2}{*}{mouse retina} & Drop-seq & 44,808 & 12 & 97 \\
Shekhar \etal & & Drop-seq & 27,499 & 5 & 96 \\
\bottomrule
\end{tabular}
\end{center}
\caption{Data sets used in our experiments. In each pair, the first data set (Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In all cases, we relied on quality control and annotations from the original studies. To facilitate comparisons, the cell annotations were harmonized using the cell type annotations from cell ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols were used to estimate gene expressions. We here report the number of cell types from each data set retained after preprocessing. Single-cell data is sparse, typically containing less than 10\% expressed genes per cell.}
\label{tab:datasets}
\end{table}

The cells in the data sets were captured from the following three tissues:
\begin{description}
\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains cells from the visual cortex exploring transcriptional changes after exposure to light. This was used as a reference for the data from Chen \etal~\cite{chen2017}, containing various cells from the mouse hypothalamus and their reaction to food deprivation. From the secondary data, we removed cells with no corresponding types in the reference, namely ependymal cells, epithelial cells, tanycytes, and unlabelled cells.

\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was created as an atlas of pancreatic cell types. We used this set as a reference for data from Xin \etal~\cite{xin2016}, who examined transcriptional differences between healthy and type 2 diabetic patients.

\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was created as an atlas of mouse retinal cell types. We used this as a reference for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for different types of retinal bipolar cells.
\end{description}

\subsection{Single-cell data preprocessing pipeline}

Due to the specific nature of single-cell data, additional steps must be taken to properly apply t-SNE. We use a standard single-cell preprocessing pipeline, consisting of selection of 3,000 representative genes (see Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation, standardization, and PCA-based representation that retains 50 principal components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to high-dimensionality of the preprocessed input data we use cosine distance to estimate similarities between reference data points~\cite{Domingos2012-CACM}. When adding new data points from secondary data set to the reference embedding, we select 1,000 genes present in both data sets and use these to estimate the similarities between the secondary data item and reference data points. The similarities are estimated using cosine similarity. We note that similarity is computed using the raw count matrices. The preprocessing stages are detailed in accompanying Python notebooks (Sec.~\ref{sec:implementation}).

\subsection{Gene selection\label{sec:gene-selection}}

Single-cell data sets suffer from a high level of technical noise and low capture efficiency, resulting in sparse expression matrices~\cite{umi}. To address this problem, we use a specialized feature-selection method, which exploits the mean-dropout relationship of expression counts as recently proposed by Kobak and Berens~\cite{art_of_using_tsne}. Here, genes with higher than expected dropout rate are regarded as potential markers for cell subpopulations and are retained in the data.

Given an expression matrix $\mathbf{X} \in \mathbb{R}^{N \times G}$ where $N$ is the number of samples and $G$ is the number of genes in the data set, we compute the fraction of cells where a gene $g$ was not expressed

\begin{equation}
d_g = \frac{1}{N} \sum_i I \left ( X_{ig} = 0\right )
\end{equation}

\noindent The mean $\log_2$ expression of the genes is computed from all the cells where gene was expressed:

\begin{equation}
m_g = \left \langle \, \log_2 X_{ig} \mid X_{ig} > t \, \right \rangle.
\end{equation}

All genes expressed in less than ten cells are discarded. In order to select a specific number of $\hat{G}$ genes, we use a binary search to find a value $b$ such that

\begin{equation}
\sum_g I \left (d_g > \exp \left [ -(m_g - b) \right ] + 0.02 \right ) = \hat{G}.
\end{equation}

\noindent In our experiments we use $t=0$ and $a=1$.

\subsection{Results and Discussion}

Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and \ref{fig:transform_retina} show the embedding of the reference data sets and their corresponding embeddings of the secondary data sets. In all the figures, the cells from the secondary data sets were positioned in the cluster of same-typed reference cells, providing strong evidence of the success of the proposed approach. There are some deviations to these observations; for instance, in Fig.~\ref{fig:transform_brain} several oligodendrocytes precursor cells (OPC) were mapped to oligodendrocytes. This may be due to differences in annotation criteria by different authors, or due to inherent similarities of these types of cells. Examples of such erroneous placements can be found in other figures as well, but they are not common and constitute less then 5\% of the cells (less than \%1 for pancreas, \%2 for retina and 5\% for brain secondary data sets).

Notice that we could simulate the split between reference and secondary data sets by cross-validation using one data set only, as this type of experiments would not incorporate batch effects. We want to remind the reader that handling batch effects were central to our endeavor and that disregard of this effect could lead to data visualizations strikingly different from ours. For example, compare the visualisations from Fig~\ref{fig:batch_effect}.a and Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and \ref{fig:transform_pancreas}.b.

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
\caption{Two-dimensional embedding of a reference brain cells (a) and corresponding mapping of secondary data from hypothalamus cells (b). Notice that the majority of hypothalamus cells were mapped to their corresponding reference cluster. For instance, the astrocyte cells marked with red on the right were mapped to an oval cluster of same-typed cells denoted with the same color in the visualization on the left.} \label{fig:transform_brain}
\end{figure}

% FIGURE: 2. t-SNE transform use-case
\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
\caption{Embedding of pancreas cells from Baron \etal~\cite{baron2016} and cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in Fig~\ref{fig:transform_brain} vast majority of the cells from the secondary data set were correctly mapped to the same-typed cluster of reference cells.}\label{fig:transform_pancreas}
\end{figure}

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/transform_retina.pdf}
\caption{Embedding of a large reference retina cells from Macosco \etal~\cite{macosko2015} (a) and mapping of cells from a smaller study that focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b). We use colors consistent with the study by Macosko \etal. Notice that Shekhar includes cells of only four different types.} \label{fig:transform_retina}
\end{figure}

There are a few tricks that we use in our procedure for t-SNE embedding of the secondary data set that were proposed recently and enhance the original t-SNE. An important one is a multi-scale extension that besides local ordering of the data points takes care about global optimization of the cluster placement. For illustration, consider visualizations with standard and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped together, while their placement in standard t-SNE is arbitrary (Fig.~\ref{fig:multiscale}.a).

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
\caption{Comparison of embedding by stanard and multiscale t-SNE on a data set from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale similarities provides a more meaningful layout of the clusters. Neuronal types occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are mainly progenitors to oligodendrocytes, but may also differentiate into neurons or astrocytes.} \label{fig:multiscale}
\end{figure}

We have also observed an important role of gene selection in crafting the reference embedding spaces. We have found that when selecting an insufficient number of genes, the resulting visualizations display fragmented clusters. When the selection is too broad and includes lowly expressed genes, the subclusters tend to overlap. These effects can all be attributed to sparseness of the data sets and may be intrinsic for single-cell data. In our studies, we found that selection of 3,000 genes yields most informative visualizations (Fig.~\ref{fig:gene_selection}).

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/hrvatin_embedding_tsne_genes.pdf}
\caption{Gene selection plays an important role when constructing the reference embedding. {\bf (a)} Using too few genes results in over-clustering. {\bf (b)} Using an intermediate number of genes reveals clustering mostly consistent with cell annotations. {\bf (c)} Including all the genes may lead to under-clustering of the more specialized cell types.}
\label{fig:gene_selection}
\end{figure}

In principle, our theoretically-grounded embedding of secondary data into the scaffold defined by reference embedding could be simplified with the application of the nearest neighbors-based procedure. For example, while describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne} proposed to position new points into a known embedding by placing them in the median position of their 10 nearest neighbors, where the neighborhood was estimated in the original data space. Notice that we use this trick as well, but only for initialization of positions of new data instances that are subject to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that nearest neighboring-based positioning is insufficient and may yield clumped visualizations where the optimal positioning using the t-SNE loss function is much more dispersed and rightfully shows a more considerable variation in the secondary data. Some data points may also fell into the neighboring space between differently typed clusters, while after optimization they typically converge closer to same-typed groups.

\begin{figure}[htbp]
\includegraphics[width=\textwidth]{figures/optimization_retina.pdf}
\caption{Comparison of data placement by nearest neighbors approach from Kobak and Berens~\cite{art_of_using_tsne} and optimized placement by our algorithm. {\bf (a)} Data points are placed to the median position of their 25 nearest neighbors in the reference set. {\bf (b)} Point positions are optimized, revealing a different, more dispersed placement that better reflects the variety of cells in the secondary data set.}
\label{fig:optimization}
\end{figure}

The proposed method assumes that all cell types from the secondary data set are present in the reference. The proposed method would fail to reveal novel cell types in the secondary data set, possibly positioning them arbitrarily close to unrelated clusters. Procedures such as scmap~\cite{scmap} were recently proposed to cope with such cases and identify the cells whose type is new and not included in the reference. Our procedure does not address such cases, and for scaling-up to a wider collection of cell types relies on emerging availability of large collections of the reference data such as those managed by Human Cell Atlas initiative~\cite{hca}.

\subsection{Implementation\label{sec:implementation}}

The procedures described in this paper are provided as Python notebooks that are, together with the data, available in an open repository~\footnote{https://github.com/biolab/tsne-embedding}. All experiments were run using openTSNE~\footnote{https://github.com/pavlin-policar/openTSNE}, our open and extensible t-SNE library for Python.

\section{Conclusion}

Almost all recent publications of single-cell studies start with a two-dimensional visualization of the data that exposes the diversity as well as different types of cells from the study. While any dimensionality reduction technique can be used to render such a visualization, different variants of t-SNE are most often used. Due to the ability to explore biological mechanisms at the cellular level, single-cell studies are increasingly widespread, and their publications in the past couple of years are abundant. One of the central tasks in single-cell studies is the classification of new cells based on finding from previous studies. Such transfer of knowledge is often difficult due to batch effects present in data from different sources. Solving the problem of the batch-effects together with prevailing means of presenting single-cell data in two-dimensional visualizations motivated the research presented in this paper.

Our proposed approach uses t-SNE embedding as a scaffold for the positioning of new cells within the visualization, and possibly for aiding their classification. The three case studies incorporating pairs of data sets from different domains but with similar classifications demonstrate that our proposed procedure can effectively deal with batch effects to construct visualizations that correctly map a secondary data set onto a reference data set from an independent study that possibly uses different data collection protocol. While we here focused on t-SNE constructed from reference data sets, the approach could be applied to any existing two-dimensional visualization.

\subsubsection*{Acknowledgements}

This work was supported by the Slovenian Research Agency Program Grant P2-0209, and by the BioPharm.SI project supported from European Regional Development Fund and the Slovenian Ministry of Education, Science and Sport. We would also like to thank Dmitry Kobak for helpful discussions on t-SNE extensions.

\bibliographystyle{splncs04}
\bibliography{references}
\end{document}

### 2019-06-13 paper/transfer_learning.tex
Fix number of neighbors for initialization to 10

@@
 We initialize the position of the data point in embedding space to the median embedding position of its $k$ nearest neighbors from the reference data set.
-While we found the algorithm to be robust to choices of $k$, we use $k=25$ in our experiments.
+While we found the algorithm to be robust to choices of $k$, we use $k=10$ in our experiments.
 
@@
 \caption{Comparison of data placement by nearest neighbors approach from Kobak and Berens~\cite{art_of_using_tsne} and optimized placement by our algorithm.
-{\bf (a)} Data points are placed to the median position of their 25 nearest neighbors in the reference set.
+{\bf (a)} Data points are placed to the median position of their 10 nearest neighbors in the reference set.
 {\bf (b)} Point positions are optimized, revealing a different, more dispersed placement that better reflects the variety of cells in the secondary data set.}

### 2019-06-13 paper/tsne_embedding.tex
Word wrap.

--- 2019-06-13 paper/transfer_learning.tex
@@
 \begin{abstract}
-Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data.
-When working with multiple data sets, the straightforward application of these methods often fails, and instead of revealing underlying classes, the resulting visualizations expose domain-specific clusters.
-To circumvent these batch effects, we here propose an embedding procedure that takes a t-SNE visualization constructed on reference data set and uses it as a scaffold for new embeddings.
-New, secondary data is embedded one data-instance at the time, thus disregarding any interactions between data items in the secondary data and implicitly mitigating batch effects.
-We demonstrate the utility of this procedure on analysis of six recently published single-cell gene expression data sets containing up to tens of thousands of cells and thousands of genes.
-In these data sets, the batch effects were particularly strong as the data comes from different institutions and was obtained using different experimental protocols.
-The visualizations constructed by the proposed approach are absent of batch effects, and the cells from secondary data sets correctly co-cluster with cells from the primary data sharing the same cell type.
+
+Dimensionality reduction techniques, such as t-SNE, can construct informative
+visualizations of high-dimensional data. When working with multiple data sets,
+the straightforward application of these methods often fails, and instead of
+revealing underlying classes, the resulting visualizations expose
+domain-specific clusters. To circumvent these batch effects, we here propose an
+embedding procedure that takes a t-SNE visualization constructed on reference
+data set and uses it as a scaffold for new embeddings. New, secondary data is
+embedded one data-instance at the time, thus disregarding any interactions
+between data items in the secondary data and implicitly mitigating batch
+effects. We demonstrate the utility of this procedure on analysis of six
+recently published single-cell gene expression data sets containing up to tens
+of thousands of cells and thousands of genes. In these data sets, the batch
+effects were particularly strong as the data comes from different institutions
+and was obtained using different experimental protocols. The visualizations
+constructed by the proposed approach are absent of batch effects, and the cells
+from secondary data sets correctly co-cluster with cells from the primary data
+sharing the same cell type.
 
@@
 
-Two-dimensional embeddings and their visualizations may assist in the analysis and interpretation of multi-dimensional data.
-Intuitively, two data instances should be co-located in the resulting visualization if their multi-dimensional profiles are similar.
-For this task, non-linear local embeddings such as t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform manifold approximation and projection~\cite{umap} have recently complemented traditional data transformation and embedding approaches such as principal component analysis (PCA) and multi-dimensional scaling~\cite{distill,umap_single_cell}.
-While useful for visualizing the data from a single coherent source, these methods may encounter problems if the data comes from different sources.
-Here, if executing the dimensionality reduction procedures on a merged data, the resulting visualizations would typically depict source-specific clusters instead of reveal clusters that are semantically enriched, say, with same-type of data instances across different sources.
-This source-specific confounding is often referred to as {\em domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or {\em dataset shift}~\cite{dataset_shift}.
-In the bioinformatics literature, the domain-specific differences are more commonly referred to as {\em batch effects}~\cite{cca,mnn,seurat}.
+Two-dimensional embeddings and their visualizations may assist in the analysis
+and interpretation of multi-dimensional data. Intuitively, two data instances
+should be co-located in the resulting visualization if their multi-dimensional
+profiles are similar. For this task, non-linear local embeddings such as
+t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
+manifold approximation and projection~\cite{umap} have recently complemented
+traditional data transformation and embedding approaches such as principal
+component analysis (PCA) and multi-dimensional
+scaling~\cite{distill,umap_single_cell}. While useful for visualizing the data
+from a single coherent source, these methods may encounter problems if the data
+comes from different sources. Here, if executing the dimensionality reduction
+procedures on a merged data, the resulting visualizations would typically
+depict source-specific clusters instead of reveal clusters that are
+semantically enriched, say, with same-type of data instances across different
+sources. This source-specific confounding is often referred to as {\em domain
+shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
+{\em dataset shift}~\cite{dataset_shift}. In the bioinformatics literature, the
+domain-specific differences are more commonly referred to as {\em batch
+effects}~\cite{cca,mnn,seurat}.
 
-Massive, multi-variate biological data sets typical suffer from source-specific biases.
-Consider an example from single-cell genomics, a domain we will focus on in this manuscript and that was --- besides current scientific challenges --- selected also due to availability and the abundance of recently published data.
-Single-cell data sets may include thousands of cells whose profiles consist of tens of thousands of cell-specific gene expressions.
-Single cell studies typically start with the analysis of cell types, where we expect that cells of the same type would cluster in two-dimensional data visualisation~\cite{seurat}.
-For instance, Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells originating from visual cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}.
-The figure reveals distinct clusters but also separates the data from the two brain regions.
-These two regions share the same cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two studies to overlap.
-Batch effects similarly prohibit the utility of t-SNE in the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from human cell atlas~\cite{baron2016} and similarly-typed cells from diabetic patients~\cite{xin2016}.
-Just like with data from brain cells, pancreas cells cluster first according to the data source, again resulting in an uninformative visualization that primarily exposes the batch effect.
+Massive, multi-variate biological data sets typical suffer from source-specific
+biases. Consider an example from single-cell genomics, a domain we will focus
+on in this manuscript and that was --- besides current scientific challenges
+--- selected also due to availability and the abundance of recently published
+data. Single-cell data sets may include thousands of cells whose profiles
+consist of tens of thousands of cell-specific gene expressions. Single cell
+studies typically start with the analysis of cell types, where we expect that
+cells of the same type would cluster in two-dimensional data
+visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
+t-SNE embedded data from mouse brain cells originating from visual
+cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
+distinct clusters but also separates the data from the two brain regions. These
+two regions share the same cell types and --- contrary to the depiction in
+Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
+studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
+the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which
+renders the data from human cell atlas~\cite{baron2016} and similarly-typed
+cells from diabetic patients~\cite{xin2016}. Just like with data from brain
+cells, pancreas cells cluster first according to the data source, again
+resulting in an uninformative visualization that primarily exposes the batch
+effect.
 
@@
 \caption{Batch effects are a driving factor of variation between the data sets.
-Depicted is a t-SNE visualisation of two pairs of data sets.
-In each pair, the data sets share cell types, so it would be expected that the cells from the reference data (blue) would mix with the cells in a secondary data sets (orange).
-Instead, t-SNE visualisation clusters data according to the data source.} \label{fig:batch_effect}
+Depicted is a t-SNE visualisation of two pairs of data sets. In each pair, the
+data sets share cell types, so it would be expected that the cells from the
+reference data (blue) would mix with the cells in a secondary data sets
+(orange). Instead, t-SNE visualisation clusters data according to the data
+source.} \label{fig:batch_effect}
 \end{figure}
 
-Current solutions to embedding the data from various data sources address the batch effect problems up-front.
-The data is typically pre-processed and transformed, possibly to the latent space, so that the batch effects are removed.
-The recently proposed procedures for batch-effect removal include canonical correlation analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.
-In these works, a proof for the success of batch effect removal is a good mixing of cells from different sources in a t-SNE visualization.
-Elimination of batch effects may require aggressive data preprocessing which may blur the boundaries between cell types.
-Another problem is also the inclusion of any new data, for which the entire data analysis pipeline has to be rerun, usually resulting in different layout and clusters that have little resemblance to original visualization and thus require reinterpretation.
+Current solutions to embedding the data from various data sources address the
+batch effect problems up-front. The data is typically pre-processed and
+transformed, possibly to the latent space, so that the batch effects are
+removed. The recently proposed procedures for batch-effect removal include
+canonical correlation analysis~\cite{cca} and mutual
+nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of
+batch effect removal is a good mixing of cells from different sources in a
+t-SNE visualization. Elimination of batch effects may require aggressive data
+preprocessing which may blur the boundaries between cell types. Another problem
+is also the inclusion of any new data, for which the entire data analysis
+pipeline has to be rerun, usually resulting in different layout and clusters
+that have little resemblance to original visualization and thus require
+reinterpretation.
 
-We propose a direct solution of rendering t-SNE visualizations that addresses batch effects.
-Our approach considers one of the data sets as a {\em reference} and aims to visually classify the cells in the other, {\em secondary data set}.
-We construct the t-SNE embedding using the reference data set, and then use it as a scaffold for the embedding of data points from the secondary data.
-Embedding is performed one data point at a time.
-Independence of each new embedding of data instances from secondary data set causes clustering landscape to only depend on a reference scaffold, thus removing data source-driven variation.
-In other words, when including new data, the scaffold inferred from the reference data set is kept unchanged and defines the ``gravitational field'' to independently embedded each new data item.
-For example, in Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from hypothalamus (Fig.~\ref{fig:transform_brain}.b).
-Unlike in their joint t-SNE visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are dispersed across entire embedding space and their cell type correctly matches the prevailing type in reference clusters.
+We propose a direct solution of rendering t-SNE visualizations that addresses
+batch effects. Our approach considers one of the data sets as a {\em reference}
+and aims to visually classify the cells in the other, {\em secondary data set}.
+We construct the t-SNE embedding using the reference data set, and then use it
+as a scaffold for the embedding of data points from the secondary data.
+Embedding is performed one data point at a time. Independence of each new
+embedding of data instances from secondary data set causes clustering landscape
+to only depend on a reference scaffold, thus removing data source-driven
+variation. In other words, when including new data, the scaffold inferred from
+the reference data set is kept unchanged and defines the ``gravitational
+field'' to independently embedded each new data item. For example, in
+Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
+scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
+hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t-SNE
+visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are
+dispersed across entire embedding space and their cell type correctly matches
+the prevailing type in reference clusters.
 
-At the core of the proposed solution is a mapping that implements embedding of new data to existing t-SNE visualization.
-In the following, we introduce t-SNE, then describe its recently proposed multi-scale extension, and outline our algorithm that embeds new data into a previously trained scaffold.
-While a utility of such an algorithm was already hinted to in recent publication~\cite{art_of_using_tsne}, we here provide its practical and theoretically-grounded implementation.
-Considering the abundance of recent publications in batch effect removal, we present surprising evidence that a computationally more direct and elegant embedding procedure solves the batch effects problem when constructing interpretable visualizations from different data sources.
+At the core of the proposed solution is a mapping that implements embedding of
+new data to existing t-SNE visualization. In the following, we introduce t-SNE,
+then describe its recently proposed multi-scale extension, and outline our
+algorithm that embeds new data into a previously trained scaffold. While a
+utility of such an algorithm was already hinted to in recent
+publication~\cite{art_of_using_tsne}, we here provide its practical and
+theoretically-grounded implementation. Considering the abundance of recent
+publications in batch effect removal, we present surprising evidence that a
+computationally more direct and elegant embedding procedure solves the batch
+effects problem when constructing interpretable visualizations from different
+data sources.
 
@@
 
-We describe an end-to-end pipeline that uses t-SNE embeddings as a scaffold for new data samples, and enables visualisation of data from different sources while mitigating batch effects.
-Our proposed approach starts with t\nobreakdash -SNE to embedd a reference data set, with the aim of constructing a two-dimensional visualisation to facilitate interpretation and cluster classification.
-We regard the resulting two-dimensional embedding as a scaffold to which we add samples from the new, secondary data source.
-Each new sample is then placed into the reference embedding and optimized independently using the t\nobreakdash -SNE loss function.
-Independent embedding of each data item from a secondary data set disregards any interactions present in that data, and prevents forming clusters that would be data-souce specific.
-Below, we start with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing the notation and relevant parts upon which we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).
+We describe an end-to-end pipeline that uses t-SNE embeddings as a scaffold for
+new data samples, and enables visualisation of data from different sources
+while mitigating batch effects. Our proposed approach starts with t\nobreakdash
+-SNE to embedd a reference data set, with the aim of constructing a
+two-dimensional visualisation to facilitate interpretation and cluster
+classification. We regard the resulting two-dimensional embedding as a scaffold
+to which we add samples from the new, secondary data source. Each new sample is
+then placed into the reference embedding and optimized independently using the
+t\nobreakdash -SNE loss function. Independent embedding of each data item from
+a secondary data set disregards any interactions present in that data, and
+prevents forming clusters that would be data-souce specific. Below, we start
+with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing
+the notation and relevant parts upon which we base our secondary data embedding
+approach (Sec.~\ref{sec:transfer}).
 
@@
 
-The embedding by t-SNE is local and non-linear, tailored to visualisation of high dimensional data sets.
-Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the number of samples in the reference data set, t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space, their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also close.
-Since t-SNE is primarily used as a visualization tool, $d$ is typically set to two.
-The similarity between two data points in t-SNE is defined as:
+The embedding by t-SNE is local and non-linear, tailored to visualisation of
+high dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
+\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
+\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
+t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
+\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$
+where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
+close in the multi-dimensional space, their corresponding embeddings
+$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
+as a visualization tool, $d$ is typically set to two. The similarity between
+two data points in t-SNE is defined as:
 
@@
 
-The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the perplexity of the distribution matches a user-specified parameter value
+The bandwidth of each Gaussian kernel $\sigma_i$ is selected such that the
+perplexity of the distribution matches a user-specified parameter value
 
@@
 
-\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying density of the data in the multi-dimensional space.
+\noindent Different bandwidths $\sigma_i$ enable t-SNE to adapt to the varying
+density of the data in the multi-dimensional space.
 
-The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the embedding space are defined using the $t$-distribution with one degree of freedom
+The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
+embedding space are defined using the $t$-distribution with one degree of
+freedom
 
@@
 
-The data transformation by t-SNE finds an embedding $\mathbf{Y}$ that minimizes the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
+The data transformation by t-SNE finds an embedding $\mathbf{Y}$ that minimizes
+the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
 
@@
 
-The time complexity needed to evaluate the similarities in Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of the algorithm impractical for any reasonably-sized data.
-To address larger data sets, we instead adapt a recent approach for low-rank approximation of similarity matrix based on polynomial interpolation that reduces the time complexity of t-SNE to $\mathcal{O}(N)$.
-This approximation enables the visualization of massive data sets possibly containing millions of data points~\cite{fi_tsne}.
+The time complexity needed to evaluate the similarities in
+Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of
+the algorithm impractical for any reasonably-sized data. To address larger data
+sets, we instead adapt a recent approach for low-rank approximation of
+similarity matrix based on polynomial interpolation that reduces the time
+complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
+visualization of massive data sets possibly containing millions of data
+points~\cite{fi_tsne}.
 
-The t-SNE embeddings substantially depend on the value of the perplexity parameter.
-The perplexity parameter can be interpreted as the number of neighbors for which the distances in the embedding space are preserved.
-Small values of perplexity result in tightly-packed clusters of points and lead to ignoring the long-range interactions between clusters.
-Larger values may result in a more globally consistent visualisations, preserving distances on a large scale and organizing clusters in a more meaningful way.
-Larger values of perplexity may often result in merging multiple small clusters, thus obscuring local aspects of the data~\cite{art_of_using_tsne}.
+The t-SNE embeddings substantially depend on the value of the perplexity
+parameter. The perplexity parameter can be interpreted as the number of
+neighbors for which the distances in the embedding space are preserved. Small
+values of perplexity result in tightly-packed clusters of points and lead to
+ignoring the long-range interactions between clusters. Larger values may result
+in a more globally consistent visualisations, preserving distances on a large
+scale and organizing clusters in a more meaningful way. Larger values of
+perplexity may often result in merging multiple small clusters, thus obscuring
+local aspects of the data~\cite{art_of_using_tsne}.
 
-A trade-off between the local organization and global consistency may be achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}.
+A trade-off between the local organization and global consistency may be
+achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
+a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}.
 Multi-scale kernels are defined as
@@
 
-\noindent where $L$ is the number of mixture components.
-The bandwidths $\sigma_{i,l}$ are selected in the same manner as in Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each $l$.
-In our experiments, we used a mixture of two Gaussian kernels with perplexity values of 50 and 500.
-We note that a similar formulation of multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the resulting embeddings are visually very similar to those obtained with the approach described above (for brevity, data not shown).
+\noindent where $L$ is the number of mixture components. The bandwidths
+$\sigma_{i,l}$ are selected in the same manner as in
+Eq.~\ref{eq:gaussian_kernel}, but with a different value of perplexity for each
+$l$. In our experiments, we used a mixture of two Gaussian kernels with
+perplexity values of 50 and 500. We note that a similar formulation of
+multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
+resulting embeddings are visually very similar to those obtained with the
+approach described above (for brevity, data not shown).
 
@@
 
-An algorithm to embedd new data point to a reference embedding consists of estimating similarities between the point and the reference data and optimizing the position of the data point in the embedding space.
-Unlike parametric models such as principal component analysis or autoencoders, t-SNE does not define an explicit mapping to embedding space, and embeddings need to be found through loss function optimization.
+An algorithm to embedd new data point to a reference embedding consists of
+estimating similarities between the point and the reference data and optimizing
+the position of the data point in the embedding space. Unlike parametric models
+such as principal component analysis or autoencoders, t-SNE does not define an
+explicit mapping to embedding space, and embeddings need to be found through
+loss function optimization.
 
-For each new data point, we first estimate its distance to each of the data instances in the reference data set.
-We initialize the position of the data point in embedding space to the median embedding position of its $k$ nearest neighbors from the reference data set.
-While we found the algorithm to be robust to choices of $k$, we use $k=10$ in our experiments.
+For each new data point, we first estimate its distance to each of the data
+instances in the reference data set. We initialize the position of the data
+point in embedding space to the median embedding position of its $k$ nearest
+neighbors from the reference data set. While we found the algorithm to be
+robust to choices of $k$, we use $k=10$ in our experiments.
 
-We adapt the standard t-SNE formulation from Eqs.~\ref{eq:gaussian_kernel}~and~\ref{eq:cauchy_kernel} with
+We adapt the standard t-SNE formulation from
+Eqs.~\ref{eq:gaussian_kernel}~and~\ref{eq:cauchy_kernel} with
 
@@
 
-\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots, \mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots, \mathbf{w}_M \right \} \in \mathbb{R}^d$.
-Additionally, we omit the symmetrization step in Eq.~\ref{eq:symmetrize}.
-This enables new points to be inserted into the embedding independently of one another.
-The gradients of $\mathbf{w}_j$ with respect to the loss (Eq.~\ref{eq:kl_divergence}) are:
+\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
+\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
+the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
+\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
+symmetrization step in Eq.~\ref{eq:symmetrize}. This enables new points to be
+inserted into the embedding independently of one another. The gradients of
+$\mathbf{w}_j$ with respect to the loss (Eq.~\ref{eq:kl_divergence}) are:
 
@@
 
-In the optimization step, we refine point positions using batch gradient descent.
-We use an adaptive learning rate scheme with momentum proposed by Jacobs to speed up the convergence~\cite{momentum}.
-We run gradient descent with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the optimization converged for all cases in our experiments.
-The time complexity needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$, however, by adapting the same polynomial interpolation based approximation, this is reduced to $\mathcal{O}(N)$.
+In the optimization step, we refine point positions using batch gradient
+descent. We use an adaptive learning rate scheme with momentum proposed by
+Jacobs to speed up the convergence~\cite{momentum}. We run gradient descent
+with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
+optimization converged for all cases in our experiments.  The time complexity
+needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
+\cdot M)$, however, by adapting the same polynomial interpolation based
+approximation, this is reduced to $\mathcal{O}(N)$.
 
-Special care must be taken to reduce the learning rate $\eta$ as the default value in most implementations ($\eta = 200$) may cause points to ``shoot off'' from the reference embedding.
-This phenomenon is caused due to the embedding to already defined t-SNE space, where the distances between data points and corresponding gradients of the optimization function may be quite large.
-When running s standard t-SNE, points are initialized and scaled to have variance 0.0001.
-The resulting gradients tend to be very small during the initial phase, resulting in stable convergence.
-When embedding new samples, the range of the embedding is much larger, resulting in much larger gradients, and the default learning rate causes points to move very far from the reference embedding.
-In our experiments, we found that lowering the learning rate to $\eta \sim 0.1$ produces stable solutions.
-This is especially important when using the interpolation-based approximation.
-Our approach places a grid of interpolation points over the embedding space, where the number of grid points is determined by the embedding range.
-Clearly, if even one point ``shoots off'' far from the embedding, the number of required grid points escalates, substantially increasing the runtime.
-The reduced learning rate suppresses this issue, and does not slow the convergence because of an adaptive learning rate scheme and running the optimization for a sufficient number of steps.
+Special care must be taken to reduce the learning rate $\eta$ as the default
+value in most implementations ($\eta = 200$) may cause points to ``shoot off''
+from the reference embedding. This phenomenon is caused due to the embedding to
+already defined t-SNE space, where the distances between data points and
+corresponding gradients of the optimization function may be quite large. When
+running s standard t-SNE, points are initialized and scaled to have variance
+0.0001. The resulting gradients tend to be very small during the initial phase,
+resulting in stable convergence. When embedding new samples, the range of the
+embedding is much larger, resulting in much larger gradients, and the default
+learning rate causes points to move very far from the reference embedding. In
+our experiments, we found that lowering the learning rate to $\eta \sim 0.1$
+produces stable solutions. This is especially important when using the
+interpolation-based approximation. Our approach places a grid of interpolation
+points over the embedding space, where the number of grid points is determined
+by the embedding range. Clearly, if even one point ``shoots off'' far from the
+embedding, the number of required grid points escalates, substantially
+increasing the runtime. The reduced learning rate suppresses this issue, and
+does not slow the convergence because of an adaptive learning rate scheme and
+running the optimization for a sufficient number of steps.
 
@@
 
-We apply the proposed transfer learning approach to t-SNE visualizations of single-cell data.
-In single-cell data sets, the data includes the variety of cells from specific tissue and characterizes the cells through the expression of its genes.
-In experiments, we considered several recently published data sets where cells were annotated with the cell type.
-Our aim was to construct t-SNE visualizations where similarly-typed cells would cluster together, despite the differences between data sources.
-Below, we enlist the data sets, describe single-cell specific data preprocessing procedures, and display the resulting data visualizations.
-Finally, we discuss the success of the proposed approach in alleviating the batch effects.
+We apply the proposed transfer learning approach to t-SNE visualizations of
+single-cell data. In single-cell data sets, the data includes the variety of
+cells from specific tissue and characterizes the cells through the expression
+of its genes. In experiments, we considered several recently published data
+sets where cells were annotated with the cell type. Our aim was to construct
+t-SNE visualizations where similarly-typed cells would cluster together,
+despite the differences between data sources. Below, we enlist the data sets,
+describe single-cell specific data preprocessing procedures, and display the
+resulting data visualizations. Finally, we discuss the success of the proposed
+approach in alleviating the batch effects.
 
@@
 
-We use three pairs of reference and secondary single-cell data sets originating from different organisms and tissues.
-The data in each pair were chosen so that the majority of cell types from the secondary data set were included in the reference set (Table~\ref{tab:datasets}).
+We use three pairs of reference and secondary single-cell data sets originating
+from different organisms and tissues. The data in each pair were chosen so that
+the majority of cell types from the secondary data set were included in the
+reference set (Table~\ref{tab:datasets}).
 
@@
 \end{center}
-\caption{Data sets used in our experiments.
-In each pair, the first data set (Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.
-In all cases, we relied on quality control and annotations from the original studies.
-To facilitate comparisons, the cell annotations were harmonized using the cell type annotations from cell ontology~\cite{cell_ontology}.
-Notice that different RNA sequencing protocols were used to estimate gene expressions.
-We here report the number of cell types from each data set retained after preprocessing.
-Single-cell data is sparse, typically containing less than 10\% expressed genes per cell.}
+\caption{Data sets used in our experiments. In each pair, the first data set
+(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
+all cases, we relied on quality control and annotations from the original
+studies. To facilitate comparisons, the cell annotations were harmonized using
+the cell type annotations from cell ontology~\cite{cell_ontology}. Notice that
+different RNA sequencing protocols were used to estimate gene expressions. We
+here report the number of cell types from each data set retained after
+preprocessing. Single-cell data is sparse, typically containing less than 10\%
+expressed genes per cell.}
 \label{tab:datasets}
@@
 \begin{description}
-\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains cells from the visual cortex exploring transcriptional changes after exposure to light.
-This was used as a reference for the data from Chen \etal~\cite{chen2017}, containing various cells from the mouse hypothalamus and their reaction to food deprivation.
-From the secondary data, we removed cells with no corresponding types in the reference, namely ependymal cells, epithelial cells, tanycytes, and unlabelled cells.
+\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
+cells from the visual cortex exploring transcriptional changes after exposure
+to light. This was used as a reference for the data from Chen
+\etal~\cite{chen2017}, containing various cells from the mouse hypothalamus and
+their reaction to food deprivation. From the secondary data, we removed cells
+with no corresponding types in the reference, namely ependymal cells,
+epithelial cells, tanycytes, and unlabelled cells.
 
-\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was created as an atlas of pancreatic cell types.
-We used this set as a reference for data from Xin \etal~\cite{xin2016}, who examined transcriptional differences between healthy and type 2 diabetic patients.
+\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was
+created as an atlas of pancreatic cell types. We used this set as a reference
+for data from Xin \etal~\cite{xin2016}, who examined transcriptional
+differences between healthy and type 2 diabetic patients.
 
-\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was created as an atlas of mouse retinal cell types.
-We used this as a reference for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for different types of retinal bipolar cells.
+\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
+created as an atlas of mouse retinal cell types. We used this as a reference
+for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
+different types of retinal bipolar cells.
 \end{description}
@@
 
-Due to the specific nature of single-cell data, additional steps must be taken to properly apply t-SNE.
-We use a standard single-cell preprocessing pipeline, consisting of selection of 3,000 representative genes (see Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation, standardization, and PCA-based representation that retains 50 principal components~\cite{seurat,scanpy}.
-To obtain the reference embedding, we apply multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}.
-Due to high-dimensionality of the preprocessed input data we use cosine distance to estimate similarities between reference data points~\cite{Domingos2012-CACM}.
-When adding new data points from secondary data set to the reference embedding, we select 1,000 genes present in both data sets and use these to estimate the similarities between the secondary data item and reference data points.
-The similarities are estimated using cosine similarity.
-We note that similarity is computed using the raw count matrices.
-The preprocessing stages are detailed in accompanying Python notebooks (Sec.~\ref{sec:implementation}).
+Due to the specific nature of single-cell data, additional steps must be taken
+to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
+consisting of selection of 3,000 representative genes (see
+Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
+standardization, and PCA-based representation that retains 50 principal
+components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
+multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
+high-dimensionality of the preprocessed input data we use cosine distance to
+estimate similarities between reference data points~\cite{Domingos2012-CACM}.
+When adding new data points from secondary data set to the reference embedding,
+we select 1,000 genes present in both data sets and use these to estimate the
+similarities between the secondary data item and reference data points. The
+similarities are estimated using cosine similarity. We note that similarity is
+computed using the raw count matrices. The preprocessing stages are detailed in
+accompanying Python notebooks (Sec.~\ref{sec:implementation}).
 
@@
 
-Single-cell data sets suffer from a high level of technical noise and low capture efficiency, resulting in sparse expression matrices~\cite{umi}.
-To address this problem, we use a specialized feature-selection method, which exploits the mean-dropout relationship of expression counts as recently proposed by Kobak and Berens~\cite{art_of_using_tsne}.
-Here, genes with higher than expected dropout rate are regarded as potential markers for cell subpopulations and are retained in the data.
+Single-cell data sets suffer from a high level of technical noise and low
+capture efficiency, resulting in sparse expression matrices~\cite{umi}. To
+address this problem, we use a specialized feature-selection method, which
+exploits the mean-dropout relationship of expression counts as recently
+proposed by Kobak and Berens~\cite{art_of_using_tsne}. Here, genes with higher
+than expected dropout rate are regarded as potential markers for cell
+subpopulations and are retained in the data.
 
-Given an expression matrix $\mathbf{X} \in \mathbb{R}^{N \times G}$ where $N$ is the number of samples and $G$ is the number of genes in the data set, we compute the fraction of cells where a gene $g$ was not expressed
+Given an expression matrix $\mathbf{X} \in \mathbb{R}^{N \times G}$ where $N$
+is the number of samples and $G$ is the number of genes in the data set, we
+compute the fraction of cells where a gene $g$ was not expressed
 
@@
 
-\noindent The mean $\log_2$ expression of the genes is computed from all the cells where gene was expressed:
+\noindent The mean $\log_2$ expression of the genes is computed from all the
+cells where gene was expressed:
 
@@
 
-All genes expressed in less than ten cells are discarded. In order to select a specific number of $\hat{G}$ genes, we use a binary search to find a value $b$ such that
+All genes expressed in less than ten cells are discarded. In order to select a
+specific number of $\hat{G}$ genes, we use a binary search to find a value $b$
+such that
 
@@
 
-Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and \ref{fig:transform_retina} show the embedding of the reference data sets and their corresponding embeddings of the secondary data sets.
-In all the figures, the cells from the secondary data sets were positioned in the cluster of same-typed reference cells, providing strong evidence of the success of the proposed approach.
-There are some deviations to these observations; for instance, in Fig.~\ref{fig:transform_brain} several oligodendrocytes precursor cells (OPC) were mapped to oligodendrocytes.
-This may be due to differences in annotation criteria by different authors, or due to inherent similarities of these types of cells.
-Examples of such erroneous placements can be found in other figures as well, but they are not common and constitute less then 5\% of the cells (less than \%1 for pancreas, \%2 for retina and 5\% for brain secondary data sets).
+Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
+\ref{fig:transform_retina} show the embedding of the reference data sets and
+their corresponding embeddings of the secondary data sets. In all the figures,
+the cells from the secondary data sets were positioned in the cluster of
+same-typed reference cells, providing strong evidence of the success of the
+proposed approach. There are some deviations to these observations; for
+instance, in Fig.~\ref{fig:transform_brain} several oligodendrocytes precursor
+cells (OPC) were mapped to oligodendrocytes. This may be due to differences in
+annotation criteria by different authors, or due to inherent similarities of
+these types of cells. Examples of such erroneous placements can be found in
+other figures as well, but they are not common and constitute less then 5\% of
+the cells (less than \%1 for pancreas, \%2 for retina and 5\% for brain
+secondary data sets).
 
-Notice that we could simulate the split between reference and secondary data sets by cross-validation using one data set only, as this type of experiments would not incorporate batch effects.
-We want to remind the reader that handling batch effects were central to our endeavor and that disregard of this effect could lead to data visualizations strikingly different from ours.
-For example, compare the visualisations from Fig~\ref{fig:batch_effect}.a and Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and \ref{fig:transform_pancreas}.b.
+Notice that we could simulate the split between reference and secondary data
+sets by cross-validation using one data set only, as this type of experiments
+would not incorporate batch effects. We want to remind the reader that handling
+batch effects were central to our endeavor and that disregard of this effect
+could lead to data visualizations strikingly different from ours. For example,
+compare the visualisations from Fig~\ref{fig:batch_effect}.a and
+Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
+\ref{fig:transform_pancreas}.b.
 
@@
 \includegraphics[width=\textwidth]{figures/transform_brain.pdf}
-\caption{Two-dimensional embedding of a reference brain cells (a) and corresponding mapping of secondary data from hypothalamus cells (b).
-Notice that the majority of hypothalamus cells were mapped to their corresponding reference cluster.
-For instance, the astrocyte cells marked with red on the right were mapped to an oval cluster of same-typed cells denoted with the same color in the visualization on the left.} \label{fig:transform_brain}
+\caption{Two-dimensional embedding of a reference brain cells (a) and
+corresponding mapping of secondary data from hypothalamus cells (b). Notice
+that the majority of hypothalamus cells were mapped to their corresponding
+reference cluster. For instance, the astrocyte cells marked with red on the
+right were mapped to an oval cluster of same-typed cells denoted with the same
+color in the visualization on the left.} \label{fig:transform_brain}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
-\caption{Embedding of pancreas cells from Baron \etal~\cite{baron2016} and cells from the same tissue from Xin \etal.~\cite{xin2016}.
-Just like in Fig~\ref{fig:transform_brain} vast majority of the cells from the secondary data set were correctly mapped to the same-typed cluster of reference cells.}\label{fig:transform_pancreas}
+\caption{Embedding of pancreas cells from Baron \etal~\cite{baron2016} and
+cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
+Fig~\ref{fig:transform_brain} vast majority of the cells from the secondary
+data set were correctly mapped to the same-typed cluster of reference
+cells.}\label{fig:transform_pancreas}
 \end{figure}
@@
 \includegraphics[width=\textwidth]{figures/transform_retina.pdf}
-\caption{Embedding of a large reference retina cells from Macosco \etal~\cite{macosko2015} (a) and mapping of cells from a smaller study that focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b).
-We use colors consistent with the study by Macosko \etal.
-Notice that Shekhar includes cells of only four different types.} \label{fig:transform_retina}
+\caption{Embedding of a large reference retina cells from Macosco
+\etal~\cite{macosko2015} (a) and mapping of cells from a smaller study that
+focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b). We use
+colors consistent with the study by Macosko \etal. Notice that Shekhar includes
+cells of only four different types.} \label{fig:transform_retina}
 \end{figure}
 
-There are a few tricks that we use in our procedure for t-SNE embedding of the secondary data set that were proposed recently and enhance the original t-SNE.
-An important one is a multi-scale extension that besides local ordering of the data points takes care about global optimization of the cluster placement.
-For illustration, consider visualizations with standard and multi-scale t-SNE in Fig.~\ref{fig:multiscale}.
-Notice, for instance, that in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped together, while their placement in standard t-SNE is arbitrary (Fig.~\ref{fig:multiscale}.a).
+There are a few tricks that we use in our procedure for t-SNE embedding of the
+secondary data set that were proposed recently and enhance the original t-SNE.
+An important one is a multi-scale extension that besides local ordering of the
+data points takes care about global optimization of the cluster placement. For
+illustration, consider visualizations with standard and multi-scale t-SNE in
+Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE
+(Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped
+together, while their placement in standard t-SNE is arbitrary
+(Fig.~\ref{fig:multiscale}.a).
 
@@
 \includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
-\caption{Comparison of embedding by stanard and multiscale t-SNE on a data set from the mouse visual cortex~\cite{hrvatin2018}.
-{\bf (a)} Standard t-SNE places clusters arbitrarily.
-{\bf (b)} Augmenting t-SNE with multi-scale similarities provides a more meaningful layout of the clusters.
-Neuronal types occupy one region of the space.
-Oligodendrocyte precursor cells (OPCs) are mainly progenitors to oligodendrocytes, but may also differentiate into neurons or astrocytes.} \label{fig:multiscale}
+\caption{Comparison of embedding by stanard and multiscale t-SNE on a data set
+from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE
+places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
+similarities provides a more meaningful layout of the clusters. Neuronal types
+occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are
+mainly progenitors to oligodendrocytes, but may also differentiate into neurons
+or astrocytes.} \label{fig:multiscale}
 \end{figure}
 
-We have also observed an important role of gene selection in crafting the reference embedding spaces.
-We have found that when selecting an insufficient number of genes, the resulting visualizations display fragmented clusters.
-When the selection is too broad and includes lowly expressed genes, the subclusters tend to overlap.
-These effects can all be attributed to sparseness of the data sets and may be intrinsic for single-cell data.
-In our studies, we found that selection of 3,000 genes yields most informative visualizations (Fig.~\ref{fig:gene_selection}).
+We have also observed an important role of gene selection in crafting the
+reference embedding spaces. We have found that when selecting an insufficient
+number of genes, the resulting visualizations display fragmented clusters. When
+the selection is too broad and includes lowly expressed genes, the subclusters
+tend to overlap. These effects can all be attributed to sparseness of the data
+sets and may be intrinsic for single-cell data. In our studies, we found that
+selection of 3,000 genes yields most informative visualizations
+(Fig.~\ref{fig:gene_selection}).
 
@@
 \includegraphics[width=\textwidth]{figures/hrvatin_embedding_tsne_genes.pdf}
-\caption{Gene selection plays an important role when constructing the reference embedding.
-{\bf (a)} Using too few genes results in over-clustering.
-{\bf (b)} Using an intermediate number of genes reveals clustering mostly consistent with cell annotations.
-{\bf (c)} Including all the genes may lead to under-clustering of the more specialized cell types.}
+\caption{Gene selection plays an important role when constructing the reference
+embedding. {\bf (a)} Using too few genes results in over-clustering. {\bf (b)}
+Using an intermediate number of genes reveals clustering mostly consistent with
+cell annotations. {\bf (c)} Including all the genes may lead to
+under-clustering of the more specialized cell types.}
 \label{fig:gene_selection}
@@
 
-In principle, our theoretically-grounded embedding of secondary data into the scaffold defined by reference embedding could be simplified with the application of the nearest neighbors-based procedure.
-For example, while describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne} proposed to position new points into a known embedding by placing them in the median position of their 10 nearest neighbors, where the neighborhood was estimated in the original data space.
-Notice that we use this trick as well, but only for initialization of positions of new data instances that are subject to further optimization.
-In Fig.~\ref{fig:optimization} we demonstrate that nearest neighboring-based positioning is insufficient and may yield clumped visualizations where the optimal positioning using the t-SNE loss function is much more dispersed and rightfully shows a more considerable variation in the secondary data.
-Some data points may also fell into the neighboring space between differently typed clusters, while after optimization they typically converge closer to same-typed groups.
+In principle, our theoretically-grounded embedding of secondary data into the
+scaffold defined by reference embedding could be simplified with the
+application of the nearest neighbors-based procedure. For example, while
+describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
+proposed to position new points into a known embedding by placing them in the
+median position of their 10 nearest neighbors, where the neighborhood was
+estimated in the original data space. Notice that we use this trick as well,
+but only for initialization of positions of new data instances that are subject
+to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
+nearest neighboring-based positioning is insufficient and may yield clumped
+visualizations where the optimal positioning using the t-SNE loss function is
+much more dispersed and rightfully shows a more considerable variation in the
+secondary data. Some data points may also fell into the neighboring space
+between differently typed clusters, while after optimization they typically
+converge closer to same-typed groups.
 
@@
 \includegraphics[width=\textwidth]{figures/optimization_retina.pdf}
-\caption{Comparison of data placement by nearest neighbors approach from Kobak and Berens~\cite{art_of_using_tsne} and optimized placement by our algorithm.
-{\bf (a)} Data points are placed to the median position of their 10 nearest neighbors in the reference set.
-{\bf (b)} Point positions are optimized, revealing a different, more dispersed placement that better reflects the variety of cells in the secondary data set.}
+\caption{Comparison of data placement by nearest neighbors approach from Kobak
+and Berens~\cite{art_of_using_tsne} and optimized placement by our algorithm.
+{\bf (a)} Data points are placed to the median position of their 10 nearest
+neighbors in the reference set. {\bf (b)} Point positions are optimized,
+revealing a different, more dispersed placement that better reflects the
+variety of cells in the secondary data set.}
 \label{fig:optimization}
@@
 
-The proposed method assumes that all cell types from the secondary data set are present in the reference.
-The proposed method would fail to reveal novel cell types in the secondary data set, possibly positioning them arbitrarily close to unrelated clusters.
-Procedures such as scmap~\cite{scmap} were recently proposed to cope with such cases and identify the cells whose type is new and not included in the reference.
-Our procedure does not address such cases, and for scaling-up to a wider collection of cell types relies on emerging availability of large collections of the reference data such as those managed by Human Cell Atlas initiative~\cite{hca}.
+The proposed method assumes that all cell types from the secondary data set are
+present in the reference. The proposed method would fail to reveal novel cell
+types in the secondary data set, possibly positioning them arbitrarily close to
+unrelated clusters. Procedures such as scmap~\cite{scmap} were recently
+proposed to cope with such cases and identify the cells whose type is new and
+not included in the reference. Our procedure does not address such cases, and
+for scaling-up to a wider collection of cell types relies on emerging
+availability of large collections of the reference data such as those managed
+by Human Cell Atlas initiative~\cite{hca}.
 
@@
 
-The procedures described in this paper are provided as Python notebooks that are, together with the data, available in an open repository~\footnote{https://github.com/biolab/tsne-embedding}.
-All experiments were run using openTSNE~\footnote{https://github.com/pavlin-policar/openTSNE}, our open and extensible t-SNE library for Python.
+The procedures described in this paper are provided as Python notebooks that
+are, together with the data, available in an open
+repository~\footnote{https://github.com/biolab/tsne-embedding}. All experiments
+were run using openTSNE~\footnote{https://github.com/pavlin-policar/openTSNE},
+our open and extensible t-SNE library for Python.
 
@@
 
-Almost all recent publications of single-cell studies start with a two-dimensional visualization of the data that exposes the diversity as well as different types of cells from the study.
-While any dimensionality reduction technique can be used to render such a visualization, different variants of t-SNE are most often used.
-Due to the ability to explore biological mechanisms at the cellular level, single-cell studies are increasingly widespread, and their publications in the past couple of years are abundant.
-One of the central tasks in single-cell studies is the classification of new cells based on finding from previous studies.
-Such transfer of knowledge is often difficult due to batch effects present in data from different sources.
-Solving the problem of the batch-effects together with prevailing means of presenting single-cell data in two-dimensional visualizations motivated the research presented in this paper.
+Almost all recent publications of single-cell studies start with a
+two-dimensional visualization of the data that exposes the diversity as well as
+different types of cells from the study. While any dimensionality reduction
+technique can be used to render such a visualization, different variants of
+t-SNE are most often used. Due to the ability to explore biological mechanisms
+at the cellular level, single-cell studies are increasingly widespread, and
+their publications in the past couple of years are abundant. One of the central
+tasks in single-cell studies is the classification of new cells based on
+finding from previous studies. Such transfer of knowledge is often difficult
+due to batch effects present in data from different sources. Solving the
+problem of the batch-effects together with prevailing means of presenting
+single-cell data in two-dimensional visualizations motivated the research
+presented in this paper.
 
-Our proposed approach uses t-SNE embedding as a scaffold for the positioning of new cells within the visualization, and possibly for aiding their classification.
-The three case studies incorporating pairs of data sets from different domains but with similar classifications demonstrate that our proposed procedure can effectively deal with batch effects to construct visualizations that correctly map a secondary data set onto a reference data set from an independent study that possibly uses different data collection protocol.
-While we here focused on t-SNE constructed from reference data sets, the approach could be applied to any existing two-dimensional visualization.
+Our proposed approach uses t-SNE embedding as a scaffold for the positioning of
+new cells within the visualization, and possibly for aiding their
+classification. The three case studies incorporating pairs of data sets from
+different domains but with similar classifications demonstrate that our
+proposed procedure can effectively deal with batch effects to construct
+visualizations that correctly map a secondary data set onto a reference data
+set from an independent study that possibly uses different data collection
+protocol. While we here focused on t-SNE constructed from reference data sets,
+the approach could be applied to any existing two-dimensional visualization.
 
@@
 
-This work was supported by the Slovenian Research Agency Program Grant P2-0209, and by the BioPharm.SI project supported from European Regional Development Fund and the Slovenian Ministry of Education, Science and Sport.
-We would also like to thank Dmitry Kobak for helpful discussions on t-SNE extensions.
+This work was supported by the Slovenian Research Agency Program Grant P2-0209,
+and by the BioPharm.SI project supported from European Regional Development
+Fund and the Slovenian Ministry of Education, Science and Sport. We would also
+like to thank Dmitry Kobak for helpful discussions on t-SNE extensions.
 

### 2019-06-13 paper/tsne_embedding.tex
Corrections and edits.

@@
 visualizations of high-dimensional data. When working with multiple data sets,
-the straightforward application of these methods often fails, and instead of
+a straightforward application of these methods often fails; instead of
 revealing underlying classes, the resulting visualizations expose
-domain-specific clusters. To circumvent these batch effects, we here propose an
-embedding procedure that takes a t-SNE visualization constructed on reference
-data set and uses it as a scaffold for new embeddings. New, secondary data is
-embedded one data-instance at the time, thus disregarding any interactions
-between data items in the secondary data and implicitly mitigating batch
-effects. We demonstrate the utility of this procedure on analysis of six
+data set-specific clusters. To circumvent these batch effects, we propose an
+embedding procedure that takes a t-SNE visualization constructed on the reference
+data set and uses it as a scaffold for embedding new data. New, secondary data is
+embedded one data-instance at the time. This prevents any interactions
+between data instances in the secondary data and implicitly mitigates batch
+effects. We demonstrate the utility of this approach on analysis of six
 recently published single-cell gene expression data sets containing up to tens
@@
 and was obtained using different experimental protocols. The visualizations
-constructed by the proposed approach are absent of batch effects, and the cells
+constructed by the proposed approach are cleared of batch effects, and the cells
 from secondary data sets correctly co-cluster with cells from the primary data
@@
 Two-dimensional embeddings and their visualizations may assist in the analysis
-and interpretation of multi-dimensional data. Intuitively, two data instances
+and interpretation of high-dimensional data. Intuitively, two data instances
 should be co-located in the resulting visualization if their multi-dimensional
@@
 component analysis (PCA) and multi-dimensional
-scaling~\cite{distill,umap_single_cell}. While useful for visualizing the data
+scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
 from a single coherent source, these methods may encounter problems if the data
 comes from different sources. Here, if executing the dimensionality reduction
-procedures on a merged data, the resulting visualizations would typically
-depict source-specific clusters instead of reveal clusters that are
-semantically enriched, say, with same-type of data instances across different
-sources. This source-specific confounding is often referred to as {\em domain
+procedures on merged data, the resulting visualizations would often
+depict source-specific clusters instead of 
+grouping same classes of data which are shared across different
+sources. This source-specific confounding is commonly referred to as {\em domain
 shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
-{\em dataset shift}~\cite{dataset_shift}. In the bioinformatics literature, the
+{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
 domain-specific differences are more commonly referred to as {\em batch
@@
 
-Massive, multi-variate biological data sets typical suffer from source-specific
+Massive, multi-variate biological data sets often suffer from these source-specific
 biases. Consider an example from single-cell genomics, a domain we will focus
@@
 --- selected also due to availability and the abundance of recently published
-data. Single-cell data sets may include thousands of cells whose profiles
-consist of tens of thousands of cell-specific gene expressions. Single cell
-studies typically start with the analysis of cell types, where we expect that
-cells of the same type would cluster in two-dimensional data
+data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
+which serves as an estimate of the expression of cell's genes.
+Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
+typically start with the analysis of cell types. Here, it is generally expected that
+cells of the same type would cluster together in two-dimensional data
 visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
@@
 the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which
-renders the data from human cell atlas~\cite{baron2016} and similarly-typed
-cells from diabetic patients~\cite{xin2016}. Just like with data from brain
-cells, pancreas cells cluster first according to the data source, again
-resulting in an uninformative visualization that primarily exposes the batch
-effect.
+renders the data from human cell atlas~\cite{baron2016} and
+related cells from diabetic patients~\cite{xin2016}. Just like with data from brain
+cells, pancreas cells cluster primarily by data source, again
+resulting in an uninformative visualization driven by batch
+effects.
 
@@
 Current solutions to embedding the data from various data sources address the
-batch effect problems up-front. The data is typically pre-processed and
-transformed, possibly to the latent space, so that the batch effects are
-removed. The recently proposed procedures for batch-effect removal include
-canonical correlation analysis~\cite{cca} and mutual
+batch effect problems upfront. The data is first preprocessed and
+transformed such that the batch effects are explicitly
+removed. Recently proposed procedures for batch effect removal include
+Canonical correlation analysis~\cite{cca} and Mutual
 nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of
@@
 preprocessing which may blur the boundaries between cell types. Another problem
-is also the inclusion of any new data, for which the entire data analysis
-pipeline has to be rerun, usually resulting in different layout and clusters
+is also the inclusion of any new data, for which the entire analysis
+pipeline has to be rerun, potentially resulting in different layout and clusters
 that have little resemblance to original visualization and thus require
@@
 batch effects. Our approach considers one of the data sets as a {\em reference}
-and aims to visually classify the cells in the other, {\em secondary data set}.
+and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
 We construct the t-SNE embedding using the reference data set, and then use it
-as a scaffold for the embedding of data points from the secondary data.
-Embedding is performed one data point at a time. Independence of each new
-embedding of data instances from secondary data set causes clustering landscape
-to only depend on a reference scaffold, thus removing data source-driven
+as a scaffold for the embedding of secondary data points.
+A key idea underpinning our approach is that the embedding is performed one data point at a time. 
+Independence of each new
+embedding of data instances from secondary data set causes the clustering landscape
+to only depend on the reference scaffold, implicitly removing data source-driven
 variation. In other words, when including new data, the scaffold inferred from
-the reference data set is kept unchanged and defines the ``gravitational
-field'' to independently embedded each new data item. For example, in
+the reference data set is kept unchanged and defines a ``gravitational
+field'', independently driving the embedding of each new instance. For example, in
 Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
@@
 
-At the core of the proposed solution is a mapping that implements embedding of
-new data to existing t-SNE visualization. In the following, we introduce t-SNE,
-then describe its recently proposed multi-scale extension, and outline our
-algorithm that embeds new data into a previously trained scaffold. While a
+The proposed solution is implemented using a mapping of
+new data to an existing t-SNE visualization. While the 
 utility of such an algorithm was already hinted to in recent
-publication~\cite{art_of_using_tsne}, we here provide its practical and
+publication~\cite{art_of_using_tsne}, we hereby provide its practical and
 theoretically-grounded implementation. Considering the abundance of recent
-publications in batch effect removal, we present surprising evidence that a
-computationally more direct and elegant embedding procedure solves the batch
+publications on batch effect removal, we present surprising evidence that a
+computationally more direct and principled embedding procedure solves the batch
 effects problem when constructing interpretable visualizations from different
@@
 
-We describe an end-to-end pipeline that uses t-SNE embeddings as a scaffold for
-new data samples, and enables visualisation of data from different sources
+We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a scaffold for
+embedding new (secondary) data, enabling joint visualisation of multiple data sources
 while mitigating batch effects. Our proposed approach starts with t\nobreakdash
--SNE to embedd a reference data set, with the aim of constructing a
+-SNE to embed the reference data set, with the aim of constructing a
 two-dimensional visualisation to facilitate interpretation and cluster
-classification. We regard the resulting two-dimensional embedding as a scaffold
-to which we add samples from the new, secondary data source. Each new sample is
-then placed into the reference embedding and optimized independently using the
-t\nobreakdash -SNE loss function. Independent embedding of each data item from
-a secondary data set disregards any interactions present in that data, and
-prevents forming clusters that would be data-souce specific. Below, we start
+classification. Then, the placement of each new sample is
+optimized independently via the
+t\nobreakdash -SNE loss function. Independent treatment of each data instance from
+a secondary data set disregards any interactions present in that data set, and
+prevents forming clusters that would be specific to the secondary data. Below, we start
 with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing
@@
 
-The embedding by t-SNE is local and non-linear, tailored to visualisation of
-high dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
+The non-linear t-SNE embedding emphasizes local similarities between data instances, 
+as it is tailored to visualisation of high dimensional data sets. 
+Given a multi-dimensional data set $\mathbf{X} =
 \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
@@
 
-\noindent where $\mathcal{D}$ is some distance measure. This is then symmetrized to
+\noindent where $\mathcal{D}$ is a distance measure. This is then symmetrized to
 
@@
 The similarity between points $\mathbf{y}_i$ and $\mathbf{y}_j$ in the
-embedding space are defined using the $t$-distribution with one degree of
+embedding space is defined using the $t$-distribution with one degree of
 freedom
@@
 
-The data transformation by t-SNE finds an embedding $\mathbf{Y}$ that minimizes
+The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes
 the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
@@
 the algorithm impractical for any reasonably-sized data. To address larger data
-sets, we instead adapt a recent approach for low-rank approximation of
+sets, we adopt a recent approach for low-rank approximation of
 similarity matrix based on polynomial interpolation that reduces the time
 complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
-visualization of massive data sets possibly containing millions of data
+visualization of massive data sets, possibly containing millions of data
 points~\cite{fi_tsne}.
 
-The t-SNE embeddings substantially depend on the value of the perplexity
-parameter. The perplexity parameter can be interpreted as the number of
+The resulting embeddings substantially depend on the value of the perplexity
+parameter. Perplexity can be interpreted as the number of
 neighbors for which the distances in the embedding space are preserved. Small
@@
 scale and organizing clusters in a more meaningful way. Larger values of
-perplexity may often result in merging multiple small clusters, thus obscuring
+perplexity can lead to merging of multiple small clusters, thus obscuring
 local aspects of the data~\cite{art_of_using_tsne}.
@@
 resulting embeddings are visually very similar to those obtained with the
-approach described above (for brevity, data not shown).
+approach described above (not shown for brevity).
 
@@
 
-An algorithm to embedd new data point to a reference embedding consists of
+An algorithm to embed a new data point to a reference embedding consists of
 estimating similarities between the point and the reference data and optimizing
@@
 such as principal component analysis or autoencoders, t-SNE does not define an
-explicit mapping to embedding space, and embeddings need to be found through
+explicit mapping to the embedding space, and embeddings need to be found through
 loss function optimization.
 
-For each new data point, we first estimate its distance to each of the data
-instances in the reference data set. We initialize the position of the data
-point in embedding space to the median embedding position of its $k$ nearest
-neighbors from the reference data set. While we found the algorithm to be
+The position of a new data
+point in embedding space is initialized 
+to the median reference embedding position of its $k$ nearest 
+neighbors. While we found the algorithm to be
 robust to choices of $k$, we use $k=10$ in our experiments.
@@
 In the optimization step, we refine point positions using batch gradient
-descent. We use an adaptive learning rate scheme with momentum proposed by
-Jacobs to speed up the convergence~\cite{momentum}. We run gradient descent
+descent. We use an adaptive learning rate scheme with momentum 
+to speed up the convergence, as proposed by Jacobs~\cite{momentum}. We run gradient descent
 with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
@@
 from the reference embedding. This phenomenon is caused due to the embedding to
-already defined t-SNE space, where the distances between data points and
+a previously defined t-SNE space, where the distances between data points and
 corresponding gradients of the optimization function may be quite large. When
@@
 0.0001. The resulting gradients tend to be very small during the initial phase,
-resulting in stable convergence. When embedding new samples, the range of the
+resulting in stable convergence. When embedding new instances, the range of the
 embedding is much larger, resulting in much larger gradients, and the default
 learning rate causes points to move very far from the reference embedding. In
-our experiments, we found that lowering the learning rate to $\eta \sim 0.1$
+our experiments, we found that decreasing the learning rate to $\eta \sim 0.1$
 produces stable solutions. This is especially important when using the
@@
 increasing the runtime. The reduced learning rate suppresses this issue, and
-does not slow the convergence because of an adaptive learning rate scheme and
-running the optimization for a sufficient number of steps.
+does not slow the convergence because of an adaptive learning rate scheme, provided
+the optimization is run for a sufficient number of steps.
 
@@
 
-We apply the proposed transfer learning approach to t-SNE visualizations of
-single-cell data. In single-cell data sets, the data includes the variety of
-cells from specific tissue and characterizes the cells through the expression
-of its genes. In experiments, we considered several recently published data
+We examine the utility of the proposed approach with t-SNE visualizations of
+single-cell data. Data in this realm include a variety of
+cells from specific tissues, and characterize them through gene expression. 
+We considered several recently published data
 sets where cells were annotated with the cell type. Our aim was to construct
-t-SNE visualizations where similarly-typed cells would cluster together,
-despite the differences between data sources. Below, we enlist the data sets,
+t-SNE visualizations where cells of similar types would cluster together,
+despite systematic differences between data sources. Below, we enlist the data sets,
 describe single-cell specific data preprocessing procedures, and display the
@@
 studies. To facilitate comparisons, the cell annotations were harmonized using
-the cell type annotations from cell ontology~\cite{cell_ontology}. Notice that
+the cell type annotations from Cell ontology~\cite{cell_ontology}. Notice that
 different RNA sequencing protocols were used to estimate gene expressions. We
@@
 
-\noindent In our experiments we use $t=0$ and $a=1$.
+\noindent In our experiments, we use $t=0$ and proceed with a subset of
+$\hat{G}$ genes.
 
@@
 reference cluster. For instance, the astrocyte cells marked with red on the
-right were mapped to an oval cluster of same-typed cells denoted with the same
+right were mapped to an oval cluster of the same type, with cells denoted using the same
 color in the visualization on the left.} \label{fig:transform_brain}
@@
 
-There are a few tricks that we use in our procedure for t-SNE embedding of the
+There are additional modifications that we use in the embedding of the
 secondary data set that were proposed recently and enhance the original t-SNE.
-An important one is a multi-scale extension that besides local ordering of the
-data points takes care about global optimization of the cluster placement. For
+An important one is a multi-scale extension that, besides local ordering of the
+data points, globally optimizes the cluster placement. For
 illustration, consider visualizations with standard and multi-scale t-SNE in
@@
 median position of their 10 nearest neighbors, where the neighborhood was
-estimated in the original data space. Notice that we use this trick as well,
+estimated in the original data space. Notice that we use this approach as well,
 but only for initialization of positions of new data instances that are subject
@@
 between differently typed clusters, while after optimization they typically
-converge closer to same-typed groups.
+converge closer to groups of the same type.
 
@@
 types in the secondary data set, possibly positioning them arbitrarily close to
-unrelated clusters. Procedures such as scmap~\cite{scmap} were recently
+unrelated clusters. Procedures such as Scmap were recently
 proposed to cope with such cases and identify the cells whose type is new and
-not included in the reference. Our procedure does not address such cases, and
+not included in the reference~\cite{scmap}. Our procedure does not address such cases, and
 for scaling-up to a wider collection of cell types relies on emerging
@@
 
-Almost all recent publications of single-cell studies start with a
+A large majority of recent publications on single-cell studies start with a
 two-dimensional visualization of the data that exposes the diversity as well as
-different types of cells from the study. While any dimensionality reduction
+different types of cells in the study. While any dimensionality reduction
 technique can be used to render such a visualization, different variants of
 t-SNE are most often used. Due to the ability to explore biological mechanisms
-at the cellular level, single-cell studies are increasingly widespread, and
+at the individual cell level, single-cell studies are increasingly widespread, and
 their publications in the past couple of years are abundant. One of the central

### 2019-06-13 paper/tsne_embedding.tex

@@
 % Fixup hyphenation
-\hyphenation{cor-res-pon-din}
+\hyphenation{cor-res-pon-ding}
 \hyphenation{cor-re-lat-tion}
@@
 from a single coherent source, these methods may encounter problems if the data
-comes from different sources. Here, if executing the dimensionality reduction
-procedures on merged data, the resulting visualizations would often
-depict source-specific clusters instead of 
-grouping same classes of data which are shared across different
-sources. This source-specific confounding is commonly referred to as {\em domain
+comes from different sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
+depict source-specific clusters instead of revealing clusters that are
+semantically enriched, say, consisting of data instances originating from
+different sources from the same class-type.. This source-specific confounding is often referred to as {\em domain
 shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
@@
 visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
-t-SNE embedded data from mouse brain cells originating from visual
+t-SNE embedded data from mouse brain cells originating from the visual
 cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
@@
 Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
-studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
-the exploration of pancreas cells in Fig.~\ref{fig:batch_effect}.b, which
-renders the data from human cell atlas~\cite{baron2016} and
-related cells from diabetic patients~\cite{xin2016}. Just like with data from brain
-cells, pancreas cells cluster primarily by data source, again
+studies to overlap. Batch effects similarly hinder the utility of t-SNE in
+the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
+renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
+cells from diabetic patients~\cite{xin2016}. Just like with data from brain
+cells, pancreatic cells cluster primarily by data source, again
 resulting in an uninformative visualization driven by batch
-effects.
+effect.
 
@@
 Current solutions to embedding the data from various data sources address the
-batch effect problems upfront. The data is first preprocessed and
+batch effect problems up-front. The data is typically preprocessed and
 transformed such that the batch effects are explicitly
-removed. Recently proposed procedures for batch effect removal include
-Canonical correlation analysis~\cite{cca} and Mutual
-nearest-neighbors~\cite{mnn,seurat}. In these works, a proof for the success of
-batch effect removal is a good mixing of cells from different sources in a
-t-SNE visualization. Elimination of batch effects may require aggressive data
+removed. Recently proposed procedures for batch-effect removal include
+canonical correlation analysis~\cite{cca} and mutual
+nearest-neighbors~\cite{mnn,seurat}.
+In these works, batch effects are deemed removed when good mixing of cells from different sources is achieved in a t-SNE visualization.
+Elimination of batch effects may require aggressive data
 preprocessing which may blur the boundaries between cell types. Another problem
-is also the inclusion of any new data, for which the entire analysis
-pipeline has to be rerun, potentially resulting in different layout and clusters
+is also the inclusion of any new data, for which the entire data analysis
+pipeline must be rerun, usually resulting in a different layout and clusters
 that have little resemblance to original visualization and thus require
@@
 and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
-We construct the t-SNE embedding using the reference data set, and then use it
-as a scaffold for the embedding of secondary data points.
+We construct the t\nobreakdash -SNE embedding using the reference data set, and then use it
+as a scaffold for the embedding of data points from the secondary data.
 A key idea underpinning our approach is that the embedding is performed one data point at a time.
 Independence of each new
-embedding of data instances from secondary data set causes the clustering landscape
-to only depend on the reference scaffold, implicitly removing data source-driven
+embedded data instance from the secondary data set causes clustering the landscape
+to depend only on the reference scaffold, thus removing data source-driven
 variation. In other words, when including new data, the scaffold inferred from
@@
 scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
-hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t-SNE
-visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamus cells are
+hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
+visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
 dispersed across entire embedding space and their cell type correctly matches
@@
 
+\begin{figure}[htb]
+\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
+\caption{A two-dimensional embedding of a reference containing brain cells (a) and
+the corresponding mapping of secondary data containing hypothalamic cells (b). Notice
+that the majority of hypothalamic cells were mapped to their corresponding
+reference cluster. For instance, astrocyte cells marked with red on the
+right were mapped to an oval cluster of same-typed cells denoted with the same
+color in the visualization on the left.} \label{fig:transform_brain}
+\end{figure}
+
 The proposed solution is implemented using a mapping of
-new data to an existing t-SNE visualization. While the 
-utility of such an algorithm was already hinted to in recent
-publication~\cite{art_of_using_tsne}, we hereby provide its practical and
+new data into an existing t-SNE visualization. While the
+utility of such an algorithm was already hinted at in recent
+publication~\cite{art_of_using_tsne}, here we provide its practical and
 theoretically-grounded implementation. Considering the abundance of recent
@@
 while mitigating batch effects. Our proposed approach starts with t\nobreakdash
--SNE to embed the reference data set, with the aim of constructing a
+-SNE to embed a reference data set, with the aim of constructing a
 two-dimensional visualisation to facilitate interpretation and cluster
@@
 prevents forming clusters that would be specific to the secondary data. Below, we start
-with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, emphasizing
-the notation and relevant parts upon which we base our secondary data embedding
+with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, introducing
+the relevant notation, upon which we base our secondary data embedding
 approach (Sec.~\ref{sec:transfer}).
@@
 
-The non-linear t-SNE embedding emphasizes local similarities between data instances, 
-as it is tailored to visualisation of high dimensional data sets. 
-Given a multi-dimensional data set $\mathbf{X} =
+t-SNE is local, non-linear dimensionality reduction method, tailored to the visualisation of
+high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
 \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
@@
 The time complexity needed to evaluate the similarities in
-Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the applications of
+Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of
 the algorithm impractical for any reasonably-sized data. To address larger data
 sets, we adopt a recent approach for low-rank approximation of
-similarity matrix based on polynomial interpolation that reduces the time
+gradients based on polynomial interpolation which reduces the time
 complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
@@
 neighbors for which the distances in the embedding space are preserved. Small
-values of perplexity result in tightly-packed clusters of points and lead to
-ignoring the long-range interactions between clusters. Larger values may result
+values of perplexity result in tightly-packed clusters of points and effectively
+ignore the long-range interactions between clusters. Larger values may result
 in a more globally consistent visualisations, preserving distances on a large
@@
 
-A trade-off between the local organization and global consistency may be
+The trade-off between the local organization and global consistency may be
 achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
@@
 
-An algorithm to embed a new data point to a reference embedding consists of
-estimating similarities between the point and the reference data and optimizing
-the position of the data point in the embedding space. Unlike parametric models
+Our algorithm, which embeds new data points to a reference embedding, consists of
+estimating similarities between each new point and the reference data and optimizing
+the position of each data point in the embedding space. Unlike parametric models
 such as principal component analysis or autoencoders, t-SNE does not define an
@@
 corresponding gradients of the optimization function may be quite large. When
-running s standard t-SNE, points are initialized and scaled to have variance
+running standard t-SNE, points are initialized and scaled to have variance
 0.0001. The resulting gradients tend to be very small during the initial phase,
-resulting in stable convergence. When embedding new instances, the range of the
+resulting in stable convergence. When embedding new samples, the span of the
 embedding is much larger, resulting in much larger gradients, and the default
@@
 produces stable solutions. This is especially important when using the
-interpolation-based approximation. Our approach places a grid of interpolation
+interpolation-based approximation, which places a grid of interpolation
 points over the embedding space, where the number of grid points is determined
-by the embedding range. Clearly, if even one point ``shoots off'' far from the
-embedding, the number of required grid points escalates, substantially
+by the span of the embedding. Clearly, if even one point ``shoots off'' far from the
+embedding, the number of required grid points may grow, substantially
 increasing the runtime. The reduced learning rate suppresses this issue, and
-does not slow the convergence because of an adaptive learning rate scheme, provided
+does not slow the convergence because of the adaptive learning rate scheme, provided
 the optimization is run for a sufficient number of steps.
@@
 
-We examine the utility of the proposed approach with t-SNE visualizations of
+We apply the proposed approach to t-SNE visualizations of
 single-cell data. Data in this realm include a variety of
 cells from specific tissues, and characterize them through gene expression.
-We considered several recently published data
+In our experiments, we considered several recently published data
 sets where cells were annotated with the cell type. Our aim was to construct
-t-SNE visualizations where cells of similar types would cluster together,
-despite systematic differences between data sources. Below, we enlist the data sets,
+t-SNE visualizations where similarly-typed cells would cluster together,
+despite systematic differences between data sources. Below, we list the data sets,
 describe single-cell specific data preprocessing procedures, and display the
@@
 (Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
-all cases, we relied on quality control and annotations from the original
-studies. To facilitate comparisons, the cell annotations were harmonized using
-the cell type annotations from Cell ontology~\cite{cell_ontology}. Notice that
+all cases, we relied on the quality control and annotations from the original
+publication. To facilitate comparisons, the cell annotations were harmonized using
+cell type annotations from the cell ontology~\cite{cell_ontology}. Notice that
 different RNA sequencing protocols were used to estimate gene expressions. We
-here report the number of cell types from each data set retained after
+report the number of cell types from each data set retained after
 preprocessing. Single-cell data is sparse, typically containing less than 10\%
@@
 
-The cells in the data sets were captured from the following three tissues:
+The cells in the data sets originates from the following three tissues:
 \begin{description}
@@
 to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
-consisting of selection of 3,000 representative genes (see
+consisting of the selection of 3,000 representative genes (see
 Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
@@
 similarities between the secondary data item and reference data points. The
-similarities are estimated using cosine similarity. We note that similarity is
+similarities are estimated using cosine similarity. We note that similarities are
 computed using the raw count matrices. The preprocessing stages are detailed in
@@
 
-Single-cell data sets suffer from a high level of technical noise and low
+Single-cell data sets suffer from high levels of technical noise and low
 capture efficiency, resulting in sparse expression matrices~\cite{umi}. To
@@
 \begin{equation}
-m_g = \left \langle \, \log_2 X_{ig} \mid X_{ig} > t \, \right \rangle.
+m_g = \left \langle \, \log_2 X_{ig} \mid X_{ig} > 0 \, \right \rangle.
 \end{equation}
@@
 
-\noindent In our experiments, we use $t=0$ and proceed with a subset of
-$\hat{G}$ genes.
-
 \subsection{Results and Discussion}
@@
 proposed approach. There are some deviations to these observations; for
-instance, in Fig.~\ref{fig:transform_brain} several oligodendrocytes precursor
-cells (OPC) were mapped to oligodendrocytes. This may be due to differences in
+instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
+cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
 annotation criteria by different authors, or due to inherent similarities of
@@
 other figures as well, but they are not common and constitute less then 5\% of
-the cells (less than \%1 for pancreas, \%2 for retina and 5\% for brain
+the cells (less than  5\% in brain, \%1 in pancreas and \%2 in retina
 secondary data sets).
@@
 Notice that we could simulate the split between reference and secondary data
-sets by cross-validation using one data set only, as this type of experiments
+sets using one data set only and perform cross-validation, however this type of experiment
 would not incorporate batch effects. We want to remind the reader that handling
-batch effects were central to our endeavor and that disregard of this effect
-could lead to data visualizations strikingly different from ours. For example,
+batch effects were central to our endeavour and that the disregard of this effect
+could lead to overly-optimistic results and data visualizations strikingly different from ours. For example,
 compare the visualisations from Fig~\ref{fig:batch_effect}.a and
@@
 
-\begin{figure}[htbp]
-\includegraphics[width=\textwidth]{figures/transform_brain.pdf}
-\caption{Two-dimensional embedding of a reference brain cells (a) and
-corresponding mapping of secondary data from hypothalamus cells (b). Notice
-that the majority of hypothalamus cells were mapped to their corresponding
-reference cluster. For instance, the astrocyte cells marked with red on the
-right were mapped to an oval cluster of the same type, with cells denoted using the same
-color in the visualization on the left.} \label{fig:transform_brain}
-\end{figure}
-
-% FIGURE: 2. t-SNE transform use-case
-\begin{figure}[htbp]
+\begin{figure}[htb]
 \includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
-\caption{Embedding of pancreas cells from Baron \etal~\cite{baron2016} and
+\caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
 cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
-Fig~\ref{fig:transform_brain} vast majority of the cells from the secondary
+Fig~\ref{fig:transform_brain} the vast majority of the cells from the secondary
 data set were correctly mapped to the same-typed cluster of reference
@@
 
-\begin{figure}[htbp]
+\begin{figure}[htb]
 \includegraphics[width=\textwidth]{figures/transform_retina.pdf}
-\caption{Embedding of a large reference retina cells from Macosco
+\caption{An embedding of a large reference of retinal cells from Macosco
 \etal~\cite{macosko2015} (a) and mapping of cells from a smaller study that
 focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b). We use
-colors consistent with the study by Macosko \etal. Notice that Shekhar includes
-cells of only four different types.} \label{fig:transform_retina}
+colors consistent with the study by Macosko \etal.} \label{fig:transform_retina}
 \end{figure}
@@
 There are additional modifications that we use in the embedding of the
-secondary data set that were proposed recently and enhance the original t-SNE.
-An important one is a multi-scale extension that, besides local ordering of the
-data points, globally optimizes the cluster placement. For
+secondary data set that were recently proposed and enhance the original t-SNE visualization.
+One important extension is the use of multi-scale similarities that, besides local ordering of the
+data points, includes global optimization of cluster placement. For
 illustration, consider visualizations with standard and multi-scale t-SNE in
 Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE
-(Fig.~\ref{fig:multiscale}.b) the clusters with neuron cells are clumped
+(Fig.~\ref{fig:multiscale}.b) the clusters with neuronal cells are clumped
 together, while their placement in standard t-SNE is arbitrary
@@
 \includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
-\caption{Comparison of embedding by stanard and multiscale t-SNE on a data set
+\caption{A comparison of standard and multi-scale t-SNE on data
 from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE
@@
 
-We have also observed an important role of gene selection in crafting the
-reference embedding spaces. We have found that when selecting an insufficient
-number of genes, the resulting visualizations display fragmented clusters. When
+We also observed the important role of gene selection in crafting the
+reference embedding spaces. We found that when selecting an insufficient
+number of genes, the resulting visualizations display overly-fragmented clusters. When
 the selection is too broad and includes lowly expressed genes, the subclusters
 tend to overlap. These effects can all be attributed to sparseness of the data
-sets and may be intrinsic for single-cell data. In our studies, we found that
+sets and may be intrinsic to single-cell data. In our studies, we found that
 selection of 3,000 genes yields most informative visualizations
@@
 \caption{Gene selection plays an important role when constructing the reference
-embedding. {\bf (a)} Using too few genes results in over-clustering. {\bf (b)}
+embedding. {\bf (a)} Using too few genes results in fragmented clusters. {\bf (b)}
 Using an intermediate number of genes reveals clustering mostly consistent with
@@
 In principle, our theoretically-grounded embedding of secondary data into the
-scaffold defined by reference embedding could be simplified with the
+scaffold defined by the reference embedding could be simplified with the
 application of the nearest neighbors-based procedure. For example, while
@@
 estimated in the original data space. Notice that we use this approach as well,
-but only for initialization of positions of new data instances that are subject
+but only for the initialization of positions of new data instances that are subject
 to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
-nearest neighboring-based positioning is insufficient and may yield clumped
+nearest neighbor-based positioning is insufficient and may yield clumped
 visualizations where the optimal positioning using the t-SNE loss function is
 much more dispersed and rightfully shows a more considerable variation in the
-secondary data. Some data points may also fell into the neighboring space
-between differently typed clusters, while after optimization they typically
-converge closer to groups of the same type.
+secondary data. Some data points may also fall into the empty regions
+between different clusters, while after optimization they typically
+move closer to same-typed groups.
 
@@
 \includegraphics[width=\textwidth]{figures/optimization_retina.pdf}
-\caption{Comparison of data placement by nearest neighbors approach from Kobak
-and Berens~\cite{art_of_using_tsne} and optimized placement by our algorithm.
+\caption{Comparison of data placement using the nearest neighbors approach from Kobak
+and Berens~\cite{art_of_using_tsne} and the optimized placement using our algorithm.
 {\bf (a)} Data points are placed to the median position of their 10 nearest
@@
 
-A large majority of recent publications on single-cell studies start with a
-two-dimensional visualization of the data that exposes the diversity as well as
-different types of cells in the study. While any dimensionality reduction
+Almost all recent publications of single-cell studies begin with a
+two-dimensional visualization of the data that reveals the cellular diversity containing many 
+different cell-types from the study. While any dimensionality reduction
 technique can be used to render such a visualization, different variants of
@@
 tasks in single-cell studies is the classification of new cells based on
-finding from previous studies. Such transfer of knowledge is often difficult
-due to batch effects present in data from different sources. Solving the
-problem of the batch-effects together with prevailing means of presenting
-single-cell data in two-dimensional visualizations motivated the research
+findings from previous studies. Such transfer of knowledge is often difficult
+due to batch effects present in data from different sources.
+Addressing batch effects by adapting and extending t-SNE, the prevailing method used
+to present single-cell data in two-dimensional visualization motivated the research
 presented in this paper.
 
-Our proposed approach uses t-SNE embedding as a scaffold for the positioning of
-new cells within the visualization, and possibly for aiding their
+Our proposed approach uses a t-SNE embedding as a scaffold for the positioning of
+new cells within the visualization, and possibly for aiding in their
 classification. The three case studies incorporating pairs of data sets from
@@
 proposed procedure can effectively deal with batch effects to construct
-visualizations that correctly map a secondary data set onto a reference data
-set from an independent study that possibly uses different data collection
-protocol. While we here focused on t-SNE constructed from reference data sets,
-the approach could be applied to any existing two-dimensional visualization.
+visualizations that correctly map secondary data sets onto a reference data
+set from an independent study that possibly uses different experimental
+protocol. While we focused here on reference visualizations constructed using t-SNE,
+this approach can be applied using any existing two-dimensional visualization.
 
@@
 Fund and the Slovenian Ministry of Education, Science and Sport. We would also
-like to thank Dmitry Kobak for helpful discussions on t-SNE extensions.
+like to thank Dmitry Kobak for discussions on t-SNE.
 

### 2019-06-13 paper/tsne_embedding.tex

@@
 data set-specific clusters. To circumvent these batch effects, we propose an
-embedding procedure that takes a t-SNE visualization constructed on the reference
-data set and uses it as a scaffold for embedding new data. New, secondary data is
-embedded one data-instance at the time. This prevents any interactions
-between data instances in the secondary data and implicitly mitigates batch
-effects. We demonstrate the utility of this approach on analysis of six
+embedding procedure that takes a t-SNE visualization constructed on a reference
+data set and uses it as a scaffold for embedding new data. The new, secondary data is
+embedded one data-point at the time. This prevents any interactions
+between instances in the secondary data and implicitly mitigates batch
+effects. We demonstrate the utility of this approach with an analysis of six
 recently published single-cell gene expression data sets containing up to tens
 of thousands of cells and thousands of genes. In these data sets, the batch
-effects were particularly strong as the data comes from different institutions
+effects are particularly strong as the data comes from different institutions
 and was obtained using different experimental protocols. The visualizations
-constructed by the proposed approach are cleared of batch effects, and the cells
+constructed by our proposed approach are cleared of batch effects, and the cells
 from secondary data sets correctly co-cluster with cells from the primary data
@@
 should be co-located in the resulting visualization if their multi-dimensional
-profiles are similar. For this task, non-linear local embeddings such as
-t-distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
+profiles are similar. For this task, non-linear embedding techniques such as
+t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
 manifold approximation and projection~\cite{umap} have recently complemented
@@
 from a single coherent source, these methods may encounter problems if the data
-comes from different sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
-depict source-specific clusters instead of revealing clusters that are
-semantically enriched, say, consisting of data instances originating from
-different sources from the same class-type.. This source-specific confounding is often referred to as {\em domain
+comes from multiple sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
+reveal source-specific clusters instead of grouping data instances of the same class-type, regardless of data sources. This source-specific confounding is often referred to as {\em domain
 shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
@@
 Massive, multi-variate biological data sets often suffer from these source-specific
-biases. Consider an example from single-cell genomics, a domain we will focus
+biases. Consider an example from single-cell genomics, the domain we will focus
 on in this manuscript and that was --- besides current scientific challenges
---- selected also due to availability and the abundance of recently published
+--- selected also due to the availability and abundance of recently published
 data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
-which serves as an estimate of the expression of cell's genes.
+which serve as an estimate of the expression of cell's genes.
 Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
@@
 t-SNE embedded data from mouse brain cells originating from the visual
-cortex~\cite{hrvatin2018} and hypothalamus~\cite{chen2017}. The figure reveals
+cortex~\cite{hrvatin2018} and the hypothalamus~\cite{chen2017}. The figure reveals
 distinct clusters but also separates the data from the two brain regions. These
@@
 Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
-studies to overlap. Batch effects similarly hinder the utility of t-SNE in
+studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
 the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
@@
 nearest-neighbors~\cite{mnn,seurat}.
-In these works, batch effects are deemed removed when good mixing of cells from different sources is achieved in a t-SNE visualization.
-Elimination of batch effects may require aggressive data
+In these works, batch effects are deemed removed when cells from different sources exhibit good mixing in a t-SNE visualization.
+The elimination of batch effects may require aggressive data
 preprocessing which may blur the boundaries between cell types. Another problem
@@
 We propose a direct solution of rendering t-SNE visualizations that addresses
-batch effects. Our approach considers one of the data sets as a {\em reference}
+batch effects. Our approach treats one of the data sets as a {\em reference}
 and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
-We construct the t\nobreakdash -SNE embedding using the reference data set, and then use it
+We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
 as a scaffold for the embedding of data points from the secondary data.
-A key idea underpinning our approach is that the embedding is performed one data point at a time.
+The key idea underpinning our approach is that the embedding is performed one data point at a time.
 Independence of each new
-embedded data instance from the secondary data set causes clustering the landscape
+embedded data instance from the secondary data set causes the clustering landscape
 to depend only on the reference scaffold, thus removing data source-driven
@@
 field'', independently driving the embedding of each new instance. For example, in
-Fig.~\ref{fig:transform_brain}, the cells from visual cortex define the
-scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
+Fig.~\ref{fig:transform_brain}, the cells from the visual cortex define the
+scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from the
 hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
 visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
-dispersed across entire embedding space and their cell type correctly matches
+dispersed across the entire embedding space and their cell type correctly matches
 the prevailing type in reference clusters.
@@
 The proposed solution is implemented using a mapping of
-new data into an existing t-SNE visualization. While the
+new data to an existing t-SNE visualization. While the
 utility of such an algorithm was already hinted at in recent
@@
 embedding new (secondary) data, enabling joint visualisation of multiple data sources
-while mitigating batch effects. Our proposed approach starts with t\nobreakdash
+while mitigating batch effects. Our proposed approach starts by using t\nobreakdash
 -SNE to embed a reference data set, with the aim of constructing a
@@
 a secondary data set disregards any interactions present in that data set, and
-prevents forming clusters that would be specific to the secondary data. Below, we start
+prevents the formation of clusters that would be specific to the secondary data. Below, we start
 with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, introducing
@@
 
-t-SNE is local, non-linear dimensionality reduction method, tailored to the visualisation of
+t-SNE is a local, non-linear dimensionality reduction method, tailored to the visualisation of
 high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
@@
 t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
-\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \in \mathbb{R}^d$
+\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \\\in \mathbb{R}^d$
 where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
@@
 estimating similarities between each new point and the reference data and optimizing
-the position of each data point in the embedding space. Unlike parametric models
+the position of each new data point in the embedding space. Unlike parametric models
 such as principal component analysis or autoencoders, t-SNE does not define an
@@
 descent. We use an adaptive learning rate scheme with momentum 
-to speed up the convergence, as proposed by Jacobs~\cite{momentum}. We run gradient descent
+to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
 with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
-optimization converged for all cases in our experiments.  The time complexity
+optimization converged in all our experiments. The time complexity
 needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
@@
 resulting in stable convergence. When embedding new samples, the span of the
-embedding is much larger, resulting in much larger gradients, and the default
+embedding is much larger, resulting in substantially larger gradients, and the default
 learning rate causes points to move very far from the reference embedding. In
@@
 by the span of the embedding. Clearly, if even one point ``shoots off'' far from the
-embedding, the number of required grid points may grow, substantially
-increasing the runtime. The reduced learning rate suppresses this issue, and
+embedding, the number of required grid points may grow dramatically,
+increasing the runtime substantially. The reduced learning rate suppresses this issue, and
 does not slow the convergence because of the adaptive learning rate scheme, provided
@@
 estimate similarities between reference data points~\cite{Domingos2012-CACM}.
-When adding new data points from secondary data set to the reference embedding,
+When adding new data points from the secondary data set to the reference embedding,
 we select 1,000 genes present in both data sets and use these to estimate the
@@
 Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
-\ref{fig:transform_retina} show the embedding of the reference data sets and
+\ref{fig:transform_retina} show the embeddings of the reference data sets and
 their corresponding embeddings of the secondary data sets. In all the figures,
@@
 could lead to overly-optimistic results and data visualizations strikingly different from ours. For example,
-compare the visualisations from Fig~\ref{fig:batch_effect}.a and
+compare the visualisations from Fig.~\ref{fig:batch_effect}.a and
 Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
@@
 cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
-Fig~\ref{fig:transform_brain} the vast majority of the cells from the secondary
+Fig.~\ref{fig:transform_brain} the vast majority of the cells from the secondary
 data set were correctly mapped to the same-typed cluster of reference
@@
 places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
-similarities provides a more meaningful layout of the clusters. Neuronal types
+similarities and using proper initialization provides a more meaningful layout of the clusters. Neuronal types
 occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are
@@
 describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
-proposed to position new points into a known embedding by placing them in the
+proposed positioning new points into a known embedding by placing them in the
 median position of their 10 nearest neighbors, where the neighborhood was
@@
 types in the secondary data set, possibly positioning them arbitrarily close to
-unrelated clusters. Procedures such as Scmap were recently
+unrelated clusters. Procedures such as scmap were recently
 proposed to cope with such cases and identify the cells whose type is new and

### 2019-06-14 paper/tsne_embedding.tex
Minor correction and re-order references.

@@
 %
-\title{Embedding to Reference t-SNE Space Adresses Batch-Effects in Single-Cell Classification}
+\title{Embedding to Reference t-SNE Space Adresses Batch Effects in Single-Cell Classification}
 %
@@
 transformed such that the batch effects are explicitly
-removed. Recently proposed procedures for batch-effect removal include
+removed. Recently proposed procedures for batch effect removal include
 canonical correlation analysis~\cite{cca} and mutual
@@
 
-\bibliographystyle{splncs04}
+% \bibliographystyle{splncs04}
+\bibliographystyle{unsrt}
 \bibliography{references}

### 2019-08-12 paper/review-response.tex
Revision edits.

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{todonotes}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{alltt}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{graphicx}

\usepackage{xr}[]
\externaldocument{policar2019-tsne-embedding}
\usepackage{url}

\renewcommand{\thetable}{\Alph{table}}
\renewcommand{\thefigure}{\Alph{figure}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqs}{\begin{equation}}
\newcommand{\eeqs}{\end{equation}}
\newcommand{\mx}[1]{\ensuremath{\boldsymbol{\mathrm{{#1}}}}}   % Simple matrix format.

% \newcommand{\review}[1]{\noindent{{\textbf{\textit{#1}}}}\vspace{3mm}}
% \definecolor{colframe}{rgb}{0.01,0.199,0.1}

\newcounter{rtaskno}
\newcommand{\rtask}[1]{\refstepcounter{rtaskno}\label{#1}}

\newcommand{\reviewc}[2]{\begin{leftbar}\noindent\rtask{#1}\ref{#1}. #2\end{leftbar}}
\newcommand{\review}[1]{\begin{leftbar}\noindent #1\end{leftbar}}
\newcommand{\answer}[1]{\noindent #1\\[0mm]}
\newcommand{\copyquote}[1]{\vspace{2mm} {\color{blue} #1}\vspace{2mm}}

\newlength{\leftbarwidth}
\setlength{\leftbarwidth}{3pt}
\newlength{\leftbarsep}
\setlength{\leftbarsep}{10pt}
\renewenvironment{leftbar}{%
    \def\FrameCommand{{\color{black}{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
    \MakeFramed {\advance \hsize -\width \FrameRestore }%
}{\endMakeFramed}

% \usepackage[colorlinks=True,citecolor=black]{hyperref}

\begin{document}

\noindent{\bf Poli\v{c}ar, Stra\v{z}ar \& Zupan: Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}

\vspace{5mm}
\noindent {\bf RESPONSE TO THE REVIEWERS}

\vspace{5mm}

\noindent Dear Editor and the Reviewers, \\

\vspace{3mm}

\noindent we thank for the comprehensive feedback that helped to improve the
manuscript, sufficient time to implement the changes, and the permission to
submit a revised version.

\vspace{3mm}

All of the reviewer remarks are addressed below with reference to the
corresponding changes in the manuscript.  In case of overlapping remarks from
different reviewers, we refer the reader to the relevant enumerated answer. A
differential document is provided to facilitate the comparison with the initial
submission. The changes to the data, code and the instructions to reproduce all
the experiments are available in the repository:
\url{https://github.com/biolab/tsne-embedding}.

\vspace{3mm}

\begin{flushright}
\noindent Kind regards, \\ 
\vspace{2mm}
Pavlin Poli\v{c}ar \\
Martin Stra\v{z}ar \\
Bla\v{z} Zupan \\ 
\vspace{2mm}
University of Ljubljana, Slovenia 
\end{flushright}

\vspace{10mm}

\subsection*{Reviewer \#1}

\reviewc{r1q1}{I am missing more comparisons to the standard t-SNE, i.e., more
examples where the proposed method removes the batch effect. The results
section is focused on demonstrating that the approach appropriately embeds the
second dataset to the reference dataset (in terms of cell types), however, it
gives few examples of the batch effect and how the proposed method resolves it.
More comparison to the standard t-SNE would be welcome.}

\answer{The batch-effect related to embedding multiple datasets is highly abundant. To reinforce this claim, we have added two additional experiments with independent datasets (...) in the accompanying notebooks. Also in these cases, the standard t-SNE would be confounded by data source of origin.

We have added a paragraph to Section~\ref{sec:results}, pointing the reader to the relevant material and findings.}

\reviewc{r1q2}{Also, the community still largely uses PCA in order to visualize high dimensional data, therefore, the comparison to PCA would be interesting to see as well - this should be done very easily since the proposed method starts from the PCA-based representation of the data.}

\answer{

Since this comparison has been made several times in the literature (...), we
opted to omit this comparison in favour of brevity.  In molecular biology for
example, PCA remains common for analysis of small-sample datasets with tens to
hundreds of samples (e.g. bulk RNA sequencing), while t-SNE is essential for
single-cell studies with tens of thousands to millions of samples.

The two methods are based on different practical objectives.  Note that the
objective of PCA is to find an optimal approximation to the data, \emph{limited
to} an orthogonal subspace of the input data. The linearity of the
transformation additionally limits the capacity of 2D embeddings, typically
used for visualization. While often useful to visualize smaller-to-medium sized
datasets, the low-rank subspace constraint does not work well with increasing
number of dimensions.}

\reviewc{r1q3}{As far as I understand, the results are not symmetric in terms
of which of the two datasets is a reference and which one is embedded? How
should one decide which dataset is better used as a reference? Furthermore, how
should more than two datasets be analyzed with the proposed approach?}

\answer{We agree with the Reviewer's observation and have explicitly mentioned this property in the discussion (Section~\ref{sec:results}). From a theoretical standpoint, the choice of referenence is largely arbitrary. In practice, however, one would start with previously published reference lanscapes and embed own, new (secondary) datasets in order to identify cell types.}

\reviewc{r1q4}{It would be interesting to report what are time and memory requirements of the proposed method. More specifically, what is the overhead of the proposed method in comparison to the standard t-SNE?}

\answer{report the time

The embedding of each cell independently essentially reduced the complexity of embedding the secondary dataset to linear.}

\subsection*{Reviewer \#2}

\reviewc{r2q1}{Can you validate the results also *quantitatively*? Visualizations are nice, but numbers may be able to provide complementary information that the eye does not see.}

\answer{The quantitative estimates of erroneus placed were estimated for each of the three reference/secondary data set pairs. Additionally, we have added two new experiments with cells from new studies, for which the error estimates are reported as well.

Empirically, the method incorrect placement ranges from 1-5 \% of the cells, assuming ground truth annotation.}

\reviewc{r2q2}{Can you *prove* anything about the results except that it is the result of an optimization?}

\answer{The results in research of t-SNE and related non-linear methods from visualization are by and large statistical, empirical and subjective. The key property of our method formulation, which can be proved trivially, is the independence of embedding each secondary data point, which in turn leads to batch-effect removal.

(see what happens with randomly-permuted cell profiles)}

\reviewc{r2q3}{In the abstract, I found the formulation "one data point at a time" misleading. I was expecting order-dependent effects, where you in fact meant that the optimizations are computed for each data point of the secondary dataset individually (and thus are in fact "embarassingly parallel").}

\answer{We thank the reviever for this careful observation. The the text was revised at relevant places (Section~\ref{sec:intro}) to explicitly state that data points are embedded independently of one another and of order.}

\reviewc{r2q4}{Fig. 7, in particular as a comparison with a simpler method, does not look great from my point of view. I fail to see why the newly proposed method is better than the simpler method. Maybe you can elaborate a bit more on that.}

\answer{As stated in Section~\ref{sec:results}, the placement od data points with t-SNE embedding is subject to optimization rather than interpolation (as with nearest neighbours approach). 
The latter is guarateed to place the secondary data points into existing clusters and perhaps prone to overclustering. With t-SNE embedding, this effect is less pronounced and data points unrelated to the reference can be identified as outliers. We have added a sentence to elucidate this difference between the methods.

(Perhaps this is shown with the permutation test.)}

\subsection*{Reviewer \#3}

\reviewc{r3q1}{no quantitative statistics are reported regarding the accuracy of the proposed algorithm in correctly classifying each cell in the correct type. Please consider adding this information}

\answer{The concern was addressed in Reviewer Remark~\ref{r2q1}.}

\reviewc{r3q2}{as the authors pointed out, one limitation of the proposed approach is that the reference dataset should already contain all possible cell types that are present in the other datasets. Would it be possible to label as a outlier any novel cell that does not fit within the cell types present in the reference dataset? Please elaborate}

\answer{The concern was addressed in Reviewer Remark~\ref{r1q3}.}

\end{document}

### 2019-08-12 paper/policar2019-tsne-embedding.tex
Revision edits.

--- 2019-06-14 paper/tsne_embedding.tex
@@
 \usepackage{booktabs}
+\usepackage{todo}
 
@@
 %
-\title{Embedding to Reference t-SNE Space Adresses Batch Effects in Single-Cell Classification}
+\title{Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
 %
@@
 \section{Introduction}
+\label{sec:intro}
 
@@
 as a scaffold for the embedding of data points from the secondary data.
-The key idea underpinning our approach is that the embedding is performed one data point at a time.
+The key idea underpinning our approach is that data points are embedded
+independently of one another.
+
 Independence of each new
@@
 \subsection{Results and Discussion}
+\label{sec:results}
 
@@
 secondary data sets).
+
+Additional experiments with embedding secondary datasets from independent
+studies are provided in the accompanying Python notebooks. We embedded brain
+cells from \cite{} to the reference embedding shown in
+Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
+in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
+of X\% and Y\%. \todo{Add errors}
 
@@
 between different clusters, while after optimization they typically
-move closer to same-typed groups.
+move closer to same-typed groups. While the nearest neighbour-based approach
+always places new data points to their most similar cluster, our proposed
+t-SNE approach can identify extreme outliers and place them far from the
+reference cluster.
 
@@
 proposed to cope with such cases and identify the cells whose type is new and
-not included in the reference~\cite{scmap}. Our procedure does not address such cases, and
-for scaling-up to a wider collection of cell types relies on emerging
-availability of large collections of the reference data such as those managed
-by Human Cell Atlas initiative~\cite{hca}.
+not included in the reference~\cite{scmap}.
+
+Our procedure and the resulting cell landscape are affectected by the asymmetry
+of determining the reference and the secondary datasets. In practice, however,
+newly produced \em secondary \em datasets would be embedded into published and
+annnotated the reference landscapes. Scaling-up to a wider collection of cell
+types is driven by increasingly large collections of data, e.g the Human Cell
+Atlas initiative~\cite{hca}.
 

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Minor edits.

@@
 \caption{A two-dimensional embedding of a reference containing brain cells (a) and
-the corresponding mapping of secondary data containing hypothalamic cells (b). Notice
-that the majority of hypothalamic cells were mapped to their corresponding
+the corresponding mapping of secondary data containing hypothalamic cells (b). 
+The majority of hypothalamic cells were mapped to their corresponding
 reference cluster. For instance, astrocyte cells marked with red on the
@@
 t-SNE visualizations where similarly-typed cells would cluster together,
-despite systematic differences between data sources. Below, we list the data sets,
+despite systematic differences between data sources. To that end, we focus
+on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
+have been substantially covered before~\todo{Add citation}. 
+Below, we list the data sets,
 describe single-cell specific data preprocessing procedures, and display the
@@
 estimated in the original data space. Notice that we use this approach as well,
-but only for the initialization of positions of new data instances that are subject
-to further optimization. In Fig.~\ref{fig:optimization} we demonstrate that
-nearest neighbor-based positioning is insufficient and may yield clumped
-visualizations where the optimal positioning using the t-SNE loss function is
-much more dispersed and rightfully shows a more considerable variation in the
-secondary data. Some data points may also fall into the empty regions
+but only for the initialization of positions of new data instances that are
+subject to further optimization. Despite both nearest-neighbours search and
+t-SNE optimization can be computed in linear time, the former dominates the
+runtime (mouse retina example; M reference, N secondary cells, X min N-search,
+Y min optimization).
+
+We demonstrate a case where nearest neighbor-based positioning is insufficient
+in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
+optimal positioning using the t-SNE loss function is much more dispersed and
+rightfully shows a more considerable variation in the secondary data. 
+Some data points may also fall into the empty regions
 between different clusters, while after optimization they typically
@@
 of determining the reference and the secondary datasets. In practice, however,
-newly produced \em secondary \em datasets would be embedded into published and
-annnotated the reference landscapes. Scaling-up to a wider collection of cell
+newly produced \em secondary \em data would be embedded into published and
+annotated reference landscapes. Scaling-up to a wider collection of cell
 types is driven by increasingly large collections of data, e.g the Human Cell

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Shortening and diff.

@@
 data set and uses it as a scaffold for embedding new data. The new, secondary data is
-embedded one data-point at the time. This prevents any interactions
+embedded one data point at the time. This prevents any interactions
 between instances in the secondary data and implicitly mitigates batch
-effects. We demonstrate the utility of this approach with an analysis of six
-recently published single-cell gene expression data sets containing up to tens
-of thousands of cells and thousands of genes. In these data sets, the batch
-effects are particularly strong as the data comes from different institutions
+effects. We demonstrate the utility of this approach by analyzing six
+recently published single-cell gene expression data sets with up to tens
+of thousands of cells and thousands of genes. The batch
+effects are particularly strong as the used data comes from different institutions
 and was obtained using different experimental protocols. The visualizations
 constructed by our proposed approach are cleared of batch effects, and the cells
-from secondary data sets correctly co-cluster with cells from the primary data
-sharing the same cell type.
+from secondary data sets correctly co-cluster with cells of the same type from the primary data.
 
@@
 scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
-from a single coherent source, these methods may encounter problems if the data
-comes from multiple sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
-reveal source-specific clusters instead of grouping data instances of the same class-type, regardless of data sources. This source-specific confounding is often referred to as {\em domain
+from a single coherent source, these methods may encounter problems 
+with multiple data sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
+reveal source-specific clusters instead of grouping data instances of the same class, regardless of data sources. This source-specific confounding is often referred to as {\em domain
 shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
-{\em dataset shift}~\cite{dataset_shift}. In bioinformatics, the
+{\em data set shift}~\cite{data set_shift}. In bioinformatics, the
 domain-specific differences are more commonly referred to as {\em batch
@@
 
-Massive, multi-variate biological data sets often suffer from these source-specific
-biases. Consider an example from single-cell genomics, the domain we will focus
-on in this manuscript and that was --- besides current scientific challenges
---- selected also due to the availability and abundance of recently published
-data. Single-cell RNA sequencing (scRNA-seq) data sets are the result of isolating RNA molecules from individual cells,
+Massive, multi-variate biological data sets often suffer from these
+source-specific biases.  The focus of this work is single-cell genomics, a
+domain that was selected due high biomedical relevance and and abundance of
+recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
+the result of isolating RNA molecules from individual cells,
 which serve as an estimate of the expression of cell's genes.
-Single-cell studies, which can exceed thousands of cells and tens of thousands of genes,
-typically start with the analysis of cell types. Here, it is generally expected that
+The studies can exceed thousands of cells and tens of thousands of genes,
+and typically start with cell type analysis. Here, it is expected that
 cells of the same type would cluster together in two-dimensional data
@@
 cells, pancreatic cells cluster primarily by data source, again
-resulting in an uninformative visualization driven by batch
-effect.
+resulting in a visualization driven by batch effects.
 
@@
 preprocessing which may blur the boundaries between cell types. Another problem
-is also the inclusion of any new data, for which the entire data analysis
+is also the inclusion of any new data, for which the entire analysis
 pipeline must be rerun, usually resulting in a different layout and clusters
@@
 We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
-as a scaffold for the embedding of data points from the secondary data.
-The key idea underpinning our approach is that data points are embedded
+as a scaffold for the embedding of the secondary data.
+The key idea underpinning our approach is that secondary data points are embedded
 independently of one another.
 
-Independence of each new
-embedded data instance from the secondary data set causes the clustering landscape
+Independent embedding of each secondary datum causes the clustering landscape
 to depend only on the reference scaffold, thus removing data source-driven
@@
 
-t-SNE is a local, non-linear dimensionality reduction method, tailored to the visualisation of
-high-dimensional data sets. Given a multi-dimensional data set $\mathbf{X} =
+Local, non-linear dimensionality reduction by t-SNE is performed as follows.
+Given a multi-dimensional data set $\mathbf{X} =
 \left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
-\mathbb{R}^D$ where $N$ is the number of samples in the reference data set,
+\mathbb{R}^D$ where $N$ is the number of data points in the reference data set,
 t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
@@
 Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of
-the algorithm impractical for any reasonably-sized data. To address larger data
-sets, we adopt a recent approach for low-rank approximation of
+the algorithm impractical for large data sets. 
+Hence, we adopt a recent approach for low-rank approximation of
 gradients based on polynomial interpolation which reduces the time
@@
 ignore the long-range interactions between clusters. Larger values may result
-in a more globally consistent visualisations, preserving distances on a large
-scale and organizing clusters in a more meaningful way. Larger values of
-perplexity can lead to merging of multiple small clusters, thus obscuring
-local aspects of the data~\cite{art_of_using_tsne}.
+in a more globally consistent visualisations --- preserving distances on a large
+scale and organizing clusters in a more meaningful way --- but can lead to merging 
+small clusters and thus obscuring local aspects of the data~\cite{art_of_using_tsne}.
 
@@
 to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
-with momentum $\alpha$ set to $0.8$ for $250$ iterations, where the
+with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
 optimization converged in all our experiments. The time complexity
@@
 single-cell data. Data in this realm include a variety of
-cells from specific tissues, and characterize them through gene expression.
+cells from specific tissues and are characterized through gene expression.
 In our experiments, we considered several recently published data
@@
 the majority of cell types from the secondary data set were included in the
-reference set (Table~\ref{tab:datasets}).
+reference set (Table~\ref{tab:data sets}).
+The cells in the data sets originates from the following three tissues:
 
+\begin{description}
+\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
+cells from the visual cortex exploring transcriptional changes after exposure
+to light. This was used as a reference for the data from Chen
+\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
+their reaction to food deprivation. From the secondary data, we removed cells
+with no corresponding types in the reference: tanycytes, ependymal,
+epithelial, and unlabelled cells.
+
+\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was
+created as an atlas of pancreatic cell types. We used this set as a reference
+for data from Xin \etal~\cite{xin2016}, who examined transcriptional
+differences between healthy and type 2 diabetic patients.
+
+\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
+created as an atlas of mouse retinal cell types. We used this as a reference
+for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for retinal bipolar cells.
+\end{description}
+
+\vspace{-3mm}
 \begin{table}[ht]
@@
 \end{center}
-\caption{Data sets used in our experiments. In each pair, the first data set
-(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. In
-all cases, we relied on the quality control and annotations from the original
-publication. To facilitate comparisons, the cell annotations were harmonized using
-cell type annotations from the cell ontology~\cite{cell_ontology}. Notice that
-different RNA sequencing protocols were used to estimate gene expressions. We
-report the number of cell types from each data set retained after
-preprocessing. Single-cell data is sparse, typically containing less than 10\%
-expressed genes per cell.}
-\label{tab:datasets}
+\caption{Data sets used in our experiments. The first data set in a pair
+(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. 
+We relied on the quality control and annotations from the original
+publication and report the humber of cell types after preprocessinng. 
+The cell annotations were made consistent to
+annotations from the Cell Ontology~\cite{cell_ontology}. Notice that
+different RNA sequencing protocols were used to estimate gene expressions.}
+\label{tab:data sets}
 \end{table}
-
-The cells in the data sets originates from the following three tissues:
-\begin{description}
-\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
-cells from the visual cortex exploring transcriptional changes after exposure
-to light. This was used as a reference for the data from Chen
-\etal~\cite{chen2017}, containing various cells from the mouse hypothalamus and
-their reaction to food deprivation. From the secondary data, we removed cells
-with no corresponding types in the reference, namely ependymal cells,
-epithelial cells, tanycytes, and unlabelled cells.
-
-\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was
-created as an atlas of pancreatic cell types. We used this set as a reference
-for data from Xin \etal~\cite{xin2016}, who examined transcriptional
-differences between healthy and type 2 diabetic patients.
-
-\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
-created as an atlas of mouse retinal cell types. We used this as a reference
-for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
-different types of retinal bipolar cells.
-\end{description}
 
@@
 When adding new data points from the secondary data set to the reference embedding,
-we select 1,000 genes present in both data sets and use these to estimate the
-similarities between the secondary data item and reference data points. The
-similarities are estimated using cosine similarity. We note that similarities are
-computed using the raw count matrices. The preprocessing stages are detailed in
-accompanying Python notebooks (Sec.~\ref{sec:implementation}).
+we select 1,000 genes present in both data sets and use the cosine similarity 
+to estimate the similarities between the secondary data item and reference data
+points.  We note that similarities are computed using the raw count matrices.
+The preprocessing stages are detailed in accompanying Python notebooks
+(Sec.~\ref{sec:implementation}).
 
@@
 
-Additional experiments with embedding secondary datasets from independent
+Additional experiments with embedding secondary data sets from independent
 studies are provided in the accompanying Python notebooks. We embedded brain
@@
 runtime (mouse retina example; M reference, N secondary cells, X min N-search,
-Y min optimization).
+Y min optimization). \todo{Add nunmbers}
 
@@
 rightfully shows a more considerable variation in the secondary data. 
-Some data points may also fall into the empty regions
-between different clusters, while after optimization they typically
+Some points may also fall into the empty regions, while after optimization they typically
 move closer to same-typed groups. While the nearest neighbour-based approach
@@
 
-Our procedure and the resulting cell landscape are affectected by the asymmetry
-of determining the reference and the secondary datasets. In practice, however,
+Our procedure and the resulting cell landscape are therefore affectected by the asymmetry
+of determining the reference and the secondary data sets. In practice, however,
 newly produced \em secondary \em data would be embedded into published and
@@
 Almost all recent publications of single-cell studies begin with a
-two-dimensional visualization of the data that reveals the cellular diversity containing many 
-different cell-types from the study. While any dimensionality reduction
-technique can be used to render such a visualization, different variants of
+two-dimensional visualization of the data that reveals the cellular diversity. 
+While most dimensionality reduction
+techniques can be used to render such a visualization, different variants of
 t-SNE are most often used. Due to the ability to explore biological mechanisms

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Fixes in formulae

@@
 shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
-{\em data set shift}~\cite{data set_shift}. In bioinformatics, the
+{\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
 domain-specific differences are more commonly referred to as {\em batch
@@
 p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_j ) / \sigma_i^2 \right )}
-{\sum_{k \neq \i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
+{\sum_{k \neq i } \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{x}_k ) / \sigma_i^2 \right )}, \quad p_{i \mid i} = 0
 \label{eq:gaussian_kernel}
@@
 \begin{align}
-p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} d(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
+p_{j \mid i} &= \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \\
 q_{j \mid i} &= \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}},
@@
 \cdot M)$, however, by adapting the same polynomial interpolation based
-approximation, this is reduced to $\mathcal{O}(N)$.
+approximation, this is reduced to $\mathcal{O}(\max \{ N, M \})$.
 
@@
 studies are provided in the accompanying Python notebooks. We embedded brain
-cells from \cite{} to the reference embedding shown in
+cells from~\cite{} to the reference embedding shown in
 Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
@@
 but only for the initialization of positions of new data instances that are
-subject to further optimization. Despite both nearest-neighbours search and
+subject to further optimization. Despite both nearest-neighbors search and
 t-SNE optimization can be computed in linear time, the former dominates the
@@
 Some points may also fall into the empty regions, while after optimization they typically
-move closer to same-typed groups. While the nearest neighbour-based approach
+move closer to same-typed groups. While the nearest neighbor-based approach
 always places new data points to their most similar cluster, our proposed

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Add missing references

@@
 on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
-have been substantially covered before~\todo{Add citation}. 
+have been substantially covered before~\cite{tsne,umap_single_cell}. 
 Below, we list the data sets,
@@
 studies are provided in the accompanying Python notebooks. We embedded brain
-cells from~\cite{} to the reference embedding shown in
-Fig.~\ref{fig:transform_brain}.a, and pancreas cells from~\cite{} to reference
+cells from Zeisel \etal~\cite{zeisel2015} to the reference embedding shown in
+Fig.~\ref{fig:transform_brain}.a, and pancreas cells from Muraro \etal~\cite{muraro2016} to reference
 in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Reformat latex

@@
 Dimensionality reduction techniques, such as t-SNE, can construct informative
-visualizations of high-dimensional data. When working with multiple data sets,
-a straightforward application of these methods often fails; instead of
-revealing underlying classes, the resulting visualizations expose
-data set-specific clusters. To circumvent these batch effects, we propose an
-embedding procedure that takes a t-SNE visualization constructed on a reference
-data set and uses it as a scaffold for embedding new data. The new, secondary data is
-embedded one data point at the time. This prevents any interactions
-between instances in the secondary data and implicitly mitigates batch
-effects. We demonstrate the utility of this approach by analyzing six
-recently published single-cell gene expression data sets with up to tens
-of thousands of cells and thousands of genes. The batch
-effects are particularly strong as the used data comes from different institutions
-and was obtained using different experimental protocols. The visualizations
-constructed by our proposed approach are cleared of batch effects, and the cells
-from secondary data sets correctly co-cluster with cells of the same type from the primary data.
+visualizations of high-dimensional data. When working with multiple data
+sets, a straightforward application of these methods often fails; instead of
+revealing underlying classes, the resulting visualizations expose data
+set-specific clusters. To circumvent these batch effects, we propose an
+embedding procedure that takes a t-SNE visualization constructed on a
+reference data set and uses it as a scaffold for embedding new data. The new,
+secondary data is embedded one data point at the time. This prevents any
+interactions between instances in the secondary data and implicitly mitigates
+batch effects. We demonstrate the utility of this approach by analyzing six
+recently published single-cell gene expression data sets with up to tens of
+thousands of cells and thousands of genes. The batch effects are particularly
+strong as the used data comes from different institutions and was obtained
+using different experimental protocols. The visualizations constructed by our
+proposed approach are cleared of batch effects, and the cells from secondary
+data sets correctly co-cluster with cells of the same type from the primary
+data.
 
@@
 profiles are similar. For this task, non-linear embedding techniques such as
-t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or uniform
-manifold approximation and projection~\cite{umap} have recently complemented
-traditional data transformation and embedding approaches such as principal
-component analysis (PCA) and multi-dimensional
-scaling~\cite{distill,umap_single_cell}. While useful for visualizing data
-from a single coherent source, these methods may encounter problems 
-with multiple data sources. Here, when performing dimensionality reduction on a merged data set, the resulting visualizations would typically
-reveal source-specific clusters instead of grouping data instances of the same class, regardless of data sources. This source-specific confounding is often referred to as {\em domain
-shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift} or
-{\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
+t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or
+uniform manifold approximation and projection~\cite{umap} have recently
+complemented traditional data transformation and embedding approaches such as
+principal component analysis (PCA) and multi-dimensional
+scaling~\cite{distill,umap_single_cell}. While useful for visualizing data from
+a single coherent source, these methods may encounter problems with multiple
+data sources. Here, when performing dimensionality reduction on a merged data
+set, the resulting visualizations would typically reveal source-specific
+clusters instead of grouping data instances of the same class, regardless of
+data sources. This source-specific confounding is often referred to as {\em
+domain shift}~\cite{domain_shift}, {\em covariate shift}~\cite{covariate_shift}
+or {\em data set shift}~\cite{dataset_shift}. In bioinformatics, the
 domain-specific differences are more commonly referred to as {\em batch
@@
 recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
-the result of isolating RNA molecules from individual cells,
-which serve as an estimate of the expression of cell's genes.
-The studies can exceed thousands of cells and tens of thousands of genes,
-and typically start with cell type analysis. Here, it is expected that
-cells of the same type would cluster together in two-dimensional data
-visualisation~\cite{seurat}. For instance, Fig.~\ref{fig:batch_effect}.a shows
-t-SNE embedded data from mouse brain cells originating from the visual
-cortex~\cite{hrvatin2018} and the hypothalamus~\cite{chen2017}. The figure reveals
-distinct clusters but also separates the data from the two brain regions. These
-two regions share the same cell types and --- contrary to the depiction in
-Fig.~\ref{fig:batch_effect}.a --- we would expect the data points from the two
-studies to overlap. Batch effects similarly prohibit the utility of t-SNE in
-the exploration of pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which
-renders the data from a pancreatic cell atlas~\cite{baron2016} and similarly-typed
-cells from diabetic patients~\cite{xin2016}. Just like with data from brain
-cells, pancreatic cells cluster primarily by data source, again
-resulting in a visualization driven by batch effects.
+the result of isolating RNA molecules from individual cells, which serve as an
+estimate of the expression of cell's genes.  The studies can exceed thousands
+of cells and tens of thousands of genes, and typically start with cell type
+analysis. Here, it is expected that cells of the same type would cluster
+together in two-dimensional data visualisation~\cite{seurat}. For instance,
+Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
+originating from the visual cortex~\cite{hrvatin2018} and the
+hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
+separates the data from the two brain regions. These two regions share the same
+cell types and --- contrary to the depiction in Fig.~\ref{fig:batch_effect}.a
+--- we would expect the data points from the two studies to overlap. Batch
+effects similarly prohibit the utility of t-SNE in the exploration of
+pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
+a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
+diabetic patients~\cite{xin2016}. Just like with data from brain cells,
+pancreatic cells cluster primarily by data source, again resulting in a
+visualization driven by batch effects.
 
@@
   \includegraphics[width=\textwidth]{figures/batch_effect.pdf}
-\caption{Batch effects are a driving factor of variation between the data sets.
-Depicted is a t-SNE visualisation of two pairs of data sets. In each pair, the
-data sets share cell types, so it would be expected that the cells from the
-reference data (blue) would mix with the cells in a secondary data sets
-(orange). Instead, t-SNE visualisation clusters data according to the data
-source.} \label{fig:batch_effect}
+  \caption{Batch effects are a driving factor of variation between the data
+  sets.  Depicted is a t-SNE visualisation of two pairs of data sets. In each
+  pair, the data sets share cell types, so it would be expected that the cells
+  from the reference data (blue) would mix with the cells in a secondary data
+  sets (orange).  Instead, t-SNE visualisation clusters data according to the
+  data source.}
+  \label{fig:batch_effect}
 \end{figure}
@@
 batch effect problems up-front. The data is typically preprocessed and
-transformed such that the batch effects are explicitly
-removed. Recently proposed procedures for batch effect removal include
-canonical correlation analysis~\cite{cca} and mutual
-nearest-neighbors~\cite{mnn,seurat}.
-In these works, batch effects are deemed removed when cells from different sources exhibit good mixing in a t-SNE visualization.
-The elimination of batch effects may require aggressive data
-preprocessing which may blur the boundaries between cell types. Another problem
-is also the inclusion of any new data, for which the entire analysis
-pipeline must be rerun, usually resulting in a different layout and clusters
-that have little resemblance to original visualization and thus require
-reinterpretation.
+transformed such that the batch effects are explicitly removed. Recently
+proposed procedures for batch effect removal include canonical correlation
+analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
+works, batch effects are deemed removed when cells from different sources
+exhibit good mixing in a t-SNE visualization.  The elimination of batch effects
+may require aggressive data preprocessing which may blur the boundaries between
+cell types. Another problem is also the inclusion of any new data, for which
+the entire analysis pipeline must be rerun, usually resulting in a different
+layout and clusters that have little resemblance to original visualization and
+thus require reinterpretation.
 
@@
 batch effects. Our approach treats one of the data sets as a {\em reference}
-and aims to embed the cells from another, {\em secondary data set} to a common low-dimensional space.
-We construct a t\nobreakdash -SNE embedding using the reference data set, and then use it
-as a scaffold for the embedding of the secondary data.
-The key idea underpinning our approach is that secondary data points are embedded
-independently of one another.
+and aims to embed the cells from another, {\em secondary data set} to a common
+low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
+reference data set, and then use it as a scaffold for the embedding of the
+secondary data.  The key idea underpinning our approach is that secondary data
+points are embedded independently of one another.
 
@@
 variation. In other words, when including new data, the scaffold inferred from
-the reference data set is kept unchanged and defines a ``gravitational
-field'', independently driving the embedding of each new instance. For example, in
+the reference data set is kept unchanged and defines a ``gravitational field'',
+independently driving the embedding of each new instance. For example, in
 Fig.~\ref{fig:transform_brain}, the cells from the visual cortex define the
-scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from the
-hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint t\nobreakdash -SNE
-visualization (Fig.~\ref{fig:batch_effect}.a), the hypothalamic cells are
-dispersed across the entire embedding space and their cell type correctly matches
-the prevailing type in reference clusters.
+scaffold (Fig.~\ref{fig:transform_brain}.a) into which we embed the cells from
+the hypothalamus (Fig.~\ref{fig:transform_brain}.b). Unlike in their joint
+t\nobreakdash -SNE visualization (Fig.~\ref{fig:batch_effect}.a), the
+hypothalamic cells are dispersed across the entire embedding space and their
+cell type correctly matches the prevailing type in reference clusters.
 
@@
   \includegraphics[width=\textwidth]{figures/transform_brain.pdf}
-\caption{A two-dimensional embedding of a reference containing brain cells (a) and
-the corresponding mapping of secondary data containing hypothalamic cells (b). 
-The majority of hypothalamic cells were mapped to their corresponding
-reference cluster. For instance, astrocyte cells marked with red on the
-right were mapped to an oval cluster of same-typed cells denoted with the same
-color in the visualization on the left.} \label{fig:transform_brain}
+  \caption{A two-dimensional embedding of a reference containing brain cells
+  (a) and the corresponding mapping of secondary data containing hypothalamic
+  cells (b).  The majority of hypothalamic cells were mapped to their
+  corresponding reference cluster. For instance, astrocyte cells marked with
+  red on the right were mapped to an oval cluster of same-typed cells denoted
+  with the same color in the visualization on the left.}
+  \label{fig:transform_brain}
 \end{figure}
 
-The proposed solution is implemented using a mapping of
-new data to an existing t-SNE visualization. While the
-utility of such an algorithm was already hinted at in recent
-publication~\cite{art_of_using_tsne}, here we provide its practical and
-theoretically-grounded implementation. Considering the abundance of recent
-publications on batch effect removal, we present surprising evidence that a
-computationally more direct and principled embedding procedure solves the batch
-effects problem when constructing interpretable visualizations from different
-data sources.
+The proposed solution is implemented using a mapping of new data to an existing
+t-SNE visualization. While the utility of such an algorithm was already hinted
+at in recent publication~\cite{art_of_using_tsne}, here we provide its
+practical and theoretically-grounded implementation. Considering the abundance
+of recent publications on batch effect removal, we present surprising evidence
+that a computationally more direct and principled embedding procedure solves
+the batch effects problem when constructing interpretable visualizations from
+different data sources.
 
@@
 
-We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a scaffold for
-embedding new (secondary) data, enabling joint visualisation of multiple data sources
-while mitigating batch effects. Our proposed approach starts by using t\nobreakdash
--SNE to embed a reference data set, with the aim of constructing a
-two-dimensional visualisation to facilitate interpretation and cluster
-classification. Then, the placement of each new sample is
-optimized independently via the
-t\nobreakdash -SNE loss function. Independent treatment of each data instance from
-a secondary data set disregards any interactions present in that data set, and
-prevents the formation of clusters that would be specific to the secondary data. Below, we start
-with a summary of t-SNE and its extensions (Sec.~\ref{sec:tsne}, introducing
-the relevant notation, upon which we base our secondary data embedding
-approach (Sec.~\ref{sec:transfer}).
+We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
+scaffold for embedding new (secondary) data, enabling joint visualisation of
+multiple data sources while mitigating batch effects. Our proposed approach
+starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
+of constructing a two-dimensional visualisation to facilitate interpretation
+and cluster classification. Then, the placement of each new sample is optimized
+independently via the t\nobreakdash -SNE loss function. Independent treatment
+of each data instance from a secondary data set disregards any interactions
+present in that data set, and prevents the formation of clusters that would be
+specific to the secondary data. Below, we start with a summary of t-SNE and its
+extensions (Sec.~\ref{sec:tsne}, introducing the relevant notation, upon which
+we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).
 
@@
 Local, non-linear dimensionality reduction by t-SNE is performed as follows.
-Given a multi-dimensional data set $\mathbf{X} =
-\left \{ \mathbf{x}_1, \mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in
-\mathbb{R}^D$ where $N$ is the number of data points in the reference data set,
-t-SNE aims to find a low dimensional embedding $\mathbf{Y} = \left \{
-\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_N \right \} \\\in \mathbb{R}^d$
-where $d \ll D$, such that if points $\mathbf{x}_i$ and $\mathbf{x}_j$ are
-close in the multi-dimensional space, their corresponding embeddings
-$\mathbf{y}_i$ and $\mathbf{y}_j$ are also close. Since t-SNE is primarily used
-as a visualization tool, $d$ is typically set to two. The similarity between
-two data points in t-SNE is defined as:
+Given a multi-dimensional data set $\mathbf{X} = \left \{ \mathbf{x}_1,
+\mathbf{x}_2, \dots, \mathbf{x}_N \right \} \in \mathbb{R}^D$ where $N$ is the
+number of data points in the reference data set, t-SNE aims to find a low
+dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
+\mathbf{y}_N \right \} \\\in \mathbb{R}^d$ where $d \ll D$, such that if points
+$\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
+their corresponding embeddings $\mathbf{y}_i$ and $\mathbf{y}_j$ are also
+close. Since t-SNE is primarily used as a visualization tool, $d$ is typically
+set to two. The similarity between two data points in t-SNE is defined as:
 
@@
 
-The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes
-the Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
+The t-SNE method finds an embedding $\mathbf{Y}$ that minimizes the
+Kullback-Leibler (KL) divergence between $\mathbf{P}$ and $\mathbf{Q}$,
 
@@
 The time complexity needed to evaluate the similarities in
-Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of
-the algorithm impractical for large data sets. 
-Hence, we adopt a recent approach for low-rank approximation of
-gradients based on polynomial interpolation which reduces the time
-complexity of t-SNE to $\mathcal{O}(N)$. This approximation enables the
-visualization of massive data sets, possibly containing millions of data
-points~\cite{fi_tsne}.
+Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of the
+algorithm impractical for large data sets.  Hence, we adopt a recent approach
+for low-rank approximation of gradients based on polynomial interpolation which
+reduces the time complexity of t-SNE to $\mathcal{O}(N)$. This approximation
+enables the visualization of massive data sets, possibly containing millions of
+data points~\cite{fi_tsne}.
 
 The resulting embeddings substantially depend on the value of the perplexity
-parameter. Perplexity can be interpreted as the number of
-neighbors for which the distances in the embedding space are preserved. Small
-values of perplexity result in tightly-packed clusters of points and effectively
-ignore the long-range interactions between clusters. Larger values may result
-in a more globally consistent visualisations --- preserving distances on a large
-scale and organizing clusters in a more meaningful way --- but can lead to merging 
-small clusters and thus obscuring local aspects of the data~\cite{art_of_using_tsne}.
+parameter. Perplexity can be interpreted as the number of neighbors for which
+the distances in the embedding space are preserved. Small values of perplexity
+result in tightly-packed clusters of points and effectively ignore the
+long-range interactions between clusters. Larger values may result in a more
+globally consistent visualisations --- preserving distances on a large scale
+and organizing clusters in a more meaningful way --- but can lead to merging
+small clusters and thus obscuring local aspects of the
+data~\cite{art_of_using_tsne}.
 
@@
 
-Our algorithm, which embeds new data points to a reference embedding, consists of
-estimating similarities between each new point and the reference data and optimizing
-the position of each new data point in the embedding space. Unlike parametric models
-such as principal component analysis or autoencoders, t-SNE does not define an
-explicit mapping to the embedding space, and embeddings need to be found through
-loss function optimization.
+Our algorithm, which embeds new data points to a reference embedding, consists
+of estimating similarities between each new point and the reference data and
+optimizing the position of each new data point in the embedding space. Unlike
+parametric models such as principal component analysis or autoencoders, t-SNE
+does not define an explicit mapping to the embedding space, and embeddings need
+to be found through loss function optimization.
 
-The position of a new data
-point in embedding space is initialized 
-to the median reference embedding position of its $k$ nearest 
-neighbors. While we found the algorithm to be
-robust to choices of $k$, we use $k=10$ in our experiments.
+The position of a new data point in embedding space is initialized to the
+median reference embedding position of its $k$ nearest neighbors. While we
+found the algorithm to be robust to choices of $k$, we use $k=10$ in our
+experiments.
 
@@
 In the optimization step, we refine point positions using batch gradient
-descent. We use an adaptive learning rate scheme with momentum 
-to speed up the convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient descent
-with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
-optimization converged in all our experiments. The time complexity
-needed to evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N
-\cdot M)$, however, by adapting the same polynomial interpolation based
-approximation, this is reduced to $\mathcal{O}(\max \{ N, M \})$.
+descent. We use an adaptive learning rate scheme with momentum to speed up the
+convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient
+descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
+optimization converged in all our experiments. The time complexity needed to
+evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
+however, by adapting the same polynomial interpolation based approximation,
+this is reduced to $\mathcal{O}(\max \{ N, M \})$.
 
@@
 resulting in stable convergence. When embedding new samples, the span of the
-embedding is much larger, resulting in substantially larger gradients, and the default
-learning rate causes points to move very far from the reference embedding. In
-our experiments, we found that decreasing the learning rate to $\eta \sim 0.1$
-produces stable solutions. This is especially important when using the
-interpolation-based approximation, which places a grid of interpolation
-points over the embedding space, where the number of grid points is determined
-by the span of the embedding. Clearly, if even one point ``shoots off'' far from the
-embedding, the number of required grid points may grow dramatically,
-increasing the runtime substantially. The reduced learning rate suppresses this issue, and
-does not slow the convergence because of the adaptive learning rate scheme, provided
-the optimization is run for a sufficient number of steps.
+embedding is much larger, resulting in substantially larger gradients, and the
+default learning rate causes points to move very far from the reference
+embedding. In our experiments, we found that decreasing the learning rate to
+$\eta \sim 0.1$ produces stable solutions. This is especially important when
+using the interpolation-based approximation, which places a grid of
+interpolation points over the embedding space, where the number of grid points
+is determined by the span of the embedding. Clearly, if even one point ``shoots
+off'' far from the embedding, the number of required grid points may grow
+dramatically, increasing the runtime substantially. The reduced learning rate
+suppresses this issue, and does not slow the convergence because of the
+adaptive learning rate scheme, provided the optimization is run for a
+sufficient number of steps.
 
@@
 
-We apply the proposed approach to t-SNE visualizations of
-single-cell data. Data in this realm include a variety of
-cells from specific tissues and are characterized through gene expression.
-In our experiments, we considered several recently published data
-sets where cells were annotated with the cell type. Our aim was to construct
-t-SNE visualizations where similarly-typed cells would cluster together,
-despite systematic differences between data sources. To that end, we focus
-on comparing different ways of using t-SNE rather than differences to embeddings like PCA or MDS, which
-have been substantially covered before~\cite{tsne,umap_single_cell}. 
-Below, we list the data sets,
+We apply the proposed approach to t-SNE visualizations of single-cell data.
+Data in this realm include a variety of cells from specific tissues and are
+characterized through gene expression.  In our experiments, we considered
+several recently published data sets where cells were annotated with the cell
+type. Our aim was to construct t-SNE visualizations where similarly-typed cells
+would cluster together, despite systematic differences between data sources. To
+that end, we focus on comparing different ways of using t-SNE rather than
+differences to embeddings like PCA or MDS, which have been substantially
+covered before~\cite{tsne,umap_single_cell}.  Below, we list the data sets,
 describe single-cell specific data preprocessing procedures, and display the
@@
 the majority of cell types from the secondary data set were included in the
-reference set (Table~\ref{tab:data sets}).
-The cells in the data sets originates from the following three tissues:
+reference set (Table~\ref{tab:data sets}). The cells in the data sets
+originates from the following three tissues:
 
@@
 created as an atlas of mouse retinal cell types. We used this as a reference
-for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for retinal bipolar cells.
+for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
+retinal bipolar cells.
 \end{description}
@@
 \caption{Data sets used in our experiments. The first data set in a pair
-(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference. 
-We relied on the quality control and annotations from the original
-publication and report the humber of cell types after preprocessinng. 
-The cell annotations were made consistent to
-annotations from the Cell Ontology~\cite{cell_ontology}. Notice that
-different RNA sequencing protocols were used to estimate gene expressions.}
+(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
+relied on the quality control and annotations from the original publication
+and report the humber of cell types after preprocessinng.  The cell
+annotations were made consistent to annotations from the Cell
+Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
+were used to estimate gene expressions.}
 \label{tab:data sets}
@@
 estimate similarities between reference data points~\cite{Domingos2012-CACM}.
-When adding new data points from the secondary data set to the reference embedding,
-we select 1,000 genes present in both data sets and use the cosine similarity 
-to estimate the similarities between the secondary data item and reference data
-points.  We note that similarities are computed using the raw count matrices.
-The preprocessing stages are detailed in accompanying Python notebooks
-(Sec.~\ref{sec:implementation}).
+When adding new data points from the secondary data set to the reference
+embedding, we select 1,000 genes present in both data sets and use the cosine
+similarity to estimate the similarities between the secondary data item and
+reference data points. We note that similarities are computed using the raw
+count matrices. The preprocessing stages are detailed in accompanying Python
+notebooks (Sec.~\ref{sec:implementation}).
 
@@
 cells from Zeisel \etal~\cite{zeisel2015} to the reference embedding shown in
-Fig.~\ref{fig:transform_brain}.a, and pancreas cells from Muraro \etal~\cite{muraro2016} to reference
-in \ref{fig:transform_pancreas}.a. The conclusions are largely consistent, with the error rates 
-of X\% and Y\%. \todo{Add errors}
+Fig.~\ref{fig:transform_brain}.a, and pancreas cells from Muraro
+\etal~\cite{muraro2016} to reference in \ref{fig:transform_pancreas}.a. The
+conclusions are largely consistent, with the error rates of X\% and Y\%.
+\todo{Add errors}
 
 Notice that we could simulate the split between reference and secondary data
-sets using one data set only and perform cross-validation, however this type of experiment
-would not incorporate batch effects. We want to remind the reader that handling
-batch effects were central to our endeavour and that the disregard of this effect
-could lead to overly-optimistic results and data visualizations strikingly different from ours. For example,
-compare the visualisations from Fig.~\ref{fig:batch_effect}.a and
+sets using one data set only and perform cross-validation, however this type of
+experiment would not incorporate batch effects. We want to remind the reader
+that handling batch effects were central to our endeavour and that the
+disregard of this effect could lead to overly-optimistic results and data
+visualizations strikingly different from ours. For example, compare the
+visualisations from Fig.~\ref{fig:batch_effect}.a and
 Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
@@
   cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
-Fig.~\ref{fig:transform_brain} the vast majority of the cells from the secondary
-data set were correctly mapped to the same-typed cluster of reference
-cells.}\label{fig:transform_pancreas}
+  Fig.~\ref{fig:transform_brain} the vast majority of the cells from the
+  secondary data set were correctly mapped to the same-typed cluster of
+  reference cells.}
+  \label{fig:transform_pancreas}
 \end{figure}
@@
   focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b). We use
-colors consistent with the study by Macosko \etal.} \label{fig:transform_retina}
+  colors consistent with the study by Macosko \etal.}
+  \label{fig:transform_retina}
 \end{figure}
@@
 There are additional modifications that we use in the embedding of the
-secondary data set that were recently proposed and enhance the original t-SNE visualization.
-One important extension is the use of multi-scale similarities that, besides local ordering of the
-data points, includes global optimization of cluster placement. For
-illustration, consider visualizations with standard and multi-scale t-SNE in
-Fig.~\ref{fig:multiscale}. Notice, for instance, that in multi-scale t-SNE
-(Fig.~\ref{fig:multiscale}.b) the clusters with neuronal cells are clumped
-together, while their placement in standard t-SNE is arbitrary
-(Fig.~\ref{fig:multiscale}.a).
+secondary data set that were recently proposed and enhance the original t-SNE
+visualization.  One important extension is the use of multi-scale similarities
+that, besides local ordering of the data points, includes global optimization
+of cluster placement. For illustration, consider visualizations with standard
+and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that
+in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuronal
+cells are clumped together, while their placement in standard t-SNE is
+arbitrary (Fig.~\ref{fig:multiscale}.a).
 
@@
   \includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
-\caption{A comparison of standard and multi-scale t-SNE on data
-from the mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE
-places clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
-similarities and using proper initialization provides a more meaningful layout of the clusters. Neuronal types
-occupy one region of the space. Oligodendrocyte precursor cells (OPCs) are
-mainly progenitors to oligodendrocytes, but may also differentiate into neurons
-or astrocytes.} \label{fig:multiscale}
+  \caption{A comparison of standard and multi-scale t-SNE on data from the
+  mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE places
+  clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
+  similarities and using proper initialization provides a more meaningful
+  layout of the clusters. Neuronal types occupy one region of the space.
+  Oligodendrocyte precursor cells (OPCs) are mainly progenitors to
+  oligodendrocytes, but may also differentiate into neurons or astrocytes.}
+  \label{fig:multiscale}
 \end{figure}
 
-We also observed the important role of gene selection in crafting the
-reference embedding spaces. We found that when selecting an insufficient
-number of genes, the resulting visualizations display overly-fragmented clusters. When
-the selection is too broad and includes lowly expressed genes, the subclusters
-tend to overlap. These effects can all be attributed to sparseness of the data
-sets and may be intrinsic to single-cell data. In our studies, we found that
+We also observed the important role of gene selection in crafting the reference
+embedding spaces. We found that when selecting an insufficient number of genes,
+the resulting visualizations display overly-fragmented clusters. When the
+selection is too broad and includes lowly expressed genes, the subclusters tend
+to overlap. These effects can all be attributed to sparseness of the data sets
+and may be intrinsic to single-cell data. In our studies, we found that
 selection of 3,000 genes yields most informative visualizations
@@
   \includegraphics[width=\textwidth]{figures/hrvatin_embedding_tsne_genes.pdf}
-\caption{Gene selection plays an important role when constructing the reference
-embedding. {\bf (a)} Using too few genes results in fragmented clusters. {\bf (b)}
-Using an intermediate number of genes reveals clustering mostly consistent with
-cell annotations. {\bf (c)} Including all the genes may lead to
-under-clustering of the more specialized cell types.}
+  \caption{Gene selection plays an important role when constructing the
+  reference embedding. {\bf (a)} Using too few genes results in fragmented
+  clusters. {\bf (b)} Using an intermediate number of genes reveals clustering
+  mostly consistent with cell annotations. {\bf (c)} Including all the genes
+  may lead to under-clustering of the more specialized cell types.}
   \label{fig:gene_selection}
@@
 optimal positioning using the t-SNE loss function is much more dispersed and
-rightfully shows a more considerable variation in the secondary data. 
-Some points may also fall into the empty regions, while after optimization they typically
-move closer to same-typed groups. While the nearest neighbor-based approach
-always places new data points to their most similar cluster, our proposed
-t-SNE approach can identify extreme outliers and place them far from the
-reference cluster.
+rightfully shows a more considerable variation in the secondary data. Some
+points may also fall into the empty regions, while after optimization they
+typically move closer to same-typed groups. While the nearest neighbor-based
+approach always places new data points to their most similar cluster, our
+proposed t-SNE approach can identify extreme outliers and place them far from
+the reference cluster.
 
@@
   \includegraphics[width=\textwidth]{figures/optimization_retina.pdf}
-\caption{Comparison of data placement using the nearest neighbors approach from Kobak
-and Berens~\cite{art_of_using_tsne} and the optimized placement using our algorithm.
-{\bf (a)} Data points are placed to the median position of their 10 nearest
-neighbors in the reference set. {\bf (b)} Point positions are optimized,
-revealing a different, more dispersed placement that better reflects the
-variety of cells in the secondary data set.}
+  \caption{Comparison of data placement using the nearest neighbors approach
+  from Kobak and Berens~\cite{art_of_using_tsne} and the optimized placement
+  using our algorithm.  {\bf (a)} Data points are placed to the median position
+  of their 10 nearest neighbors in the reference set. {\bf (b)} Point positions
+  are optimized, revealing a different, more dispersed placement that better
+  reflects the variety of cells in the secondary data set.}
   \label{fig:optimization}
@@
 types in the secondary data set, possibly positioning them arbitrarily close to
-unrelated clusters. Procedures such as scmap were recently
-proposed to cope with such cases and identify the cells whose type is new and
-not included in the reference~\cite{scmap}.
+unrelated clusters. Procedures such as scmap were recently proposed to cope
+with such cases and identify the cells whose type is new and not included in
+the reference~\cite{scmap}.
 
-Our procedure and the resulting cell landscape are therefore affectected by the asymmetry
-of determining the reference and the secondary data sets. In practice, however,
-newly produced \em secondary \em data would be embedded into published and
-annotated reference landscapes. Scaling-up to a wider collection of cell
-types is driven by increasingly large collections of data, e.g the Human Cell
-Atlas initiative~\cite{hca}.
+Our procedure and the resulting cell landscape are therefore affectected by the
+asymmetry of determining the reference and the secondary data sets. In
+practice, however, newly produced \em secondary \em data would be embedded into
+published and annotated reference landscapes. Scaling-up to a wider collection
+of cell types is driven by increasingly large collections of data, e.g the
+Human Cell Atlas initiative~\cite{hca}.
 
@@
 two-dimensional visualization of the data that reveals the cellular diversity.
-While most dimensionality reduction
-techniques can be used to render such a visualization, different variants of
-t-SNE are most often used. Due to the ability to explore biological mechanisms
-at the individual cell level, single-cell studies are increasingly widespread, and
-their publications in the past couple of years are abundant. One of the central
-tasks in single-cell studies is the classification of new cells based on
-findings from previous studies. Such transfer of knowledge is often difficult
-due to batch effects present in data from different sources.
-Addressing batch effects by adapting and extending t-SNE, the prevailing method used
-to present single-cell data in two-dimensional visualization motivated the research
-presented in this paper.
+While most dimensionality reduction techniques can be used to render such a
+visualization, different variants of t-SNE are most often used. Due to the
+ability to explore biological mechanisms at the individual cell level,
+single-cell studies are increasingly widespread, and their publications in the
+past couple of years are abundant. One of the central tasks in single-cell
+studies is the classification of new cells based on findings from previous
+studies. Such transfer of knowledge is often difficult due to batch effects
+present in data from different sources.  Addressing batch effects by adapting
+and extending t-SNE, the prevailing method used to present single-cell data in
+two-dimensional visualization motivated the research presented in this paper.
 
-Our proposed approach uses a t-SNE embedding as a scaffold for the positioning of
-new cells within the visualization, and possibly for aiding in their
+Our proposed approach uses a t-SNE embedding as a scaffold for the positioning
+of new cells within the visualization, and possibly for aiding in their
 classification. The three case studies incorporating pairs of data sets from
@@
 proposed procedure can effectively deal with batch effects to construct
-visualizations that correctly map secondary data sets onto a reference data
-set from an independent study that possibly uses different experimental
-protocol. While we focused here on reference visualizations constructed using t-SNE,
-this approach can be applied using any existing two-dimensional visualization.
+visualizations that correctly map secondary data sets onto a reference data set
+from an independent study that possibly uses different experimental protocol.
+While we focused here on reference visualizations constructed using t-SNE, this
+approach can be applied using any existing two-dimensional visualization.
 

### 2019-08-13 paper/policar2019-tsne-embedding.tex
Shorten manuscript

@@
 set-specific clusters. To circumvent these batch effects, we propose an
-embedding procedure that takes a t-SNE visualization constructed on a
-reference data set and uses it as a scaffold for embedding new data. The new,
+embedding procedure that uses a t-SNE visualization constructed on a
+reference data set as a scaffold for embedding new data points. The new,
 secondary data is embedded one data point at the time. This prevents any
@@
 
-\keywords{Batch effects \and Embedding \and t-SNE \and Visualisation \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
+\keywords{Batch effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
 \end{abstract}
@@
 analysis. Here, it is expected that cells of the same type would cluster
-together in two-dimensional data visualisation~\cite{seurat}. For instance,
+together in two-dimensional data visualization~\cite{seurat}. For instance,
 Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
@@
   \caption{Batch effects are a driving factor of variation between the data
-  sets.  Depicted is a t-SNE visualisation of two pairs of data sets. In each
-  pair, the data sets share cell types, so it would be expected that the cells
-  from the reference data (blue) would mix with the cells in a secondary data
-  sets (orange).  Instead, t-SNE visualisation clusters data according to the
+  sets. We depict a t-SNE visualization of two pairs of data sets. In each
+  pair, the data sets share cell types, so we would expect cells
+  from the reference data (blue) to mix with the cells in a secondary data
+  sets (orange). Instead, t-SNE clusters data according to the
   data source.}
@@
 
-We propose a direct solution of rendering t-SNE visualizations that addresses
+We propose a direct solution of rendering t-SNE visualizations to address
 batch effects. Our approach treats one of the data sets as a {\em reference}
-and aims to embed the cells from another, {\em secondary data set} to a common
+and embeds the cells from another, {\em secondary data set} to a joint
 low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
-reference data set, and then use it as a scaffold for the embedding of the
+reference data set, which is then used as a scaffold to embed the
 secondary data. The key idea underpinning our approach is that secondary data
@@
 
-The proposed solution is implemented using a mapping of new data to an existing
+The proposed solution implements a mapping of new data into an existing
 t-SNE visualization. While the utility of such an algorithm was already hinted
@@
 We describe an end-to-end pipeline that uses fixed t-SNE coordinates as a
-scaffold for embedding new (secondary) data, enabling joint visualisation of
+scaffold for embedding new (secondary) data, enabling joint visualization of
 multiple data sources while mitigating batch effects. Our proposed approach
 starts by using t\nobreakdash -SNE to embed a reference data set, with the aim
-of constructing a two-dimensional visualisation to facilitate interpretation
+of constructing a two-dimensional visualization to facilitate interpretation
 and cluster classification. Then, the placement of each new sample is optimized
@@
 dimensional embedding $\mathbf{Y} = \left \{ \mathbf{y}_1, \mathbf{y}_2, \dots,
-\mathbf{y}_N \right \} \\\in \mathbb{R}^d$ where $d \ll D$, such that if points
+\mathbf{y}_N \right \} \in \mathbb{R}^d$ where $d \ll D$, such that if points
 $\mathbf{x}_i$ and $\mathbf{x}_j$ are close in the multi-dimensional space,
@@
 The time complexity needed to evaluate the similarities in
-Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making the application of the
-algorithm impractical for large data sets.  Hence, we adopt a recent approach
+Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making its application
+impractical for large data sets. We adopt a recent approach
 for low-rank approximation of gradients based on polynomial interpolation which
-reduces the time complexity of t-SNE to $\mathcal{O}(N)$. This approximation
+reduces its time complexity to $\mathcal{O}(N)$. This approximation
 enables the visualization of massive data sets, possibly containing millions of
@@
 long-range interactions between clusters. Larger values may result in a more
-globally consistent visualisations --- preserving distances on a large scale
+globally consistent visualizations --- preserving distances on a large scale
 and organizing clusters in a more meaningful way --- but can lead to merging
@@
 $l$. In our experiments, we used a mixture of two Gaussian kernels with
-perplexity values of 50 and 500. We note that a similar formulation of
+perplexity values of 50 and 500. A similar formulation of
 multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
@@
 \mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
-the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
+the secondary data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
 \mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
@@
 reference set (Table~\ref{tab:data sets}). The cells in the data sets
-originates from the following three tissues:
+originate from the following three tissues:
 
@@
 
-\item[Human pancreas.] The data set from Baron \etal~\cite{baron2016} was
-created as an atlas of pancreatic cell types. We used this set as a reference
+\item[Human pancreas.] Baron \etal~\cite{baron2016} created
+an atlas of pancreatic cell types. We used this set as a reference
 for data from Xin \etal~\cite{xin2016}, who examined transcriptional
@@
 
-\item[Mouse retina.] The data set from Macosko \etal~\cite{macosko2015} was
-created as an atlas of mouse retinal cell types. We used this as a reference
+\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
+an atlas of mouse retinal cell types. We used this as a reference
 for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
@@
 
-\noindent The mean $\log_2$ expression of the genes is computed from all the
-cells where gene was expressed:
+\noindent The mean $\log_2$ expression of gene $g$ considers only cells
+expressing $g$:
 
@@
 the cells from the secondary data sets were positioned in the cluster of
-same-typed reference cells, providing strong evidence of the success of the
-proposed approach. There are some deviations to these observations; for
+same-typed reference cells, providing strong evidence of the success of our approach. There are some deviations to these observations; for
 instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
@@
 these types of cells. Examples of such erroneous placements can be found in
-other figures as well, but they are not common and constitute less then 5\% of
-the cells (less than  5\% in brain, \%1 in pancreas and \%2 in retina
-secondary data sets).
+other figures as well, but are uncommon and constitute less then 5\% of
+the cells (less than 5\% in brain, 1\% in pancreas and 2\% in retina
+secondary data).
 
@@
 visualizations strikingly different from ours. For example, compare the
-visualisations from Fig.~\ref{fig:batch_effect}.a and
+visualizations from Fig.~\ref{fig:batch_effect}.a and
 Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
@@
 
-There are additional modifications that we use in the embedding of the
-secondary data set that were recently proposed and enhance the original t-SNE
-visualization.  One important extension is the use of multi-scale similarities
-that, besides local ordering of the data points, includes global optimization
-of cluster placement. For illustration, consider visualizations with standard
+We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
+One important extension is the use of multi-scale similarities, which, in addition to finding local structure, also ensure consistent organization of the resulting clusters.
+For illustration, consider visualizations with standard
 and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that
@@
 
-The proposed method assumes that all cell types from the secondary data set are
-present in the reference. The proposed method would fail to reveal novel cell
+Our approach assumes that all cell types from the secondary data set are
+present in the reference. The method would fail to reveal novel cell
 types in the secondary data set, possibly positioning them arbitrarily close to
@@
 
-Our procedure and the resulting cell landscape are therefore affectected by the
-asymmetry of determining the reference and the secondary data sets. In
-practice, however, newly produced \em secondary \em data would be embedded into
-published and annotated reference landscapes. Scaling-up to a wider collection
-of cell types is driven by increasingly large collections of data, e.g the
-Human Cell Atlas initiative~\cite{hca}.
+Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
+In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
+Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.
 
@@
 are, together with the data, available in an open
-repository~\footnote{https://github.com/biolab/tsne-embedding}. All experiments
-were run using openTSNE~\footnote{https://github.com/pavlin-policar/openTSNE},
-our open and extensible t-SNE library for Python.
+repository\footnote{https://github.com/biolab/tsne-embedding}. All experiments
+were run using {\tt openTSNE},
+our open and extensible t-SNE library for Python~\cite{opentsne}.
 
@@
 Almost all recent publications of single-cell studies begin with a
-two-dimensional visualization of the data that reveals the cellular diversity.
-While most dimensionality reduction techniques can be used to render such a
-visualization, different variants of t-SNE are most often used. Due to the
-ability to explore biological mechanisms at the individual cell level,
-single-cell studies are increasingly widespread, and their publications in the
-past couple of years are abundant. One of the central tasks in single-cell
+two-dimensional visualization of the data that reveals cellular diversity.
+While many dimensionality reduction techniques are available, different 
+variants of t-SNE are most often used to produce such
+visualizations. Single-cell studies enable the exploration of biological mechanisms at a
+cellular level, and their publications in the past couple of years are abundant.
+One of the central tasks in single-cell
 studies is the classification of new cells based on findings from previous
@@
 and extending t-SNE, the prevailing method used to present single-cell data in
-two-dimensional visualization motivated the research presented in this paper.
+two-dimensional visualization, motivated the research presented in this paper.
 
@@
 Fund and the Slovenian Ministry of Education, Science and Sport. We would also
-like to thank Dmitry Kobak for discussions on t-SNE.
+like to thank Dmitry Kobak for many helpful discussions on t-SNE.
 

### 2021-05-13 paper-revision/response-to-reviewers.tex
Add revision files

\documentclass{article}

\usepackage[T1]{fontenc}
\usepackage{natbib}
\usepackage{todonotes}
\usepackage{gensymb}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{alltt}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{comment}
\usepackage{graphicx}

\usepackage{xr}[]
\externaldocument{policar2019-tsne-embedding}
\usepackage{url}

\renewcommand{\thetable}{\Alph{table}}
\renewcommand{\thefigure}{\Alph{figure}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}
\newcommand{\beqs}{\begin{equation}}
\newcommand{\eeqs}{\end{equation}}
\newcommand{\mx}[1]{\ensuremath{\boldsymbol{\mathrm{{#1}}}}}   % Simple matrix format.

% \newcommand{\review}[1]{\noindent{{\textbf{\textit{#1}}}}\vspace{3mm}}
% \definecolor{colframe}{rgb}{0.01,0.199,0.1}

\newcounter{rtaskno}
\newcommand{\rtask}[1]{\refstepcounter{rtaskno}\label{#1}}

\newcommand{\reviewc}[2]{\begin{leftbar}\noindent\rtask{#1}\ref{#1}. #2\end{leftbar}}
\newcommand{\review}[1]{\begin{leftbar}\noindent #1\end{leftbar}}
\newcommand{\answer}[1]{\noindent #1\\[0mm]}
\newcommand{\copyquote}[1]{\vspace{2mm} {\color{blue} #1}\vspace{2mm}}

\newlength{\leftbarwidth}
\setlength{\leftbarwidth}{3pt}
\newlength{\leftbarsep}
\setlength{\leftbarsep}{10pt}
\renewenvironment{leftbar}{%
    \def\FrameCommand{{\color{black}{\vrule width \leftbarwidth\relax\hspace {\leftbarsep}}}}%
    \MakeFramed {\advance \hsize -\width \FrameRestore }%
}{\endMakeFramed}

% \usepackage[colorlinks=True,citecolor=black]{hyperref}

\begin{document}

\noindent{\bf Poli\v{c}ar, Stra\v{z}ar \& Zupan: Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}

\vspace{5mm}
\noindent {\bf RESPONSE TO THE REVIEWERS}

\vspace{5mm}

\noindent Dear Editor and the Reviewers, \\

\vspace{3mm}

\noindent we thank for the comprehensive feedback that helped to improve the
manuscript, sufficient time to implement the changes, and the permission to
submit a revised version.

\vspace{3mm}

All of the reviewer remarks are addressed below with reference to the
corresponding changes in the manuscript.  In case of overlapping remarks from
different reviewers, we refer the reader to the relevant enumerated answer. A
differential document is provided to facilitate the comparison with the initial
submission. The changes to the data, code and the instructions to reproduce all
the experiments are available in the repository:
\url{https://github.com/biolab/tsne-embedding}.

\vspace{3mm}

\begin{flushright}
\noindent Kind regards, \\ 
\vspace{2mm}
Pavlin Poli\v{c}ar \\
Martin Stra\v{z}ar \\
Bla\v{z} Zupan \\ 
\vspace{2mm}
University of Ljubljana, Slovenia 
\end{flushright}

\vspace{10mm}

\subsection*{Reviewer \#1}

\reviewc{r1q1}{Most important is to add in theoretical part investigation about which dataset can be a reference dataset. It is necessary because it is crucial in estimation of level of trust worth.
Some information about it is provided in testing part but only in form of general statement on specific case (based on some data sets), but it is necessary to have a procedure/test when given data set is a good reference data set.}

\answer{
	We thank the reviewer for this comment and have expanded our discussion of potential failure cases in the ``Experiments and Discussion'' section. We have added an example of a failure case and have pointed out that care must be taken when using our procedure.

	A theoretical investigation of which data set can be used as a reference is beyond the scope of this paper. Conceptually, our approach for classification is very similar to a $k$-nearest neighbor classifier. It would be very difficult to theoretically quantify which reference data sets would be appropriate for such classifiers.
}

\reviewc{r1q2}{Page 1
Bibliography:
positions 3 and 4 is not about PCA and MDS. Please provide appropriate citations for PCA and MDS.}

\answer{We thank the reviewer for catching this mistake. We have fixed these citations.}

\reviewc{r1q3}{
Page 6
How do you select L and $\sigma_{i,l}$?
}

\answer{
	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough comparison. We moved this section to the appendix.
}

\reviewc{r1q4}{page 9
The "I(.)" in Eq. 11 is not defined.}

\answer{
	We've added a short sentence, explaining that this is the indicator function. We moved this section to the appendix.
}

\reviewc{r1q5}{Please precise what do you mean by the right side of Eq. 12 (the meaning of brackets and "|").}

\answer{We have modified the text so that the equations in this section are clearer.}

\reviewc{r1q6}{Generally subsection 3.3 is not as clear as it should be. Please rewrite it.}

\answer{We agree that this subsection could be more clear. We have modified some parts of it and added sentences which should clarify the gene selection procedure.}

\reviewc{r1q7}{Figs 1-4
It would be much more readable when Figures would be bigger. Please provide subfigures vertically, not horizontally.}

\answer{}

\reviewc{r1q8}{GENERAL statement: provide better quality of all Figures (better resolution, now we see large pixels too early...)}

\answer{}

\reviewc{r1q9}{Fig 5
Please modify comparison presented in Fig 5.
Please provide 4 subfigures:
- t-SNE without PCA and without multiscale
- t-SNE with PCA and without multiscale
- t-SNE without PCA and with multiscale
- t-SNE with PCA and with multiscale
only in such case we can be sure what is real source of success.
It would be fine to see such comparison on (at least) two datasets. (!)}

\answer{
	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough compoarison.
}

\reviewc{r1q10}{Fig 6
Authors claim that in case of sub-figure c all genes may lead to under-clustering but the figures b and c are very similar in the number of clusters.}

\answer{
	How to select the number of genes is pretty irrelevant to our work. This is a question of single-cell processing pipeline, which we do not care about in this work.
}

\subsection*{Reviewer \#2}

\reviewc{r2q1}{Since the focus, from an application point of view, is on bioinformatics and specifically on batch effects, I think that a more detailed description of this phenomenon should be reported in the introduction. }

\answer{
	We've addressed this in related work, where we've added a paragraph dedicated to batch effects.
}

\reviewc{r2q2}{The structure of the paper can significantly be improved. In particular:

- A section devoted to the description of related work should be introduced, reporting existing work on both the application and methodological viewpoints. In this respect, some descriptions currently reported in the experiments sections (see, for example, the description of scMap-cluster) should be moved in this section. Moreover, I actually can see a strong similarity with transfer learning methods, specifically with those focusing on domain adaptation. I suggest the authors to consider some relevant recent works in this field in their discussion (see, for example, the survey 10.1186/s40537-016-0043-6), also specifically focusing on bioinformatics (see, for example, the work 10.1093/bioinformatics/btz781).

- Related to the previous comment, some details related to existing approaches should be removed (or strongly simplified) from the section devoted to experiments, and moved to a new section devoted to related work.

- Some parameter values reported in the methods section should be moved in the experiments, since they are part of the experimental setting.}

\answer{We thank the reviewer for this comment. We have restructured the paper.

	We have added a ``Related work'' section, where we describe batch-effects in more detail. We have also moved the description of scMap-Cluster from the ``Experiments'' section to this section. While it is true that these approaches are, in essence, transfer learning, the data sets produced by single-cell RNA-seq are sufficiently different from data sets from other fields of machine learning that they requires specialized methods, which we describe in the ``Related work'' section.
}

\reviewc{r2q3}{The current paper is an extension of a conference paper submitted to the conference Discovery Science 2019. The authors should explicitly provide in the text the actual novel contributions with respect to the conference version.}

\answer{}

\reviewc{r2q4}{The authors should better emphasize the contribution from a methodological viewpoint. Indeed, without a proper emphasis on this aspect, the paper seems to be the application of an existing tool (namely, openTSNE) to a new dataset. If the contribution from a methodological viewpoint is actually limited, the authors should still put some effort in emphasizing the advantages of this approach in this specific domain.}

\answer{
	We have changed the working in ``Availability and Implementation'' section, which emphasizes that the extension we describe in this paper was developed by us and later incorporated into \textsf{openTSNE}. As this is the only mention of \textsf{openTSNE} in the manuscript, we believe there can be no further confusion in this respect.
}

\reviewc{r2q5}{PCA-based initialization seems to help a lot. I would appreciate to see a comparison of the results with those achievable with the simple application of PCA with just 2 principal components.}

\answer{
	This isn't part of our work, and has been moved to the supplement. Please, refer to the original publication for a thorough comparison. However, we would also refer the reviewer to a recent comment by Kobak and Linderman ``Initialization is critical for preserving global data structure in both t-SNE and UMAP'', \textit{Nature Biotechnology}, 2021, which demonstrates that good initialization schemes are important for both t-SNE and UMAP.
}

\reviewc{r2q6}{At the beginning of Section 3.2, a quite long preprocessing pipeline is shown. Actually, I would like to understand the underline motivations behind that. Few additional sentences would help in this respect.}

\answer{
	This is a question of single-cell processing pipeline, which is not the focus of our work. We provide a reference to the original publication presenting the pipeline in the text.
}

\subsection*{Reviewer \#3}

\reviewc{r3q1}{The research heavily relies on the ref. [15]. Although the authors claimed that they provide both practical and theoretical grounded implementation, the contribution in terms of theoretical justification is not convincing enough. The equations, given in Section 2, closely correlate with those provided in [15], except for two small modifications - replacing x and y with v and w. }

\answer{
	We agree that the original manuscript included sections that are important to the single-cell analysis pipeline which are perhaps not the most relevant to our particular contribution i.e. the extension of t-SNE to allow embedding new data points into existing embeddings. For this particular task, the details of the single-cell analysis are not crucial. While it is true that our single-cell pipeline relies on the work from [15], we have moved most of these sections to the appendix, as they are not directly related to our work. We believe that this improves the readability of our paper. However, the work done in [15] introduced ideas which are now considered standard practice for t-SNE visualizations, so a brief review of the most important concepts is kept in Section 3.1.

	Regarding the contribution in terms of theoretical justification: the extension is indeed very minor, and requires changing only two variables, which results in different interactions being calculated. However, additional minor but important changes need to made, including the omission of the symmetrization step and a minor change in the gradient calculation.

	Perhaps more important is the practical aspect of our work i.e. how do we get this to work. While the equations change only slightly, a straightforward application of our approach using the currently accepted t-SNE parameter settings will result in wildly unstable optimization, so special care must be taken to mitigate this erratic behaviour. We describe this in the last paragraph of Section 3.2, where we emphasize the importance of lowering the learning rate and introducing optimization tricks e.g. gradient clipping. This is especially important for the approximation schemes used today by virtually all t-SNE implementations, as a poor parameter setting will result in points ``shooting-off'' far away from the reference points and will dramatically impact the runtime of these approximation schemes. Without these modifications, the t-SNE transform functionality becomes unusable in virtually all settings.
}

\reviewc{r3q2}{From the Abstract, it is not immediately clear what the reference and secondary datasets are. The authors explained it later in the main body, but it seems reasonable to introduce the terms earlier to keep readers interested. Further, the authors claim that they show "the utility of the proposed approach by analyzing 6 recently published ... datasets … with up to tens of thousands of cells and thousands of genes ..."  and so on. Mentioning more specific quantitative results with regard to particular datasets, in comparison with the known results, would be desirable. }

\answer{
	We have slightly reworded the abstract to indicate that the secondary data set contains new, unseen points.

	We disagree that mentioning more qualitative results would improve this manuscript. There are no robust, agreed-up baselines in the single-cell literature for batch effect removal in general. Only a handful of methods for cross-data set cell-type classification methods have been proposed, and none of them are directly related to our approach. This manuscript highlights that a simple extension to a visualization technique, t-SNE, performs competitively to specialized machine learning and bioinformatics methods for cell-type classification. These methods are not directly comparable, and our aim is not to improve the accuracy of such approaches. We have, however, reworded the last paragraph of Section 4.6 to emphasize this point.
}

\reviewc{r3q3}{The Results and Discussion section is difficult to follow. Please, if possible, consider splitting it in subsections.}

\answer{
	We thank the reviewer for this comment and we agree that splitting the section into subsections would improve the readability of the manuscript. We have split up the ``Results and Discussion'' section into several subsections.
}

\reviewc{r3q4}{Consider replacing the legends in Fig. 1 with the names or descriptions of the datasets to increase the readability. }

\answer{
	The legends of Figure 1 show the names of the data sets, or rather, their original publication. The figure demonstrates that two data sets with shared cell types cluster primarily by data set and not by cell-type. In this instance one would reasonably assume and even hope that the samples would cluster by cell-type and that the data set source would not interfere with the visualization. We feel that a description of the data sets is unnecessary here, as it does not help illustrate our point, and would needlessly burden the reader with biological details.
}

\reviewc{r3q5}{Please extend Section 3.5, which includes only two sentences, with more details of implementation or consider moving this text into Section 3.4 (possibly at the beginning or at the end).}

\answer{We agree that this is not the most suitable location for this information, and we have moved it to the end of the manuscript, after the conclusion and before the acknowledgments, as is standard practice in many journals.}

\reviewc{r3q6}{Please update the preprint [15] with the published manuscript.}

\answer{The thank the reviewer for the reminder. We have updated the citation.}

\end{document}

### 2021-05-13 paper-revision/diffbase.tex
Add revision files

--- 2019-08-13 paper/policar2019-tsne-embedding.tex
@@
 %
-\institute{University of Ljubljana, SI-1000 Ljubljana, Slovenia
+\institute{University of Ljubljana, Faculty of Computer and Information Science, Večna pot 113, Ljubljana
 \email{\{pavlin.policar,martin.strazar,blaz.zupan\}@fri.uni-lj.si}\\[2pt]
@@
 
-Dimensionality reduction techniques, such as t-SNE, can construct informative
-visualizations of high-dimensional data. When working with multiple data
-sets, a straightforward application of these methods often fails; instead of
-revealing underlying classes, the resulting visualizations expose data
-set-specific clusters. To circumvent these batch effects, we propose an
-embedding procedure that uses a t-SNE visualization constructed on a
-reference data set as a scaffold for embedding new data points. The new,
-secondary data is embedded one data point at the time. This prevents any
-interactions between instances in the secondary data and implicitly mitigates
-batch effects. We demonstrate the utility of this approach by analyzing six
-recently published single-cell gene expression data sets with up to tens of
-thousands of cells and thousands of genes. The batch effects are particularly
-strong as the used data comes from different institutions and was obtained
-using different experimental protocols. The visualizations constructed by our
-proposed approach are cleared of batch effects, and the cells from secondary
-data sets correctly co-cluster with cells of the same type from the primary
-data.
+Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data.
+When jointly visualising multiple data sets, a straightforward application of these methods often fails; instead of revealing underlying classes, the resulting visualizations expose dataset-specific clusters.
+To circumvent these batch effects, we propose an embedding procedure that uses a t-SNE visualization constructed on a reference data set as a scaffold for embedding new data points.
+Each data instance in the secondary data is embedded independently and does not change the reference embedding.
+This prevents any interactions between instances in the secondary data and implicitly mitigates batch effects.
+We demonstrate the utility of this approach by analyzing six recently published single-cell gene expression data sets with up to tens of thousands of cells and thousands of genes.
+The batch effects in our studies are particularly strong as the data comes from different institutions using different experimental protocols.
+The visualizations constructed by our proposed approach are clear of batch effects, and the cells from secondary data sets correctly co-cluster with cells of the same type from the primary data.
+We also show the predictive power of our visual classification approach in t-SNE space matches the accuracy of machine learning techniques that consider the entire compendium of features that profile single cells.
 
-\keywords{Batch effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
+\keywords{Batch Effects \and Embedding \and t-SNE \and Visualization \and Single-Cell Transcriptomics \and Data Integration \and Domain Adaptation.}
 \end{abstract}
@@
 source-specific biases.  The focus of this work is single-cell genomics, a
-domain that was selected due high biomedical relevance and and abundance of
+domain that was selected due to high biomedical relevance and abundance of
 recently published data. Single-cell RNA sequencing (scRNA-seq) data sets are
@@
 \begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{figures/batch_effect.pdf}
+  \includegraphics[width=\textwidth]{batch_effect.pdf}
   \caption{Batch effects are a driving factor of variation between the data
@@
 the entire analysis pipeline must be rerun, usually resulting in a different
-layout and clusters that have little resemblance to original visualization and
+embedding layout and clusters that have little resemblance to original visualization and
 thus require reinterpretation.
@@
 batch effects. Our approach treats one of the data sets as a {\em reference}
-and embeds the cells from another, {\em secondary data set} to a joint
+and embeds the cells from another, {\em secondary data set} to a reference-defined
 low-dimensional space.  We construct a t\nobreakdash -SNE embedding using the
@@
 \begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/transform_brain.pdf}
+  \includegraphics[width=\textwidth]{transform_brain.pdf}
   \caption{A two-dimensional embedding of a reference containing brain cells
-  (a) and the corresponding mapping of secondary data containing hypothalamic
-  cells (b).  The majority of hypothalamic cells were mapped to their
+  \textbf{(a)} and the corresponding mapping of secondary data containing hypothalamic
+  cells \textbf{(b)}.  The majority of hypothalamic cells were mapped to their
   corresponding reference cluster. For instance, astrocyte cells marked with
@@
 t-SNE visualization. While the utility of such an algorithm was already hinted
-at in recent publication~\cite{art_of_using_tsne}, here we provide its
+at in recent publication~\cite{art_of_using_tsne}, we here provide its
 practical and theoretically-grounded implementation. Considering the abundance
@@
 specific to the secondary data. Below, we start with a summary of t-SNE and its
-extensions (Sec.~\ref{sec:tsne}, introducing the relevant notation, upon which
+extensions (Sec.~\ref{sec:tsne}), introducing the relevant notation, upon which
 we base our secondary data embedding approach (Sec.~\ref{sec:transfer}).
 
-\subsection{Data embedding by t-SNE and its extensions\label{sec:tsne}}
+\subsection{Data Embedding by t-SNE and Its Extensions\label{sec:tsne}}
 
@@
 
-\subsection{Adding new data points to reference embedding\label{sec:transfer}}
+When using t-SNE on larger data sets, the standard learning rate $\eta = 200$
+has been shown to lead to slower convergence and requires more iterations to
+achieve consistent embeddings~\cite{belkina2019automated}. We follow the
+recommendation of Belkina \etal ~and use a higher learning rate $\eta = N / 12$
+when visualizing larger data sets.
+
+\subsection{Adding New Data Points to Reference Embedding\label{sec:transfer}}
 
@@
 however, by adapting the same polynomial interpolation based approximation,
-this is reduced to $\mathcal{O}(\max \{ N, M \})$.
+this is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
+reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding remains fixed.
 
@@
 embedding. In our experiments, we found that decreasing the learning rate to
-$\eta \sim 0.1$ produces stable solutions. This is especially important when
+$\eta \sim 0.1$ produces stable solutions. Alternatively, we can employ gradient
+clipping to achieve similar behaviour. This is especially important when
 using the interpolation-based approximation, which places a grid of
@@
 \end{center}
-\caption{Data sets used in our experiments. The first data set in a pair
+\caption{Data sets used in our experiments. The first data set in each pair
 (Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
 relied on the quality control and annotations from the original publication
-and report the humber of cell types after preprocessinng.  The cell
+and report the number of cell types after preprocessinng.  The cell
 annotations were made consistent to annotations from the Cell
@@
 
-\subsection{Single-cell data preprocessing pipeline}
+\subsection{Results and Discussion}
+\label{sec:results}
+
+Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
+\ref{fig:transform_retina} show the embeddings of the reference data sets and
+their corresponding embeddings of the secondary data sets. In all the figures,
+the cells from the secondary data sets were positioned in the cluster of
+same-typed reference cells, providing strong evidence of the success of our approach. There are some deviations to these observations; for
+instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
+cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
+annotation criteria by different authors, or due to inherent similarities of
+these types of cells. Examples of such erroneous placements can be found in
+other figures as well, but are uncommon and constitute less then 5\% of
+the cells (less than 5\% in brain, 1\% in pancreas and 2\% in retina
+secondary data).
+
+Notice that we could simulate the split between reference and secondary data
+sets using one data set only and perform cross-validation, however this type of
+experiment would not incorporate batch effects. We want to remind the reader
+that handling batch effects were central to our endeavour and that the
+disregard of this effect could lead to overly-optimistic results and data
+visualizations strikingly different from ours. For example, compare the
+visualizations from Fig.~\ref{fig:batch_effect}.a and
+Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
+\ref{fig:transform_pancreas}.b.
+
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{transform_pancreas.pdf}
+  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
+  cells from the same tissue from Xin \etal~\cite{xin2016}. Just like in
+  Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
+  secondary data set were correctly mapped to the same-typed cluster of
+  reference cells.}
+  \label{fig:transform_pancreas}
+\end{figure}
+
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{transform_retina.pdf}
+  \caption{An embedding of a large reference of retinal cells from Macosco
+  \etal~\cite{macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
+  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} \textbf{(b)}. We use
+  colors consistent with the study by Macosko \etal}
+  \label{fig:transform_retina}
+\end{figure}
+
+We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
+Kobak and Linderman have shown that the global consistency of embeddings produced by
+popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
+more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
+as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
+Another important extension is the use of multi-scale similarities, which, in addition
+to considering short range interactions, also models wider point neighborhoods. Coupled
+with PCA-based initializations, this produces even more meaningful visualizations where
+clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
+developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
+is less apparent when using PCA-based initialization alone in
+Fig.~\ref{fig:multiscale}.b.
+
+We also observed the important role of gene selection in crafting the reference
+embedding spaces. We found that when selecting an insufficient number of genes,
+the resulting visualizations display overly-fragmented clusters. When the
+selection is too broad and includes lowly expressed genes, the subclusters tend
+to overlap. These effects can all be attributed to sparseness of the data sets
+and may be intrinsic to single-cell data. In our studies, we found that
+selection of 3,000 genes yields most informative visualizations
+(Fig.~\ref{fig:gene_selection}).
+
+In principle, our theoretically-grounded embedding of secondary data into the
+scaffold defined by the reference embedding could be simplified with the
+application of the nearest neighbors-based procedure. For example, while
+describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
+proposed positioning new points into a known embedding by placing them in the
+median position of their 10 nearest neighbors, where the neighborhood was
+estimated in the original data space. Notice that we use this approach as well,
+but only for the initialization of positions of new data instances that are
+subject to further optimization. Despite both nearest-neighbors search and
+t-SNE optimization can be computed in linear time, the former dominates the
+runtime (mouse retina example; 44,808 reference, 26,830 secondary cells,
+9min NN-search, 13sec optimization).
+
+Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based 
+positioning alone is insufficient. We construct a reference embedding using only
+neurons from Hrvatin \etal~\cite{hrvatin2018} (Fig.~\ref{fig:optimization}.a)
+and use that to position neuronal
+cells from the data set from Campbell \etal~\cite{campbell2017}. We utilize the
+weighted mean and median positions to initialize point positions from the secondary
+data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
+After initialization, we optimize point positions using the procedure described
+above for 500 iterations. The resulting visualizations from both initializations
+are visually very similar, indicating stable convergence. We show one of the resulting
+visualizations in Fig.~\ref{fig:optimization}.b.
+
+Notice that both neighbor-based initialization schemes generally position data points
+such that their classification is unclear. Median-based initialization produces
+a sort of grid-like structure, while median based initialization positions the points
+almost continuously across the embedding space. Optimization reveals strong
+correspondence of several points to reference-defined clusters, while other points
+from the secondary data set are pushed away from their initial clusters, possibly
+indicating dissimilarity.
+
+\begin{figure}[htbp]
+  \includegraphics[width=\textwidth]{optimization.pdf}
+  \caption{Comparison of different initialization schemes for positioning new data
+  points onto reference embeddings. {\bf (a)} We construct a reference embedding
+  using only neuronal subtypes from Hrvatin \etal~\cite{hrvatin2018}. {\bf (b)}
+  We position neuronal cells from Campbell \etal~\cite{campbell2017} using the
+  median initizaliation scheme from Kobak and Berens~\cite{art_of_using_tsne} and
+  run optimization for 500 iterations. Compare the optimized embedding with the 
+  initial median initialization {\bf (c)} or by using a simple weighted mean
+  initialization {\bf (d)}.}
+  \label{fig:optimization}
+\end{figure}
+
+The approach assumes that all cell types from the secondary data set are
+present in the reference. The method may fail to reveal novel cell
+types in the secondary data set, possibly positioning them arbitrarily close to
+unrelated clusters. Procedures such as scmap were recently proposed to cope
+with such cases and identify the cells whose type is new and not included in
+the reference~\cite{scmap}.
+
+Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
+In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
+Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.
+
+To quantitatively evaluate the predictive accuracy of the described procedure, we fit $k$-nearest neighbors classifiers on each reference t-SNE embedding from Figs.~\ref{fig:transform_brain}.a, \ref{fig:transform_pancreas}.a and \ref{fig:transform_retina}.a and use them to predict the cell types for the secondary data set embeddings from Figs.~\ref{fig:transform_brain}.b, \ref{fig:transform_pancreas}.b and \ref{fig:transform_retina}.b.
+The accuracy measures are reported in Table~\ref{tab:scores}.
+Our procedure of embedding new data points into two-dimensional t-SNE plane results in similar accuracy to approaches like random forests that use full compendium of cell-characterizing features.
+The results indicate that positioning of new cells onto a cell visualisation plane is not only indicative but also an accurate instrument for cell type characterization.
+
+\begin{table}[ht]
+\centering
+  \setlength\tabcolsep{6pt}
+  \begin{tabular}{c c r r}
+  \toprule
+  Tissue & Method & Accuracy & ARI \\
+  \toprule
+  \multirow{4}{*}{mouse brain} & t-SNE & 0.96 & 0.93 \\
+   & KNN & 0.96 & 0.93 \\
+   & Random Forest & \textbf{0.98} & \textbf{0.96} \\
+   & scMap-Cluster & 0.66 & 0.70 \\
+  \midrule
+  \multirow{4}{*}{human pancreas} & t-SNE & \textbf{0.99} & \textbf{0.99} \\
+   & KNN & \textbf{0.99} & 0.98 \\
+   & Random Forest & 0.96 & 0.89 \\
+   & scMap-Cluster & 0.95 & 0.93 \\
+  \midrule
+  \multirow{4}{*}{mouse retina} & t-SNE & \textbf{0.99} & 0.94 \\
+   & KNN & \textbf{0.99} & 0.96 \\
+   & Random Forest & \textbf{0.99} & \textbf{0.99} \\
+   & scMap-Cluster & 0.88 & 0.59 \\
+  \bottomrule
+  \end{tabular}
+  \caption{We compare our approach (t-SNE) to three other methods, evaluating performance using classification accuracy and the adjusted rand index (ARI).
+Notice that while the proposed approach classifies cells only based on their position on the two-dimensional plane, it performs comparably to other methods that use full compendium of features (gene expressions) that characterize the cells.}
+  \label{tab:scores}
+\end{table}
+
+We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{scmap}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data.
+scMap-Cluster uses three correlation-based distance measures and uses a voting scheme to perform classification.
+A threshold may also be set so that new cell types may be labeled as {\tt unrecognized}, but since our secondary data sets were chosen such that this does not happen, this was disabled.
+For the two machine learning approaches, we apply the typical single-cell preprocessing pipeline described in Sec.~\ref{sec:sc-pipeline}, i.e., library-size normalization, log-transformation, and select 1,000 most informative genes.
+Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest neighbors in the KNN model.
+We used 100 trees in the random forest ensemble.
+The models were fit on the reference data set, and no hyper-parameter tuning was performed.
+
+Surprisingly, both the random forest and $k$-nearest neighbor models outperform scMap-Cluster, which is specifically tailored to scRNA-seq data.
+However, these results may be skewed, as, in our examples, all the cell-types from the secondary data set were present in the reference data set.
+One of the core features of scMap-Cluster is the detection of novel cell types, which none of the other methods support.
+In other words, the other three methods would always assign a cell-type to a given cell, regardless of cell origin.
+Additionally, scMap-Cluster was primarily designed and tested on data sets produced by full-length sequencing protocols, which tend to detect a much higher number of molecules than other, sequencing protocols based on unique molecular identifiers (UMI).
+These two classes of sequencing protocols produce data sets with different sparsity and variance characteristics.
+This is consistent with the results in Table~\ref{tab:scores}, as only the data sets from the human pancreas, were produced using a full-length sequencing protocol, where scMap-Cluster achieves reasonably high accuracy.
+
+The aim of t-SNE is to construct embeddings, in which neighborhoods are preserved, therefore it is unsurprising that the accuracy of our t-SNE based approach is largely consistent with the $k$-nearest neighbors model.
+Our approach is comparable to the other models in terms of accuracy, but we would note that the goal of t-SNE embeddings is to serve as visual aids in exploratory data analysis.
+However, our procedure, in addition to providing the end-user with a cell-type prediction, allows the user to examine the low-dimensional embedding space, which may provide richer insight and interpretation of the resulting predictions.
+
+\section{Conclusion}
+
+Almost all recent publications of single-cell studies begin with a
+two-dimensional visualization of the data that reveals cellular diversity.
+While many dimensionality reduction techniques are available, different 
+variants of t-SNE are most often used to produce such
+visualizations. Single-cell studies enable the exploration of biological mechanisms at a
+cellular level, and their publications in the past couple of years are abundant.
+One of the central tasks in single-cell
+studies is the classification of new cells based on findings from previous
+studies. Such transfer of knowledge is often difficult due to batch effects
+present in data from different sources. Addressing batch effects by adapting
+and extending t-SNE, the prevailing method used to present single-cell data in
+two-dimensional visualization, motivated the research presented in this paper.
+
+The proposed approach uses a t-SNE embedding as a scaffold for the positioning of new cells within the visualization, and possibly for aiding in their classification.
+The three case studies incorporating pairs of data sets from different domains but with similar classifications demonstrate that our proposed procedure can effectively deal with batch effects to construct visualizations that correctly map secondary data sets onto an embedding of the data from an independent study that possibly uses different experimental protocol.
+We quantitatively evaluate the predictive accuracy of our approach by fitting a $k$-nearest neighbors model on the resulting two-dimensional embeddings and compare its predictive accuracy to other machine learning methods that use the entire compendium of gene expressions that characterize the cells.
+Experiments show that our approach is successful in predicting cell types and performs comparably to other methods.
+This encouraging result indicates that by using our procedure, scientists can quickly and accurately determine the composition of new data by merely visualizing and inspecting resulting visualizations.
+While we focused here on reference visualizations constructed using t-SNE, this approach can be applied using any existing two-dimensional visualization.
+
+\subsubsection*{Implementation\label{sec:implementation}}
+
+The procedures described in this paper are provided as Python notebooks that
+are, together with the data, available in an open
+repository\footnote{https://github.com/biolab/tsne-embedding}. All experiments
+were run using {\tt openTSNE},
+our open and extensible t-SNE library for Python~\cite{opentsne}.
+
+\subsubsection*{Acknowledgements}
+
+This work was supported by the Slovenian Research Agen\-cy Program Grant P2-0209,
+and by the BioPharm.SI project supported from European Regional Development
+Fund and the Slovenian Ministry of Education, Science and Sport. We would also
+like to thank Dmitry Kobak for many helpful discussions on t-SNE.
+
+% \bibliographystyle{splncs04}
+\bibliographystyle{unsrt}
+\bibliography{references}
+
+\appendix
+
+\section{Single-Cell Data Preprocessing Pipeline\label{sec:sc-pipeline}}
 
@@
 
-\subsection{Gene selection\label{sec:gene-selection}}
+\section{Gene Selection\label{sec:gene-selection}}
+
+% TODO: This section (3.3) is not clear
 
@@
 
-\subsection{Results and Discussion}
-\label{sec:results}
-
-Figs.~\ref{fig:transform_brain}, \ref{fig:transform_pancreas}, and
-\ref{fig:transform_retina} show the embeddings of the reference data sets and
-their corresponding embeddings of the secondary data sets. In all the figures,
-the cells from the secondary data sets were positioned in the cluster of
-same-typed reference cells, providing strong evidence of the success of our approach. There are some deviations to these observations; for
-instance, in Fig.~\ref{fig:transform_brain} several oligodendrocyte precursor
-cells (OPCs) were mapped to oligodendrocytes. This may be due to differences in
-annotation criteria by different authors, or due to inherent similarities of
-these types of cells. Examples of such erroneous placements can be found in
-other figures as well, but are uncommon and constitute less then 5\% of
-the cells (less than 5\% in brain, 1\% in pancreas and 2\% in retina
-secondary data).
-
-Additional experiments with embedding secondary data sets from independent
-studies are provided in the accompanying Python notebooks. We embedded brain
-cells from Zeisel \etal~\cite{zeisel2015} to the reference embedding shown in
-Fig.~\ref{fig:transform_brain}.a, and pancreas cells from Muraro
-\etal~\cite{muraro2016} to reference in \ref{fig:transform_pancreas}.a. The
-conclusions are largely consistent, with the error rates of X\% and Y\%.
-\todo{Add errors}
-
-Notice that we could simulate the split between reference and secondary data
-sets using one data set only and perform cross-validation, however this type of
-experiment would not incorporate batch effects. We want to remind the reader
-that handling batch effects were central to our endeavour and that the
-disregard of this effect could lead to overly-optimistic results and data
-visualizations strikingly different from ours. For example, compare the
-visualizations from Fig.~\ref{fig:batch_effect}.a and
-Fig.~\ref{fig:transform_brain}.b, or Figs.~\ref{fig:batch_effect}.b and
-\ref{fig:transform_pancreas}.b.
-
-\begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/transform_pancreas.pdf}
-  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
-  cells from the same tissue from Xin \etal.~\cite{xin2016}. Just like in
-  Fig.~\ref{fig:transform_brain} the vast majority of the cells from the
-  secondary data set were correctly mapped to the same-typed cluster of
-  reference cells.}
-  \label{fig:transform_pancreas}
-\end{figure}
-
-\begin{figure}[htb]
-  \includegraphics[width=\textwidth]{figures/transform_retina.pdf}
-  \caption{An embedding of a large reference of retinal cells from Macosco
-  \etal~\cite{macosko2015} (a) and mapping of cells from a smaller study that
-  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} (b). We use
-  colors consistent with the study by Macosko \etal.}
-  \label{fig:transform_retina}
-\end{figure}
-
-We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
-One important extension is the use of multi-scale similarities, which, in addition to finding local structure, also ensure consistent organization of the resulting clusters.
-For illustration, consider visualizations with standard
-and multi-scale t-SNE in Fig.~\ref{fig:multiscale}. Notice, for instance, that
-in multi-scale t-SNE (Fig.~\ref{fig:multiscale}.b) the clusters with neuronal
-cells are clumped together, while their placement in standard t-SNE is
-arbitrary (Fig.~\ref{fig:multiscale}.a).
-
 \begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{figures/hrvatin_multiscale_tsne.pdf}
+  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
   \caption{A comparison of standard and multi-scale t-SNE on data from the
-  mouse visual cortex~\cite{hrvatin2018}. {\bf (a)} Standard t-SNE places
-  clusters arbitrarily. {\bf (b)} Augmenting t-SNE with multi-scale
+  mouse neocortex~\cite{tasic2018}. {\bf (a)} Standard t-SNE using random 
+  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
+  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
+  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
+  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
+  {\bf (c)} Augmenting t-SNE with multi-scale
   similarities and using proper initialization provides a more meaningful
-  layout of the clusters. Neuronal types occupy one region of the space.
-  Oligodendrocyte precursor cells (OPCs) are mainly progenitors to
-  oligodendrocytes, but may also differentiate into neurons or astrocytes.}
+  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
+  the same region of the embedding. There are two clear sub-groups of GABAergic
+  neurons corresponding to their developmental origins, which was not as apparent
+  when using clever initialization alone.}
   \label{fig:multiscale}
@@
 
-We also observed the important role of gene selection in crafting the reference
-embedding spaces. We found that when selecting an insufficient number of genes,
-the resulting visualizations display overly-fragmented clusters. When the
-selection is too broad and includes lowly expressed genes, the subclusters tend
-to overlap. These effects can all be attributed to sparseness of the data sets
-and may be intrinsic to single-cell data. In our studies, we found that
-selection of 3,000 genes yields most informative visualizations
-(Fig.~\ref{fig:gene_selection}).
-
 \begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{figures/hrvatin_embedding_tsne_genes.pdf}
+  \includegraphics[width=\textwidth]{hrvatin_embedding_tsne_genes.pdf}
   \caption{Gene selection plays an important role when constructing the
@@
 
-In principle, our theoretically-grounded embedding of secondary data into the
-scaffold defined by the reference embedding could be simplified with the
-application of the nearest neighbors-based procedure. For example, while
-describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
-proposed positioning new points into a known embedding by placing them in the
-median position of their 10 nearest neighbors, where the neighborhood was
-estimated in the original data space. Notice that we use this approach as well,
-but only for the initialization of positions of new data instances that are
-subject to further optimization. Despite both nearest-neighbors search and
-t-SNE optimization can be computed in linear time, the former dominates the
-runtime (mouse retina example; M reference, N secondary cells, X min N-search,
-Y min optimization). \todo{Add nunmbers}
-
-We demonstrate a case where nearest neighbor-based positioning is insufficient
-in Fig.~\ref{fig:optimization}. It may yield clumped visualizations where the
-optimal positioning using the t-SNE loss function is much more dispersed and
-rightfully shows a more considerable variation in the secondary data. Some
-points may also fall into the empty regions, while after optimization they
-typically move closer to same-typed groups. While the nearest neighbor-based
-approach always places new data points to their most similar cluster, our
-proposed t-SNE approach can identify extreme outliers and place them far from
-the reference cluster.
-
-\begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{figures/optimization_retina.pdf}
-  \caption{Comparison of data placement using the nearest neighbors approach
-  from Kobak and Berens~\cite{art_of_using_tsne} and the optimized placement
-  using our algorithm.  {\bf (a)} Data points are placed to the median position
-  of their 10 nearest neighbors in the reference set. {\bf (b)} Point positions
-  are optimized, revealing a different, more dispersed placement that better
-  reflects the variety of cells in the secondary data set.}
-  \label{fig:optimization}
-\end{figure}
-
-Our approach assumes that all cell types from the secondary data set are
-present in the reference. The method would fail to reveal novel cell
-types in the secondary data set, possibly positioning them arbitrarily close to
-unrelated clusters. Procedures such as scmap were recently proposed to cope
-with such cases and identify the cells whose type is new and not included in
-the reference~\cite{scmap}.
-
-Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
-In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
-Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.
-
-\subsection{Implementation\label{sec:implementation}}
-
-The procedures described in this paper are provided as Python notebooks that
-are, together with the data, available in an open
-repository\footnote{https://github.com/biolab/tsne-embedding}. All experiments
-were run using {\tt openTSNE},
-our open and extensible t-SNE library for Python~\cite{opentsne}.
-
-\section{Conclusion}
-
-Almost all recent publications of single-cell studies begin with a
-two-dimensional visualization of the data that reveals cellular diversity.
-While many dimensionality reduction techniques are available, different 
-variants of t-SNE are most often used to produce such
-visualizations. Single-cell studies enable the exploration of biological mechanisms at a
-cellular level, and their publications in the past couple of years are abundant.
-One of the central tasks in single-cell
-studies is the classification of new cells based on findings from previous
-studies. Such transfer of knowledge is often difficult due to batch effects
-present in data from different sources. Addressing batch effects by adapting
-and extending t-SNE, the prevailing method used to present single-cell data in
-two-dimensional visualization, motivated the research presented in this paper.
-
-Our proposed approach uses a t-SNE embedding as a scaffold for the positioning
-of new cells within the visualization, and possibly for aiding in their
-classification. The three case studies incorporating pairs of data sets from
-different domains but with similar classifications demonstrate that our
-proposed procedure can effectively deal with batch effects to construct
-visualizations that correctly map secondary data sets onto a reference data set
-from an independent study that possibly uses different experimental protocol.
-While we focused here on reference visualizations constructed using t-SNE, this
-approach can be applied using any existing two-dimensional visualization.
-
-\subsubsection*{Acknowledgements}
-
-This work was supported by the Slovenian Research Agency Program Grant P2-0209,
-and by the BioPharm.SI project supported from European Regional Development
-Fund and the Slovenian Ministry of Education, Science and Sport. We would also
-like to thank Dmitry Kobak for many helpful discussions on t-SNE.
-
-% \bibliographystyle{splncs04}
-\bibliographystyle{unsrt}
-\bibliography{references}
 \end{document}

### 2021-05-13 paper-revision/policar2020-tsne-embedding.tex
Add revision files

--- 2021-05-13 paper-revision/diffbase.tex
@@
 \usepackage{booktabs}
-\usepackage{todo}
+\usepackage{hyperref}
 
@@
 \newcommand{\etal}{\textit{et al.}}
+\newcommand{\beginsupplement}{%
+  \setcounter{table}{0}
+  \renewcommand{\thetable}{S\arabic{table}}%
+  \setcounter{figure}{0}
+  \renewcommand{\thefigure}{S\arabic{figure}}%
+}
 
@@
 
-Dimensionality reduction techniques, such as t-SNE, can construct informative visualizations of high-dimensional data.
-When jointly visualising multiple data sets, a straightforward application of these methods often fails; instead of revealing underlying classes, the resulting visualizations expose dataset-specific clusters.
-To circumvent these batch effects, we propose an embedding procedure that uses a t-SNE visualization constructed on a reference data set as a scaffold for embedding new data points.
-Each data instance in the secondary data is embedded independently and does not change the reference embedding.
-This prevents any interactions between instances in the secondary data and implicitly mitigates batch effects.
-We demonstrate the utility of this approach by analyzing six recently published single-cell gene expression data sets with up to tens of thousands of cells and thousands of genes.
-The batch effects in our studies are particularly strong as the data comes from different institutions using different experimental protocols.
-The visualizations constructed by our proposed approach are clear of batch effects, and the cells from secondary data sets correctly co-cluster with cells of the same type from the primary data.
-We also show the predictive power of our visual classification approach in t-SNE space matches the accuracy of machine learning techniques that consider the entire compendium of features that profile single cells.
+Dimensionality reduction techniques, such as t-SNE, can construct informative
+visualizations of high-dimensional data. When jointly visualising multiple data
+sets, a straightforward application of these methods often fails; instead of
+revealing underlying classes, the resulting visualizations expose
+dataset-specific clusters. To circumvent these batch effects, we propose an
+embedding procedure that uses a t-SNE visualization constructed on a reference
+data set as a scaffold for embedding new data points. Each data instance from a
+new, unseen, secondary data is embedded independently and does not change the
+reference embedding. This prevents any interactions between instances in the
+secondary data and implicitly mitigates batch effects. We demonstrate the
+utility of this approach by analyzing six recently published single-cell gene
+expression data sets with up to tens of thousands of cells and thousands of
+genes. The batch effects in our studies are particularly strong as the data
+comes from different institutions using different experimental protocols. The
+visualizations constructed by our proposed approach are clear of batch effects,
+and the cells from secondary data sets correctly co-cluster with cells of the
+same type from the primary data. We also show the predictive power of our
+simple, visual classification approach in t-SNE space matches the accuracy of
+specialized machine learning techniques that consider the entire compendium of
+features that profile single cells.
 
@@
 profiles are similar. For this task, non-linear embedding techniques such as
-t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or
-uniform manifold approximation and projection~\cite{umap} have recently
+t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{Maaten2008} or
+uniform manifold approximation and projection~\cite{McInnes2018} have recently
 complemented traditional data transformation and embedding approaches such as
-principal component analysis (PCA) and multi-dimensional
-scaling~\cite{distill,umap_single_cell}. While useful for visualizing data from
+principal component analysis (PCA)~\cite{wold1987principal} and multi-dimensional
+scaling~\cite{cox2008multidimensional}. While useful for visualizing data from
 a single coherent source, these methods may encounter problems with multiple
@@
 domain-specific differences are more commonly referred to as {\em batch
-effects}~\cite{cca,mnn,seurat}.
+effects}~\cite{Butler2018,Haghverdi2018,Stuart2019}.
 
@@
 analysis. Here, it is expected that cells of the same type would cluster
-together in two-dimensional data visualization~\cite{seurat}. For instance,
+together in two-dimensional data visualization~\cite{Wolf2018}. For instance,
 Fig.~\ref{fig:batch_effect}.a shows t-SNE embedded data from mouse brain cells
-originating from the visual cortex~\cite{hrvatin2018} and the
-hypothalamus~\cite{chen2017}. The figure reveals distinct clusters but also
+originating from the visual cortex~\cite{Hrvatin2018} and the
+hypothalamus~\cite{Chen2017}. The figure reveals distinct clusters but also
 separates the data from the two brain regions. These two regions share the same
@@
 pancreatic cells in Fig.~\ref{fig:batch_effect}.b, which renders the data from
-a pancreatic cell atlas~\cite{baron2016} and similarly-typed cells from
-diabetic patients~\cite{xin2016}. Just like with data from brain cells,
+a pancreatic cell atlas~\cite{Baron2016} and similarly-typed cells from
+diabetic patients~\cite{Xin2016}. Just like with data from brain cells,
 pancreatic cells cluster primarily by data source, again resulting in a
@@
 proposed procedures for batch effect removal include canonical correlation
-analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
-works, batch effects are deemed removed when cells from different sources
-exhibit good mixing in a t-SNE visualization.  The elimination of batch effects
-may require aggressive data preprocessing which may blur the boundaries between
-cell types. Another problem is also the inclusion of any new data, for which
-the entire analysis pipeline must be rerun, usually resulting in a different
-embedding layout and clusters that have little resemblance to original visualization and
+analysis~\cite{Butler2018} and mutual
+nearest-neighbors~\cite{Haghverdi2018,Stuart2019}.  In these works, batch
+effects are deemed removed when cells from different sources exhibit good mixing
+in a t-SNE visualization. The elimination of batch effects may require
+aggressive data preprocessing which may blur the boundaries between cell types.
+Another problem is also the inclusion of any new data, for which the entire
+analysis pipeline must be rerun, usually resulting in a different embedding
+layout and clusters that have little resemblance to original visualization and
 thus require reinterpretation.
@@
 
-The proposed solution implements a mapping of new data into an existing
-t-SNE visualization. While the utility of such an algorithm was already hinted
-at in recent publication~\cite{art_of_using_tsne}, we here provide its
-practical and theoretically-grounded implementation. Considering the abundance
-of recent publications on batch effect removal, we present surprising evidence
-that a computationally more direct and principled embedding procedure solves
-the batch effects problem when constructing interpretable visualizations from
-different data sources.
+The proposed solution implements a mapping of new data into an existing t-SNE
+visualization. While the utility of such an algorithm was already hinted at in
+recent publication~\cite{Kobak2019}, we here provide its practical and
+theoretically-grounded implementation. Considering the abundance of recent
+publications on batch effect removal, we present surprising evidence that a
+computationally more direct and principled embedding procedure solves the batch
+effects problem when constructing interpretable visualizations from different
+data sources.
+
+\section{Related Work}
+
+Batch effects are systematic biases between biological data sets caused by
+technical factors in the data collection and preparation process. It has been
+well documented that even small differences in the experimental setup of
+cell-dissociation, handling protocols, library-preparation technologies, or
+sequencing platforms can significantly affect the resulting gene-expression
+measurements~\cite{Tung2017,Hicks2018}. When performing downstream comparative
+analyses, batch effects may confound real biological variability and may
+introduce spurious correlations, leading to misleading conclusions.
+
+Due to their severity, numerous computational approaches have been proposed to
+directlty remove batch effects when performing joint analysis on two or more data sets.
+Batch effect removal is typically performed as a preprocessing step. Existing
+approaches involve either modifying the original data matrix or finding a joint
+lower-dimensional space, where batch-effects are removed. Current methods
+broadly fall into two categories: 1) mutual nearest neighbor based approaches
+aim to identify matching populations of cells across the data sets, using them
+to either find and correct the data sets~\cite{Haghverdi2018} or directly
+construct a batch-corrected k-nearest neighbor graph used in downstream
+analyses~\cite{Park2018}. 2) embedding multiple data sets into a joint
+lower-dimensional space, where batch effects are removed. Some of these
+approaches opt for linear dimensionality-reduction methods such as
+PCA~\cite{Korsunsky2019} or MultiCCA~\cite{Butler2018}, while others employ
+non-linear techniques from deep learning~\cite{Li2020,Lopez2018} Still, other
+approaches use a combination of the two~\cite{Stuart2019,Hie2019}.
+
+Orthogonal approaches aimed at cell-type attempt to circumvent batch effects
+altogether by identifying a subset of representative genes which act as markers
+for a specific cell-type. Instead of considering the entire gene-expression
+profile, which may be noisy and affected by batch effects, the idea is instead
+to inspect the cells for the presence of a handful of genes, which could
+adequately be used to determine the cell type. One such procedures is
+scMap-Cluster, a consensus-based KNN method, tailored explicitly to scRNA-seq
+gene-expression data~\cite{Kiselev2018}. scMap-Cluster uses three
+correlation-based distance measures and uses a voting scheme to perform
+classification. To identify novel cell types, scMap-Cluster heuristically
+determines a distance threshold.
+
+Broadly speaking, our approach falls into the second category, alongside
+scMap-Cluster. The approach we describe uses standard single-cell pipelines
+based on identifying representative genes. However, we emphasize that the
+primary purpose of our approach is not classification, but the visualization of
+the various cell-types. We can apply a KNN classifier to the resulting
+visualizations to obtain accuracy estimates and compare our approach to other
+classification methods. However, the classification aspect of our approach is
+secondary: the primary purpose of t-SNE is to aid in scientists in exploratory
+data analysis and help them better understand the underlying data landscape.
 
@@
 Eq.~\ref{eq:cauchy_kernel} is $\mathcal{O}(N^2)$, making its application
-impractical for large data sets. We adopt a recent approach
-for low-rank approximation of gradients based on polynomial interpolation which
-reduces its time complexity to $\mathcal{O}(N)$. This approximation
-enables the visualization of massive data sets, possibly containing millions of
-data points~\cite{fi_tsne}.
+impractical for large data sets. We adopt a recent approach for low-rank
+approximation of gradients based on polynomial interpolation which reduces its
+time complexity to $\mathcal{O}(N)$. This approximation enables the
+visualization of massive data sets, possibly containing millions of data
+points~\cite{Linderman2019}.
 
@@
 long-range interactions between clusters. Larger values may result in a more
-globally consistent visualizations --- preserving distances on a large scale
-and organizing clusters in a more meaningful way --- but can lead to merging
-small clusters and thus obscuring local aspects of the
-data~\cite{art_of_using_tsne}.
+globally consistent visualizations --- preserving distances on a large scale and
+organizing clusters in a more meaningful way --- but can lead to merging small
+clusters and thus obscuring local aspects of the data~\cite{Kobak2019}.
 
@@
 achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
-a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}.
-Multi-scale kernels are defined as
+a mixture of Gaussians of varying bandwidths~\cite{Lee2015}. Multi-scale kernels
+are defined as
 
@@
 $l$. In our experiments, we used a mixture of two Gaussian kernels with
-perplexity values of 50 and 500. A similar formulation of
-multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
-resulting embeddings are visually very similar to those obtained with the
-approach described above (not shown for brevity).
+perplexity values of 50 and 500. A similar formulation of multi-scale kernels
+was proposed in~\cite{Kobak2019}, and we found the resulting embeddings are
+visually very similar to those obtained with the approach described above (not
+shown for brevity).
 
@@
 has been shown to lead to slower convergence and requires more iterations to
-achieve consistent embeddings~\cite{belkina2019automated}. We follow the
-recommendation of Belkina \etal ~and use a higher learning rate $\eta = N / 12$
-when visualizing larger data sets.
+achieve consistent embeddings~\cite{Belkina2019}. We follow the recommendation
+of Belkina \etal ~and use a higher learning rate $\eta = N / 12$ when
+visualizing larger data sets.
 
@@
 
-The position of a new data point in embedding space is initialized to the
-median reference embedding position of its $k$ nearest neighbors. While we
-found the algorithm to be robust to choices of $k$, we use $k=10$ in our
-experiments.
+The position of a new data point in embedding space is initialized to the median
+reference embedding position of its $k$ nearest neighbors. While we found the
+algorithm to be robust to choices of $k$, we use $k=10$ in our experiments.
 
@@
 descent. We use an adaptive learning rate scheme with momentum to speed up the
-convergence, as proposed by Jacobs~\cite{momentum,bh_tsne}. We run gradient
+convergence, as proposed by Jacobs~\cite{Jacobs1988,Maaten2014}. We run gradient
 descent with momentum $\alpha$ of $0.8$ for $250$ iterations, where the
@@
 evaluate the gradients in Eq.~\ref{eq:gradient} is $\mathcal{O}(N \cdot M)$,
-however, by adapting the same polynomial interpolation based approximation,
-this is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
-reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding remains fixed.
+however, by adapting the same polynomial interpolation based approximation, this
+is reduced to $\mathcal{O}(\max \{ N, M \})$. The time complexity can further be
+reduced to $\mathcal{O}(M)$ by exploiting the fact that the reference embedding
+remains fixed.
 
@@
 $\eta \sim 0.1$ produces stable solutions. Alternatively, we can employ gradient
-clipping to achieve similar behaviour. This is especially important when
-using the interpolation-based approximation, which places a grid of
-interpolation points over the embedding space, where the number of grid points
-is determined by the span of the embedding. Clearly, if even one point ``shoots
-off'' far from the embedding, the number of required grid points may grow
-dramatically, increasing the runtime substantially. The reduced learning rate
-suppresses this issue, and does not slow the convergence because of the
-adaptive learning rate scheme, provided the optimization is run for a
-sufficient number of steps.
+clipping to achieve similar behaviour. This is especially important when using
+the interpolation-based approximation, which places a grid of interpolation
+points over the embedding space, where the number of grid points is determined
+by the span of the embedding. Clearly, if even one point ``shoots off'' far from
+the embedding, the number of required grid points may grow dramatically,
+increasing the runtime substantially. The reduced learning rate suppresses this
+issue, and does not slow the convergence because of the adaptive learning rate
+scheme, provided the optimization is run for a sufficient number of steps.
 
@@
 
-We apply the proposed approach to t-SNE visualizations of single-cell data.
-Data in this realm include a variety of cells from specific tissues and are
+We apply the proposed approach to t-SNE visualizations of single-cell data. Data
+in this realm include a variety of cells from specific tissues and are
 characterized through gene expression.  In our experiments, we considered
@@
 that end, we focus on comparing different ways of using t-SNE rather than
-differences to embeddings like PCA or MDS, which have been substantially
-covered before~\cite{tsne,umap_single_cell}.  Below, we list the data sets,
-describe single-cell specific data preprocessing procedures, and display the
-resulting data visualizations. Finally, we discuss the success of the proposed
-approach in alleviating the batch effects.
+differences to embeddings like PCA or MDS, which have been substantially covered
+before~\cite{Maaten2008,Becht2019}. Below, we list the data sets used in our
+experiments, and display the resulting data visualizations. Due to the unique
+nature of single-cell data, we apply a specialized single-cell pipeline for all
+our experiments, as described in Appendix~\ref{sec:sc-pipeline}. Finally, we
+discuss the success of the proposed approach in alleviating the batch effects.
 
@@
 the majority of cell types from the secondary data set were included in the
-reference set (Table~\ref{tab:data sets}). The cells in the data sets
-originate from the following three tissues:
+reference set (Table~\ref{tab:data sets}). The cells in the data sets originate
+from the following three tissues:
 
 \begin{description}
-\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
+\item[Mouse brain.] The data set from Hrvatin \etal~\cite{Hrvatin2018} contains
 cells from the visual cortex exploring transcriptional changes after exposure
 to light. This was used as a reference for the data from Chen
-\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
+\etal~\cite{Chen2017}, containing cells from the mouse hypothalamus and
 their reaction to food deprivation. From the secondary data, we removed cells
@@
 
-\item[Human pancreas.] Baron \etal~\cite{baron2016} created
+\item[Human pancreas.] Baron \etal~\cite{Baron2016} created
 an atlas of pancreatic cell types. We used this set as a reference
-for data from Xin \etal~\cite{xin2016}, who examined transcriptional
+for data from Xin \etal~\cite{Xin2016}, who examined transcriptional
 differences between healthy and type 2 diabetic patients.
 
-\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
+\item[Mouse retina.] Macosko \etal~\cite{Macosko2015} created
 an atlas of mouse retinal cell types. We used this as a reference
-for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
+for the data from Shekhar \etal~\cite{Shekhar2016}, who built an atlas for
 retinal bipolar cells.
@@
 \caption{Data sets used in our experiments. The first data set in each pair
-(Hrvatin \etal, Baron \etal, and Macoscko \etal) was used as a reference.  We
+(Hrvatin \etal, Baron \etal, and Macosko \etal) was used as a reference.  We
 relied on the quality control and annotations from the original publication
@@
 annotations were made consistent to annotations from the Cell
-Ontology~\cite{cell_ontology}. Notice that different RNA sequencing protocols
+Ontology~\cite{Bard2005}. Notice that different RNA sequencing protocols
 were used to estimate gene expressions.}
@@
 
-\subsection{Results and Discussion}
+\subsection{t-SNE transform successfully alleviates batch effects}
 \label{sec:results}
@@
   \includegraphics[width=\textwidth]{transform_pancreas.pdf}
-  \caption{Embedding of pancreatic cells from Baron \etal~\cite{baron2016} and
-  cells from the same tissue from Xin \etal~\cite{xin2016}. Just like in
+  \caption{Embedding of pancreatic cells from Baron \etal~\cite{Baron2016} and
+  cells from the same tissue from Xin \etal~\cite{Xin2016}. Just like in
   Fig.~\ref{fig:transform_brain}, the vast majority of the cells from the
@@
   \caption{An embedding of a large reference of retinal cells from Macosco
-  \etal~\cite{macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
-  focuses on bipolar cells from Shekhar \etal~\cite{shekhar2016} \textbf{(b)}. We use
+  \etal~\cite{Macosko2015} \textbf{(a)} and mapping of cells from a smaller study that
+  focuses on bipolar cells from Shekhar \etal~\cite{Shekhar2016} \textbf{(b)}. We use
   colors consistent with the study by Macosko \etal}
@@
 
+\subsection{Constructing a reference embedding}
+
 We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
 Kobak and Linderman have shown that the global consistency of embeddings produced by
-popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
+popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
 more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
@@
 clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
-developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
+developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
 is less apparent when using PCA-based initialization alone in
@@
 
+\subsection{Optimization is crucial to producing meaningful point embeddings}
+
 In principle, our theoretically-grounded embedding of secondary data into the
@@
 application of the nearest neighbors-based procedure. For example, while
-describing a set of tricks for t-SNE, Kobak and Berens~\cite{art_of_using_tsne}
+describing a set of tricks for t-SNE, Kobak and Berens~\cite{Kobak2019}
 proposed positioning new points into a known embedding by placing them in the
@@
 positioning alone is insufficient. We construct a reference embedding using only
-neurons from Hrvatin \etal~\cite{hrvatin2018} (Fig.~\ref{fig:optimization}.a)
+neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a)
 and use that to position neuronal
-cells from the data set from Campbell \etal~\cite{campbell2017}. We utilize the
+cells from the data set from Campbell \etal~\cite{Campbell2017}. We utilize the
 weighted mean and median positions to initialize point positions from the secondary
@@
   points onto reference embeddings. {\bf (a)} We construct a reference embedding
-  using only neuronal subtypes from Hrvatin \etal~\cite{hrvatin2018}. {\bf (b)}
-  We position neuronal cells from Campbell \etal~\cite{campbell2017} using the
-  median initizaliation scheme from Kobak and Berens~\cite{art_of_using_tsne} and
+  using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)}
+  We position neuronal cells from Campbell \etal~\cite{Campbell2017} using the
+  median initizaliation scheme from Kobak and Berens~\cite{Kobak2019} and
   run optimization for 500 iterations. Compare the optimized embedding with the 
@@
 
-The approach assumes that all cell types from the secondary data set are
-present in the reference. The method may fail to reveal novel cell
-types in the secondary data set, possibly positioning them arbitrarily close to
-unrelated clusters. Procedures such as scmap were recently proposed to cope
-with such cases and identify the cells whose type is new and not included in
-the reference~\cite{scmap}.
+\subsection{Our approach requires a complete reference set}
 
-Our procedure is, therefore, asymmetrical in the choice of reference and secondary data set.
-In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
-Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.
+Our approach assumes that all cell types from the secondary data set are present
+in the reference. Intuitively, using t-SNE in such a way is conceptually similar
+to classification via $k$-nearest neighbor classifiers and is similarily
+limited. The method may fail to reveal unseen cell types in the secondary data
+set, likely positioning them arbitrarily close to unrelated clusters. In some
+instances, unknown cell types may be sufficiently different from the reference
+data that t-SNE will repel them from existing clusters. However, we caution that
+this approach is unreliable and depends heavily on the chosen preprocessing
+pipeline.
 
-To quantitatively evaluate the predictive accuracy of the described procedure, we fit $k$-nearest neighbors classifiers on each reference t-SNE embedding from Figs.~\ref{fig:transform_brain}.a, \ref{fig:transform_pancreas}.a and \ref{fig:transform_retina}.a and use them to predict the cell types for the secondary data set embeddings from Figs.~\ref{fig:transform_brain}.b, \ref{fig:transform_pancreas}.b and \ref{fig:transform_retina}.b.
-The accuracy measures are reported in Table~\ref{tab:scores}.
-Our procedure of embedding new data points into two-dimensional t-SNE plane results in similar accuracy to approaches like random forests that use full compendium of cell-characterizing features.
-The results indicate that positioning of new cells onto a cell visualisation plane is not only indicative but also an accurate instrument for cell type characterization.
+We illustrate this with Fig.~\ref{fig:incomplete_reference}, where we first fit
+create a reference embedding containing only neuronal cells from Hrvatin
+\etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
+\etal~\cite{Campbell2017} and add them to the reference embedding in
+Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal
+are scattered somewhat arbitrarily around several clusters in the reference
+embedding. Interestingly, the secondary data points form a ``ring'' around one
+of the clusters, indicating that these data points are very different from the
+cells in this cluster. Notice also that the points from the secondary data set
+exhibit little to no clustering and the different cell types seem to be mixed
+amongst each other. We hypothesize that this effect is due primarily to the
+single-cell preprocessing pipeline and not the limitations of our proceedure
+itself, as the informative genes selected to create the reference neuronal
+embedding likely do not differentiate supportive glial cells from the secondary
+data set. This effect is similar to procedures such as scMap-Cluster, a consensus KNN
+method, which heuristically determines a distance threshold to identify unknown
+cell types~\cite{Kiselev2018}.
+
+\begin{figure}[htb]
+  \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
+  \caption{A reference embedding must contain all the cell-types in the
+  secondary embedding to produce reliable results. {\bf (a)} We construct a
+  reference embedding containing only neuronal cells from Hrvatin
+  \etal~\cite{Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from
+  Campbell \etal~\cite{Campbell2017} so that no overlap exists between the
+  cell-types between the data sets. Conceptually, t-SNE behaves similarily to a
+  KNN classifier, and places the non-neurnal cells to their the most similar
+  points in the reference. In some instances, the non-neuronal cells are
+  sufficiently different from the neuronal cells so that they are repelled from
+  the reference clusters. Such behaviour results in the ``ring'' seen on the
+  right hand side of the embedding.} \label{fig:incomplete_reference}
+\end{figure}
+
+Our procedure is, therefore, asymmetrical in the choice of reference and
+secondary data set. In practice, however, newly produced {\em secondary} data
+would be embedded into previously-prepared reference landscapes. Large
+collections of data {\em e.g.} the Human Cell Atlas
+initiative~\cite{Rozenblatt-Rosen2017} make it possible to scale up our approach
+to wider sets of cell types. Identifying potential failure cases where rare
+cell-types may still be missing from constructed reference embeddings is a
+problem that plagues the bioinformatics community and is an active area of
+research.
+
+\subsection{Comparison to other similar batch-effect methods}
+
+To quantitatively evaluate the predictive accuracy of the described procedure,
+we fit $k$-nearest neighbors classifiers on each reference t-SNE embedding from
+Figs.~\ref{fig:transform_brain}.a, \ref{fig:transform_pancreas}.a and
+\ref{fig:transform_retina}.a and use them to predict the cell types for the
+secondary data set embeddings from Figs.~\ref{fig:transform_brain}.b,
+\ref{fig:transform_pancreas}.b and \ref{fig:transform_retina}.b. The accuracy
+measures are reported in Table~\ref{tab:scores}. Our procedure of embedding new
+data points into two-dimensional t-SNE plane results in similar accuracy to
+approaches like random forests that use full compendium of cell-characterizing
+features. The results indicate that positioning of new cells onto a cell
+visualisation plane is not only indicative but also an accurate instrument for
+cell type characterization.
 
@@
   \end{tabular}
-  \caption{We compare our approach (t-SNE) to three other methods, evaluating performance using classification accuracy and the adjusted rand index (ARI).
-Notice that while the proposed approach classifies cells only based on their position on the two-dimensional plane, it performs comparably to other methods that use full compendium of features (gene expressions) that characterize the cells.}
+  \caption{We compare our approach (t-SNE) to three other methods, evaluating
+    performance using classification accuracy and the adjusted rand index (ARI).
+    Notice that while the proposed approach classifies cells only based on their
+    position on the two-dimensional plane, it performs comparably to other
+    methods that use full compendium of features (gene expressions) that
+    characterize the cells.}
   \label{tab:scores}
@@
 
-We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{scmap}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data.
-scMap-Cluster uses three correlation-based distance measures and uses a voting scheme to perform classification.
-A threshold may also be set so that new cell types may be labeled as {\tt unrecognized}, but since our secondary data sets were chosen such that this does not happen, this was disabled.
-For the two machine learning approaches, we apply the typical single-cell preprocessing pipeline described in Sec.~\ref{sec:sc-pipeline}, i.e., library-size normalization, log-transformation, and select 1,000 most informative genes.
-Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest neighbors in the KNN model.
-We used 100 trees in the random forest ensemble.
-The models were fit on the reference data set, and no hyper-parameter tuning was performed.
+We compare our approach to two machine learning techniques, namely a $k$-nearest
+neighbor classifier (KNN) and a random forest ensemble, and
+scMap-Cluster~\cite{Kiselev2018}. For scMap-Cluster, we disable the distance
+threshold heuristic for identifying novel cell types, as our secondary data sets
+were chosen such that there is complete overlap between cell-types. For the two
+machine learning approaches, we apply the typical single-cell preprocessing
+pipeline described in Appendix~\ref{sec:sc-pipeline}, i.e., library-size
+normalization, log-transformation, and select 1,000 most informative genes.
+Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest
+neighbors in the KNN model. We used 100 trees in the random forest ensemble. The
+models were fit on the reference data set, and no hyper-parameter tuning was
+performed.
 
-Surprisingly, both the random forest and $k$-nearest neighbor models outperform scMap-Cluster, which is specifically tailored to scRNA-seq data.
-However, these results may be skewed, as, in our examples, all the cell-types from the secondary data set were present in the reference data set.
-One of the core features of scMap-Cluster is the detection of novel cell types, which none of the other methods support.
-In other words, the other three methods would always assign a cell-type to a given cell, regardless of cell origin.
-Additionally, scMap-Cluster was primarily designed and tested on data sets produced by full-length sequencing protocols, which tend to detect a much higher number of molecules than other, sequencing protocols based on unique molecular identifiers (UMI).
-These two classes of sequencing protocols produce data sets with different sparsity and variance characteristics.
-This is consistent with the results in Table~\ref{tab:scores}, as only the data sets from the human pancreas, were produced using a full-length sequencing protocol, where scMap-Cluster achieves reasonably high accuracy.
+Surprisingly, both the random forest and $k$-nearest neighbor models outperform
+scMap-Cluster, which is specifically tailored to scRNA-seq data. However, these
+results may be skewed, as, in our examples, all the cell-types from the
+secondary data set were present in the reference data set. One of the core
+features of scMap-Cluster is the detection of novel cell types, which none of
+the other methods support. In other words, the other three methods would always
+assign a cell-type to a given cell, regardless of cell origin. Additionally,
+scMap-Cluster was primarily designed and tested on data sets produced by
+full-length sequencing protocols, which tend to detect a much higher number of
+molecules than other, sequencing protocols based on unique molecular identifiers
+(UMI). These two classes of sequencing protocols produce data sets with
+different sparsity and variance characteristics. This is consistent with the
+results in Table~\ref{tab:scores}, as only the data sets from the human
+pancreas, were produced using a full-length sequencing protocol, where
+scMap-Cluster achieves reasonably high accuracy.
 
-The aim of t-SNE is to construct embeddings, in which neighborhoods are preserved, therefore it is unsurprising that the accuracy of our t-SNE based approach is largely consistent with the $k$-nearest neighbors model.
-Our approach is comparable to the other models in terms of accuracy, but we would note that the goal of t-SNE embeddings is to serve as visual aids in exploratory data analysis.
-However, our procedure, in addition to providing the end-user with a cell-type prediction, allows the user to examine the low-dimensional embedding space, which may provide richer insight and interpretation of the resulting predictions.
+The aim of t-SNE is to construct embeddings, in which neighborhoods are
+preserved, therefore it is unsurprising that the accuracy of our t-SNE based
+approach is largely consistent with the $k$-nearest neighbors model. While our
+approach is comparable to the other models in terms of accuracy, we emphasize
+that the goal of t-SNE embeddings is to serve as visual aids in exploratory data
+analysis. Therefore, it is surprising that our simple procedure performs
+competitievly to specialized classification methods. Therefore, our procedure,
+in addition to providing the end-user with a cell-type prediction, allows the
+user to examine the low-dimensional embedding space, which may provide richer
+insight and interpretation of the resulting predictions.
 
@@
 Almost all recent publications of single-cell studies begin with a
-two-dimensional visualization of the data that reveals cellular diversity.
-While many dimensionality reduction techniques are available, different 
-variants of t-SNE are most often used to produce such
-visualizations. Single-cell studies enable the exploration of biological mechanisms at a
-cellular level, and their publications in the past couple of years are abundant.
-One of the central tasks in single-cell
-studies is the classification of new cells based on findings from previous
-studies. Such transfer of knowledge is often difficult due to batch effects
-present in data from different sources. Addressing batch effects by adapting
-and extending t-SNE, the prevailing method used to present single-cell data in
-two-dimensional visualization, motivated the research presented in this paper.
+two-dimensional visualization of the data that reveals cellular diversity. While
+many dimensionality reduction techniques are available, different  variants of
+t-SNE are most often used to produce such visualizations. Single-cell studies
+enable the exploration of biological mechanisms at a cellular level, and their
+publications in the past couple of years are abundant. One of the central tasks
+in single-cell studies is the classification of new cells based on findings from
+previous studies. Such transfer of knowledge is often difficult due to batch
+effects present in data from different sources. Addressing batch effects by
+adapting and extending t-SNE, the prevailing method used to present single-cell
+data in two-dimensional visualization, motivated the research presented in this
+paper.
 
-The proposed approach uses a t-SNE embedding as a scaffold for the positioning of new cells within the visualization, and possibly for aiding in their classification.
-The three case studies incorporating pairs of data sets from different domains but with similar classifications demonstrate that our proposed procedure can effectively deal with batch effects to construct visualizations that correctly map secondary data sets onto an embedding of the data from an independent study that possibly uses different experimental protocol.
-We quantitatively evaluate the predictive accuracy of our approach by fitting a $k$-nearest neighbors model on the resulting two-dimensional embeddings and compare its predictive accuracy to other machine learning methods that use the entire compendium of gene expressions that characterize the cells.
-Experiments show that our approach is successful in predicting cell types and performs comparably to other methods.
-This encouraging result indicates that by using our procedure, scientists can quickly and accurately determine the composition of new data by merely visualizing and inspecting resulting visualizations.
-While we focused here on reference visualizations constructed using t-SNE, this approach can be applied using any existing two-dimensional visualization.
+The proposed approach uses a t-SNE embedding as a scaffold for the positioning
+of new cells within the visualization, and possibly for aiding in their
+classification. The three case studies incorporating pairs of data sets from
+different domains but with similar classifications demonstrate that our proposed
+procedure can effectively deal with batch effects to construct visualizations
+that correctly map secondary data sets onto an embedding of the data from an
+independent study that possibly uses different experimental protocol. We
+quantitatively evaluate the predictive accuracy of our approach by fitting a
+$k$-nearest neighbors model on the resulting two-dimensional embeddings and
+compare its predictive accuracy to other machine learning methods that use the
+entire compendium of gene expressions that characterize the cells. Experiments
+show that our approach is successful in predicting cell types and performs
+comparably to other methods. This encouraging result indicates that by using our
+procedure, scientists can quickly and accurately determine the composition of
+new data by merely visualizing and inspecting resulting visualizations. While we
+focused here on reference visualizations constructed using t-SNE, this approach
+can be applied using any existing two-dimensional visualization.
 
-\subsubsection*{Implementation\label{sec:implementation}}
+\subsubsection*{Availability and Implementation\label{sec:implementation}}
 
@@
 are, together with the data, available in an open
-repository\footnote{https://github.com/biolab/tsne-embedding}. All experiments
-were run using {\tt openTSNE},
-our open and extensible t-SNE library for Python~\cite{opentsne}.
+repository\footnote{https://github.com/biolab/tsne-embedding}. The described
+methods were implemented and incorporated into \textsf{openTSNE}, our
+open-source, extensible t-SNE library for Python~\cite{Policar2019}.
 
@@
 
+\newpage
 \appendix
+\beginsupplement
 
@@
 consisting of the selection of 3,000 representative genes (see
-Sec.~\ref{sec:gene-selection}), library size normalization, log-transformation,
-standardization, and PCA-based representation that retains 50 principal
-components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
-multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
-high-dimensionality of the preprocessed input data we use cosine distance to
-estimate similarities between reference data points~\cite{Domingos2012-CACM}.
-When adding new data points from the secondary data set to the reference
-embedding, we select 1,000 genes present in both data sets and use the cosine
-similarity to estimate the similarities between the secondary data item and
-reference data points. We note that similarities are computed using the raw
-count matrices. The preprocessing stages are detailed in accompanying Python
-notebooks (Sec.~\ref{sec:implementation}).
+Appendix~\ref{sec:gene-selection}), library size normalization,
+log-transformation, standardization, and PCA-based representation that retains
+50 principal components~\cite{Stuart2019,Wolf2018}. To obtain the reference
+embedding, we apply multi-scale t-SNE using PCA initialization~\cite{Kobak2019}.
+Due to high-dimensionality of the preprocessed input data we use cosine distance
+to estimate similarities between reference data points~\cite{Domingos2012}. When
+adding new data points from the secondary data set to the reference embedding,
+we select 1,000 genes present in both data sets and use the cosine similarity to
+estimate the similarities between the secondary data item and reference data
+points. We note that similarities are computed using the raw count matrices. The
+preprocessing stages are detailed in accompanying Python notebooks
+(Sec.~\ref{sec:implementation}).
 
@@
 
-% TODO: This section (3.3) is not clear
+Single-cell data sets suffer from high levels of technical noise and low capture
+efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A
+common occurence in these data sets is ``dropout'', where an expressed gene is
+not measured, and its corresponding matrix entry is set to zero. To address this
+problem, we use a specialized feature-selection method, which exploits the
+mean-dropout relationship of expression counts as recently proposed by Kobak and
+Berens~\cite{Kobak2019}. In this context, we will refer to all genes with matrix
+entries set to zero as dropouts. Intuitively, if a gene has high mean expression
+over all cells, but is detected in only a handful of them (i.e. has high mean
+dropout rate), then this gene is likely specific to a specific cell-type and
+will serve as a good feature for any subsequent analysis where we wish to
+discriminate between cell types.
 
-Single-cell data sets suffer from high levels of technical noise and low
-capture efficiency, resulting in sparse expression matrices~\cite{umi}. To
-address this problem, we use a specialized feature-selection method, which
-exploits the mean-dropout relationship of expression counts as recently
-proposed by Kobak and Berens~\cite{art_of_using_tsne}. Here, genes with higher
-than expected dropout rate are regarded as potential markers for cell
-subpopulations and are retained in the data.
-
-Given an expression matrix $\mathbf{X} \in \mathbb{R}^{N \times G}$ where $N$
-is the number of samples and $G$ is the number of genes in the data set, we
-compute the fraction of cells where a gene $g$ was not expressed
-
+More formally, given an expression matrix $\mathbf{X} \in \mathbb{R}^{N \times
+G}$ where $N$ is the number of cells and $G$ is the number of genes in the data
+set, we compute the fraction of cells where a gene $g$ was not expressed i.e.
+its dropout rate
 \begin{equation}
-d_g = \frac{1}{N} \sum_i I \left ( X_{ig} = 0\right )
+d_g = \frac{1}{N} \sum_i I \left ( X_{ig} = 0\right ),
 \end{equation}
-
-\noindent The mean $\log_2$ expression of gene $g$ considers only cells
-expressing $g$:
-
+where $I$ is the indicator function. The mean $\log_2$ expression of gene $g$
+considers only cells $i$ in which gene $g$ was expressed
 \begin{equation}
@@
 All genes expressed in less than ten cells are discarded. In order to select a
-specific number of $\hat{G}$ genes, we use a binary search to find a value $b$
-such that
-
+desired number of $\hat{G}$ genes, we use binary search to find a parameter
+value of $b$ such that
 \begin{equation}
@@
 \end{equation}
+In this way, we are able to select a desired number of genes $\hat{G}$, which
+appear discriminative between cell types for the given gene-expression data set.
+
+\newpage
 
@@
   \caption{A comparison of standard and multi-scale t-SNE on data from the
-  mouse neocortex~\cite{tasic2018}. {\bf (a)} Standard t-SNE using random 
+  mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random 
   initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,

### 2021-05-13 paper-revision/response-to-reviewers.tex
Add paragraph listing contributions

@@
 
-\answer{}
+\answer{
+	We added a paragraph to the end of the introduction explicitly listing our contributions.
+The differences from the conference version are mainly an improved experimental section, where we show that a naive placement of the data points leads to misleading conclusions, and that optimization produces more reasonable results.
+Additionally, we provide the classification evaluation where we compare the classification accuracy of our proposed approach and standard machine learning methods for scRNA-seq data.
+}
 

### 2021-05-13 paper-revision/policar2020-tsne-embedding.tex
Add paragraph listing contributions

@@
 data sources.
+
+Our contributions are two-fold. 
+\begin{enumerate}
+  \item We introduce a theoretically-grounded extension of the t-SNE
+  visualization algorithm, which allows the embedding of new data points into
+  existing, reference visualizations. Our extension is easily incorporated into
+  existing approximation schemes, enabling it to be applied to large data sets.
+  We show that optimization using the default t-SNE parameters is highly
+  unstable and propose parameter values which lead to stable convergence.
+  \item We show that our straightforward extension can be used to mitigate batch
+  effects in single-cell RNA-seq data.
+\end{enumerate}
 

### 2021-06-18 paper-revision/policar2020-tsne-embedding.tex
revision

--- 2021-05-13 paper-revision/policar2020-tsne-embedding.tex
@@
 
+Our contributions are two-fold: 
+\begin{enumerate}
+  \item We introduce a theoretically-grounded extension of the t-SNE
+   visualization algorithm that supports embedding new data points into
+   existing reference visualizations. Our extension is readily incorporated
+   into existing approximation schemes, enabling its applications to large
+   data sets. We show that optimization using the default t-SNE parameters is
+   highly unstable and proposes parameter values leading to stable
+   convergence.
+
+  \item We show that the proposed t-SNE extensions can mitigate batch effects
+   in the data sets and demonstrate this feature in treating single-cell gene
+   expression data.
+\end{enumerate}
+
 \section{Related Work}
@@
 sequencing platforms can significantly affect the resulting gene-expression
-measurements~\cite{Tung2017,Hicks2018}. When performing downstream comparative
-analyses, batch effects may confound real biological variability and may
-introduce spurious correlations, leading to misleading conclusions.
+measurements~\cite{Tung2017,Hicks2018}. When performing downstream
+comparative analyses, batch effects may confound real biological variability
+and introduce spurious correlations, leading to misleading conclusions.
 
 Due to their severity, numerous computational approaches have been proposed to
-directlty remove batch effects when performing joint analysis on two or more data sets.
-Batch effect removal is typically performed as a preprocessing step. Existing
-approaches involve either modifying the original data matrix or finding a joint
-lower-dimensional space, where batch-effects are removed. Current methods
-broadly fall into two categories: 1) mutual nearest neighbor based approaches
-aim to identify matching populations of cells across the data sets, using them
-to either find and correct the data sets~\cite{Haghverdi2018} or directly
-construct a batch-corrected k-nearest neighbor graph used in downstream
-analyses~\cite{Park2018}. 2) embedding multiple data sets into a joint
-lower-dimensional space, where batch effects are removed. Some of these
-approaches opt for linear dimensionality-reduction methods such as
-PCA~\cite{Korsunsky2019} or MultiCCA~\cite{Butler2018}, while others employ
-non-linear techniques from deep learning~\cite{Li2020,Lopez2018} Still, other
-approaches use a combination of the two~\cite{Stuart2019,Hie2019}.
+directly remove batch effects when performing joint analysis on two or more
+data sets. Batch effect removal is typically performed as a preprocessing
+step. Existing approaches involve either modifying the original data matrix
+or finding a joint lower-dimensional space, where batch effects are removed.
+Current methods broadly fall into two categories:
+\begin{enumerate}
+\item mutual nearest neighbor-based approaches aim to identify matching
+ populations of cells across the data sets, using them to either find and
+ correct the data sets~\cite{Haghverdi2018} or directly construct a
+ batch-corrected k-nearest neighbor graph used in downstream analyses~\cite
+ {Park2018}.
+\item embedding multiple data sets into a joint lower-dimensional space, where
+ batch effects are removed. Some of these approaches opt for linear
+ dimensionality-reduction methods such as PCA~\cite{Korsunsky2019} or
+ MultiCCA~\cite{Butler2018}, while others employ non-linear techniques from
+ deep learning~\cite{Li2020,Lopez2018}. Still, other approaches use a
+ combination of the two~\cite{Stuart2019,Hie2019}.
+\end{enumerate}
 
-Orthogonal approaches aimed at cell-type attempt to circumvent batch effects
-altogether by identifying a subset of representative genes which act as markers
-for a specific cell-type. Instead of considering the entire gene-expression
-profile, which may be noisy and affected by batch effects, the idea is instead
-to inspect the cells for the presence of a handful of genes, which could
-adequately be used to determine the cell type. One such procedures is
-scMap-Cluster, a consensus-based KNN method, tailored explicitly to scRNA-seq
+Besides computational techniques, approaches for the removal of batch effects
+can also use domain knowledge. For example, in the analysis of single-cell
+gene expression data, these approaches act on a subset of representative
+marker genes for a specific cell type. Instead of considering the entire
+gene-expression profile, which may be noisy and affected by batch effects,
+the idea is to profile the cells with a handful of genes that can
+collectively determine the cell type. One such procedure is scMap-Cluster, a
+consensus-based $k$-nearest neighbor method tailored explicitly to scRNA-seq
 gene-expression data~\cite{Kiselev2018}. scMap-Cluster uses three
@@
 
-Broadly speaking, our approach falls into the second category, alongside
-scMap-Cluster. The approach we describe uses standard single-cell pipelines
-based on identifying representative genes. However, we emphasize that the
-primary purpose of our approach is not classification, but the visualization of
-the various cell-types. We can apply a KNN classifier to the resulting
+Our approach to batch effect removal falls into the second category, as we
+lose the batch effects through dimensionality reduction. Alongside
+scMap-Cluster, we also benefit from a standard single-cell data preprocessing
+pipeline that profiles the cells with representative genes. Unlike other
+batch effect removal procedures, the primary purpose of our approach is not
+classification but the visualization of the various cell-types. If required,
+we can apply a $k$-nearest neighbor classifier to the resulting
 visualizations to obtain accuracy estimates and compare our approach to other
 classification methods. However, the classification aspect of our approach is
-secondary: the primary purpose of t-SNE is to aid in scientists in exploratory
-data analysis and help them better understand the underlying data landscape.
+secondary: the primary purpose of t-SNE is to aid in scientists in
+exploratory data analysis and help them better understand the underlying data
+landscape.
 
@@
 relied on the quality control and annotations from the original publication
-and report the number of cell types after preprocessinng.  The cell
+and report the number of cell types after prepossessing.  The cell
 annotations were made consistent to annotations from the Cell
@@
 experiment would not incorporate batch effects. We want to remind the reader
-that handling batch effects were central to our endeavour and that the
+that handling batch effects were central to our endeavor and that the
 disregard of this effect could lead to overly-optimistic results and data
@@
 
-\subsection{Constructing a reference embedding}
+\subsection{Construction of a reference embedding}
 
-We use a number of additional, recently proposed modifications to enhance the t-SNE visualization of the reference data set.
-Kobak and Linderman have shown that the global consistency of embeddings produced by
-popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
-more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
-as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
-Another important extension is the use of multi-scale similarities, which, in addition
-to considering short range interactions, also models wider point neighborhoods. Coupled
-with PCA-based initializations, this produces even more meaningful visualizations where
-clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
-developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
-is less apparent when using PCA-based initialization alone in
-Fig.~\ref{fig:multiscale}.b.
+We use a number of additional, recently proposed modifications to enhance the
+t-SNE visualization of the reference data set. Kobak and Linderman have shown
+that the global consistency of embeddings produced by popular visualization
+algorithms are largely dependent on their initialization~\cite
+{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
+more meaningful layouts of the resulting clusters (Fig.~\ref
+{fig:multiscale}.b) as opposed to using randomly initialized embeddings
+(Fig.~\ref{fig:multiscale}.a). Another important extension is the use of
+multi-scale similarities, which, in addition to considering short range
+interactions, also models wider point neighborhoods. Coupled with PCA-based
+initialization, this produces even more meaningful visualizations where
+clusters form interpretable structures. For instance, consider Fig.~\ref
+{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
+neurons, corresponding to their developmental origin, as discussed in
+Tasic \etal~\cite{Tasic2018}, while this division is less apparent when using
+PCA-based initialization alone in Fig.~\ref{fig:multiscale}.b.
 
-We also observed the important role of gene selection in crafting the reference
-embedding spaces. We found that when selecting an insufficient number of genes,
-the resulting visualizations display overly-fragmented clusters. When the
-selection is too broad and includes lowly expressed genes, the subclusters tend
-to overlap. These effects can all be attributed to sparseness of the data sets
-and may be intrinsic to single-cell data. In our studies, we found that
-selection of 3,000 genes yields most informative visualizations
-(Fig.~\ref{fig:gene_selection}).
+We also observed the important role of gene selection in crafting the
+reference embedding spaces. We found that when selecting an insufficient
+number of genes, the resulting visualizations display overly-fragmented
+clusters. When the selection is too broad and includes lowly expressed genes,
+the subclusters tend to overlap. These effects can all be attributed to
+sparseness of the data sets and may be intrinsic to single-cell data. In our
+studies, we found that selection of 3,000 genes yields most informative
+visualizations(Fig.~\ref{fig:gene_selection}).
 
@@
 Fig.~\ref{fig:optimization} demonstrates a case where nearest neighbor-based
-positioning alone is insufficient. We construct a reference embedding using only
-neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a)
-and use that to position neuronal
-cells from the data set from Campbell \etal~\cite{Campbell2017}. We utilize the
-weighted mean and median positions to initialize point positions from the secondary
-data set, as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c.
-After initialization, we optimize point positions using the procedure described
-above for 500 iterations. The resulting visualizations from both initializations
-are visually very similar, indicating stable convergence. We show one of the resulting
-visualizations in Fig.~\ref{fig:optimization}.b.
+positioning alone is insufficient. We construct a reference embedding using
+only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref
+{fig:optimization}.a) and use that to position neuronal cells from the data
+set from Campbell \etal~\cite{Campbell2017}. We utilize the weighted mean and
+median positions to initialize point positions from the secondary data set,
+as shown in Figs.~\ref{fig:optimization}.b and \ref
+{fig:optimization}.c. After initialization, we optimize point positions using
+the procedure described above for 500 iterations. The resulting
+visualizations from both initializations are visually very similar,
+indicating stable convergence. We show one of the resulting visualizations in
+Fig.~\ref{fig:optimization}.b.
 
-Notice that both neighbor-based initialization schemes generally position data points
-such that their classification is unclear. Median-based initialization produces
-a sort of grid-like structure, while median based initialization positions the points
-almost continuously across the embedding space. Optimization reveals strong
-correspondence of several points to reference-defined clusters, while other points
-from the secondary data set are pushed away from their initial clusters, possibly
-indicating dissimilarity.
+Notice that both neighbor-based initialization schemes generally position data
+points such that their classification is unclear. Median-based initialization
+produces a sort of grid-like structure, while median based initialization
+positions the points almost continuously across the embedding space.
+Optimization reveals strong correspondence of several points to
+reference-defined clusters, while other points from the secondary data set
+are pushed away from their initial clusters, possibly indicating
+dissimilarity.
 
@@
   \includegraphics[width=\textwidth]{optimization.pdf}
-  \caption{Comparison of different initialization schemes for positioning new data
-  points onto reference embeddings. {\bf (a)} We construct a reference embedding
-  using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)}
-  We position neuronal cells from Campbell \etal~\cite{Campbell2017} using the
-  median initizaliation scheme from Kobak and Berens~\cite{Kobak2019} and
-  run optimization for 500 iterations. Compare the optimized embedding with the 
-  initial median initialization {\bf (c)} or by using a simple weighted mean
+  \caption{Comparison of different initialization schemes for positioning new
+   data points onto reference embeddings. {\bf (a)} We construct a reference
+   embedding using only neuronal subtypes from Hrvatin \etal~\cite
+   {Hrvatin2018}. {\bf (b)} We position neuronal cells from
+   Campbell \etal~\cite{Campbell2017} using the median initialization scheme
+   from Kobak and Berens~\cite{Kobak2019} and run optimization for 500
+   iterations. Compare the optimized embedding with the initial median
+   initialization {\bf (c)} or by using a simple weighted mean
    initialization {\bf (d)}.}
@@
 
-\subsection{Our approach requires a complete reference set}
+\subsection{On requirement of a complete reference set}
 
-Our approach assumes that all cell types from the secondary data set are present
-in the reference. Intuitively, using t-SNE in such a way is conceptually similar
-to classification via $k$-nearest neighbor classifiers and is similarily
-limited. The method may fail to reveal unseen cell types in the secondary data
-set, likely positioning them arbitrarily close to unrelated clusters. In some
-instances, unknown cell types may be sufficiently different from the reference
-data that t-SNE will repel them from existing clusters. However, we caution that
-this approach is unreliable and depends heavily on the chosen preprocessing
-pipeline.
+Our approach assumes that all cell types from the secondary data set are
+present in the reference. Intuitively, using t-SNE in such a way is
+conceptually similar to classification via $k$-nearest neighbor classifiers
+and is similarly limited. The method may fail to reveal unseen cell types in
+the secondary data set, likely positioning them arbitrarily close to
+unrelated clusters. In some instances, unknown cell types may be sufficiently
+different from the reference data that t-SNE will repel them from existing
+clusters. However, we caution that this approach is unreliable and depends
+heavily on the chosen preprocessing pipeline.
 
@@
 \etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
-\etal~\cite{Campbell2017} and add them to the reference embedding in
-Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal
-are scattered somewhat arbitrarily around several clusters in the reference
-embedding. Interestingly, the secondary data points form a ``ring'' around one
-of the clusters, indicating that these data points are very different from the
-cells in this cluster. Notice also that the points from the secondary data set
-exhibit little to no clustering and the different cell types seem to be mixed
-amongst each other. We hypothesize that this effect is due primarily to the
-single-cell preprocessing pipeline and not the limitations of our proceedure
-itself, as the informative genes selected to create the reference neuronal
-embedding likely do not differentiate supportive glial cells from the secondary
-data set. This effect is similar to procedures such as scMap-Cluster, a consensus KNN
-method, which heuristically determines a distance threshold to identify unknown
-cell types~\cite{Kiselev2018}.
+\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref
+{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
+scattered somewhat arbitrarily around several clusters in the reference
+embedding. Interestingly, the secondary data points form a ``ring'' around
+one of the clusters, indicating that these data points are very different
+from the cells in this cluster. Notice also that the points from the
+secondary data set exhibit little to no clustering and the different cell
+types seem to be mixed among each other. We hypothesize that this effect is
+due primarily to the single-cell preprocessing pipeline and not the
+limitations of our procedure itself, as the informative genes selected to
+create the reference neuronal embedding likely do not differentiate
+supportive glial cells from the secondary data set. This effect is similar
+to procedures such as scMap-Cluster, a consensus $k$-nearest neighbor 
+method, which heuristically determines a distance threshold to identify
+unknown cell types~\cite{Kiselev2018}.
 
@@
   \includegraphics[width=\textwidth]{incomplete_reference-brain.pdf}
-  \caption{A reference embedding must contain all the cell-types in the
-  secondary embedding to produce reliable results. {\bf (a)} We construct a
-  reference embedding containing only neuronal cells from Hrvatin
-  \etal~\cite{Hrvatin2018}. {\bf (b)} We select only non-neuronal cells from
-  Campbell \etal~\cite{Campbell2017} so that no overlap exists between the
-  cell-types between the data sets. Conceptually, t-SNE behaves similarily to a
-  KNN classifier, and places the non-neurnal cells to their the most similar
-  points in the reference. In some instances, the non-neuronal cells are
-  sufficiently different from the neuronal cells so that they are repelled from
-  the reference clusters. Such behaviour results in the ``ring'' seen on the
-  right hand side of the embedding.} \label{fig:incomplete_reference}
+  \caption{A reference embedding must contain all the cell types in the secondary embedding to produce reliable results.
+{\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite {Hrvatin2018}.
+{\bf (b)} We select only non-neuronal cells from Campbell \etal~\cite{Campbell2017} so that no overlap exists between the cell types between the data sets.
+Conceptually, t-SNE behaves similarly to a $k$-NN classifier and places the non-neuronal cells to their most similar points in the reference.
+In some instances, the non-neuronal cells are sufficiently different from the neuronal cells so that they are repelled from the reference clusters.
+Such behavior results in the ``ring'' seen on the right-hand side of the embedding.} \label{fig:incomplete_reference}
 \end{figure}
@@
 features. The results indicate that positioning of new cells onto a cell
-visualisation plane is not only indicative but also an accurate instrument for
+visualization plane is not only indicative but also an accurate instrument for
 cell type characterization.
@@
 normalization, log-transformation, and select 1,000 most informative genes.
-Similarily to scMap-Cluster, we use the cosine distance to find the 5 nearest
+Similarly to scMap-Cluster, we use the cosine distance to find the 5 nearest
 neighbors in the KNN model. We used 100 trees in the random forest ensemble. The
@@
 analysis. Therefore, it is surprising that our simple procedure performs
-competitievly to specialized classification methods. Therefore, our procedure,
+competitively to specialized classification methods. Therefore, our procedure,
 in addition to providing the end-user with a cell-type prediction, allows the
@@
 efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A
-common occurence in these data sets is ``dropout'', where an expressed gene is
+common occurrence in these data sets is ``dropout'', where an expressed gene is
 not measured, and its corresponding matrix entry is set to zero. To address this
@@
    mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random
-  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
-  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
-  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
-  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
-  {\bf (c)} Augmenting t-SNE with multi-scale
-  similarities and using proper initialization provides a more meaningful
-  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
-  the same region of the embedding. There are two clear sub-groups of GABAergic
-  neurons corresponding to their developmental origins, which was not as apparent
-  when using clever initialization alone.}
+   initialization places clusters arbitrarily. The resulting clustering
+   structure is not globally consistent, as clusters of the same type of
+   cells are dispersed throughout the landscape. Non-Neuronal clusters, for
+   instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
+   {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the
+   clusters are organized in a more meaningful layout, where clusters of
+   cells of the same type appear closer together.{\bf (c)} Augmenting t-SNE
+   with multi-scale similarities and using proper initialization provides a
+   more meaningful layout of the clusters. Non-Neuronal and Endothelial cell
+   types are now placed in the same region of the embedding. There are two
+   clear sub-groups of GABAergic neurons corresponding to their developmental
+   origins, which was not as apparent when using clever initialization
+   alone.}
   \label{fig:multiscale}

### 2021-06-18 paper-revision/response-to-reviewers.tex
revision

@@
 
-\noindent{\bf Poli\v{c}ar, Stra\v{z}ar \& Zupan: Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
+\noindent{\bf Poli\v{c}ar, Stra\v{z}ar \& Zupan: 
+\\
+Embedding to Reference t-SNE Space Addresses Batch Effects in Single-Cell Classification}
 
@@
 
-\noindent we thank for the comprehensive feedback that helped to improve the
+\noindent we thank for the comprehensive feedback that helped us to improve the
 manuscript, sufficient time to implement the changes, and the permission to
-submit a revised version.
+submit a revised version. Below, we provide the detailed response to reviewers. Please note that the submission also includes a version of the manuscript with highlighted changes.
 
@@
 
-All of the reviewer remarks are addressed below with reference to the
-corresponding changes in the manuscript.  In case of overlapping remarks from
-different reviewers, we refer the reader to the relevant enumerated answer. A
-differential document is provided to facilitate the comparison with the initial
-submission. The changes to the data, code and the instructions to reproduce all
-the experiments are available in the repository:
-\url{https://github.com/biolab/tsne-embedding}.
-
+\noindent Kind regards,
 \vspace{3mm}
 
-\begin{flushright}
-\noindent Kind regards, \\ 
-\vspace{2mm}
-Pavlin Poli\v{c}ar \\
-Martin Stra\v{z}ar \\
+\noindent Pavlin Poli\v{c}ar, 
+Martin Stra\v{z}ar, 
 Bla\v{z} Zupan \\ 
-\vspace{2mm}
-University of Ljubljana, Slovenia 
-\end{flushright}
+\noindent Ljubljana, June 18, 2021
 
@@
 
-\reviewc{r1q1}{Most important is to add in theoretical part investigation about which dataset can be a reference dataset. It is necessary because it is crucial in estimation of level of trust worth.
+\reviewc{r1q1}{Most important is to add in theoretical part investigation about which dataset can be a reference dataset.
+It is necessary because it is crucial in estimation of level of trust worth.
 Some information about it is provided in testing part but only in form of general statement on specific case (based on some data sets), but it is necessary to have a procedure/test when given data set is a good reference data set.}
@@
 \answer{
-	We thank the reviewer for this comment and have expanded our discussion of potential failure cases in the ``Experiments and Discussion'' section.
-We have added an example of a failure case and have pointed out that care must be taken when using our procedure.
-
-	A theoretical investigation of which data set can be used as a reference is beyond the scope of this paper.
-Conceptually, our approach for classification is very similar to a $k$-nearest neighbor classifier.
-It would be very difficult to theoretically quantify which reference data sets would be appropriate for such classifiers.
+We thank the reviewer for this comment.
+We have expanded our discussion with section 4.5, which highlights the need for a complete reference set.
+We also include an example of a failure case (Fig.6) where the reference data set violates this requirement.
 }
@@
 
-\answer{We thank the reviewer for catching this mistake. We have fixed these citations.}
+\answer{Thank you, we have fixed the error.}
 
@@
 \answer{
-	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough comparison. We moved this section to the appendix.
+These are parameters of the t-SNE method, which was proposed in related work.
+We note that $L$ and the perplexities are user-specified parameters and overview the procedure for selection of $\sigma_{i,l}$ in the text (Sec.
+3.1).
 }
@@
 
-\answer{We agree that this subsection could be more clear. We have modified some parts of it and added sentences which should clarify the gene selection procedure.}
+\answer{As suggested, we have improved the writing in Sec. 3.3. This background information on preprocessing of gene expression data is domain-specific and was hence moved to appendix.}
 
@@
 
-\answer{}
+\answer{Thank you for this suggestion.
+We have experimented with a different image layout and found that vertical alignment due to very large layout reduces readability of the text.
+We would kindly ask to leave this issue to the technical editor.}
 
@@
 
-\answer{}
+\answer{Thank you, all the figure are now rendered in higher quality.}
 
@@
 \answer{
-	This isn't part of our work, and has been moved to the supplement. Refer to the original publication for a thorough compoarison.
+Multiscale kernels were introduced and studied by Kobak and Berens.
+With Fig.
+5, we point that our pipeline can benefit from their approach as it produces a more coherent reference.
+Please note that our work focuses on adding new points into existing visualizations, regardless of how they were produced.
+Hence, we did not include further studies of creating reference visualizations as proposed by the reviewer.
 }
@@
 \answer{
-	How to select the number of genes is pretty irrelevant to our work. This is a question of single-cell processing pipeline, which we do not care about in this work.
+Globally, (b) and (c) are indeed similar, but there is a substantial difference in the structure of points describing neuronal cells.
+To clarify this point, we have addd ``In our example, the neuronal subclusters are more clearly defined in (b).'' to the caption of the figure.
 }
@@
 \answer{
-	We've addressed this in related work, where we've added a paragraph dedicated to batch effects.
+We added a paragraph on batch effects to the section on related work.
 }
@@
 
-\answer{We thank the reviewer for this comment. We have restructured the paper.
-
+\answer{We thank the reviewer for this comment.
+We have restructured the paper accordingly.
 We have added a ``Related work'' section, where we describe batch-effects in more detail.
 We have also moved the description of scMap-Cluster from the ``Experiments'' section to this section.
-While it is true that these approaches are, in essence, transfer learning, the data sets produced by single-cell RNA-seq are sufficiently different from data sets from other fields of machine learning that they requires specialized methods, which we describe in the ``Related work'' section.
+A relation with transfer learning is now addressed in Related work (second item in the second paragraph).
 }
@@
 \answer{
-	We added a paragraph to the end of the introduction explicitly listing our contributions.
+We added a paragraph to the end of the introduction explicitly listing our contributions of the paper.
 The differences from the conference version are mainly an improved experimental section, where we show that a naive placement of the data points leads to misleading conclusions, and that optimization produces more reasonable results.
@@
 \answer{
-	We have changed the working in ``Availability and Implementation'' section, which emphasizes that the extension we describe in this paper was developed by us and later incorporated into \textsf{openTSNE}.
+We have altered the wording in ``Availability and Implementation'' section to emphasize that the extension we describe in this paper was developed by us and later incorporated into \textsf{openTSNE}.
 As this is the only mention of \textsf{openTSNE} in the manuscript, we believe there can be no further confusion in this respect.
@@
 \answer{
-	This isn't part of our work, and has been moved to the supplement.
-Please, refer to the original publication for a thorough comparison.
-However, we would also refer the reviewer to a recent comment by Kobak and Linderman ``Initialization is critical for preserving global data structure in both t-SNE and UMAP'', \textit{Nature Biotechnology}, 2021, which demonstrates that good initialization schemes are important for both t-SNE and UMAP.
+PCA fails to produce cluster-informative visualizations of any high-dimensional data, as it focuses on variance instead of clustering structure.
+Comparison of these two methods have been already and extensively published (see, for instance, XXX).
 }
@@
 \answer{
-	This is a question of single-cell processing pipeline, which is not the focus of our work. We provide a reference to the original publication presenting the pipeline in the text.
-}
+The single-cell processing pipeline we use is standard, and we have provided references to original proposals of the methods.
+We include the description of preprocessing pipeline in the paper for the reasons of reproducibility.
+As data preprocessing is not the focus of our work, we have moved the description to the Appendix.}
 
@@
 \answer{
-	We agree that the original manuscript included sections that are important to the single-cell analysis pipeline which are perhaps not the most relevant to our particular contribution i.e.
-the extension of t-SNE to allow embedding new data points into existing embeddings.
-For this particular task, the details of the single-cell analysis are not crucial.
-While it is true that our single-cell pipeline relies on the work from [15], we have moved most of these sections to the appendix, as they are not directly related to our work.
-We believe that this improves the readability of our paper.
-However, the work done in [15] introduced ideas which are now considered standard practice for t-SNE visualizations, so a brief review of the most important concepts is kept in Section 3.1.
-
-	Regarding the contribution in terms of theoretical justification: the extension is indeed very minor, and requires changing only two variables, which results in different interactions being calculated.
-However, additional minor but important changes need to made, including the omission of the symmetrization step and a minor change in the gradient calculation.
-
-	Perhaps more important is the practical aspect of our work i.e.
-how do we get this to work.
-While the equations change only slightly, a straightforward application of our approach using the currently accepted t-SNE parameter settings will result in wildly unstable optimization, so special care must be taken to mitigate this erratic behaviour.
-We describe this in the last paragraph of Section 3.2, where we emphasize the importance of lowering the learning rate and introducing optimization tricks e.g.
-gradient clipping.
-This is especially important for the approximation schemes used today by virtually all t-SNE implementations, as a poor parameter setting will result in points ``shooting-off'' far away from the reference points and will dramatically impact the runtime of these approximation schemes.
-Without these modifications, the t-SNE transform functionality becomes unusable in virtually all settings.
+Our work extends the original t-SNE approach and uses a bag of tricks by Kobak and Berens.
+The reviewer is right: our formulation of the loss function for new data points is (intentionally) very similar to that of original t-SNE's.
+Our main contributions are the first demonstration that such an approach can work, produces excellent results, and even removes the batch effects.
+We expose these novelties in the new paragraph at the end of the introduction and highlight the importance of steps in our approach in the last paragraph of Sec.3.2.
 }
@@
 \answer{
-	We have slightly reworded the abstract to indicate that the secondary data set contains new, unseen points.
-
-	We disagree that mentioning more qualitative results would improve this manuscript.
-There are no robust, agreed-up baselines in the single-cell literature for batch effect removal in general.
-Only a handful of methods for cross-data set cell-type classification methods have been proposed, and none of them are directly related to our approach.
-This manuscript highlights that a simple extension to a visualization technique, t-SNE, performs competitively to specialized machine learning and bioinformatics methods for cell-type classification.
-These methods are not directly comparable, and our aim is not to improve the accuracy of such approaches.
-We have, however, reworded the last paragraph of Section 4.6 to emphasize this point.
+We have reworded the abstract to indicate that the secondary data set contains new, unseen points.
+We agree with a reviewer that quantifying the results is important.
+Hence, the manuscript includes a comparison with other techniques (Table 2).
+We, however, refrain from reporting the accuracy estimates in the abstract, as the focus of our work is a visualization and not optimization of classification accuracy.
 }
@@
 \answer{
-	We thank the reviewer for this comment and we agree that splitting the section into subsections would improve the readability of the manuscript.
-We have split up the ``Results and Discussion'' section into several subsections.
+We thank the reviewer for this comment. We have split up the ``Results and Discussion'' section into several subsections.
 }
@@
 \answer{
-	The legends of Figure 1 show the names of the data sets, or rather, their original publication.
-The figure demonstrates that two data sets with shared cell types cluster primarily by data set and not by cell-type.
-In this instance one would reasonably assume and even hope that the samples would cluster by cell-type and that the data set source would not interfere with the visualization.
-We feel that a description of the data sets is unnecessary here, as it does not help illustrate our point, and would needlessly burden the reader with biological details.
+We want to thank the reviewer for this comment.
+Both datasets contain cells as indicated in the panel title (brain, pancreas) and use a similar approach to gather the data.
+The main difference is in the group that performed the experiments and published the data.
+Hence, we use the first authors of the publications to distinguish between two batches of data.
 }
@@
 
-\answer{We agree that this is not the most suitable location for this information, and we have moved it to the end of the manuscript, after the conclusion and before the acknowledgments, as is standard practice in many journals.}
+\answer{We agree that a particular section is not the most appropriate location for this kind of information.
+We have moved the note on implementation to a more suitable place in the note before the acknowledgments.}
 
@@
 
-\answer{The thank the reviewer for the reminder. We have updated the citation.}
+\answer{The thank the reviewer for the reminder. We have updated the citation accordingly.}
 

### 2021-06-18 paper-revision/policar2020-tsne-embedding.tex
revision

@@
  combination of the two~\cite{Stuart2019,Hie2019}.
+Note that these approaches bear similarity with transfer learning~\cite{https://journalofbigdata.springeropen.com/articles/10.1186/s40537-016-0043-6}, which has also been used in domain adaptation~\cite{https://jieliu6.github.io/publication/conference-paper/jointly-embedding-multiple-single-cell/}.
 \end{enumerate}
@@
 
-\noindent where $L$ is the number of mixture components. The bandwidths
+\noindent where $L$ is the number of mixture components as specified by the user. The bandwidths
 $\sigma_{i,l}$ are selected in the same manner as in
@@
 
+\begin{figure}[htbp]
+  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
+  \caption{A comparison of standard and multi-scale t-SNE on data from the
+   mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random
+   initialization places clusters arbitrarily. The resulting clustering
+   structure is not globally consistent, as clusters of the same type of
+   cells are dispersed throughout the landscape. Non-Neuronal clusters, for
+   instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
+   {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the
+   clusters are organized in a more meaningful layout, where clusters of
+   cells of the same type appear closer together.{\bf (c)} Augmenting t-SNE
+   with multi-scale similarities and using proper initialization provides a
+   more meaningful layout of the clusters. Non-Neuronal and Endothelial cell
+   types are now placed in the same region of the embedding. There are two
+   clear sub-groups of GABAergic neurons corresponding to their developmental
+   origins, which was not as apparent when using clever initialization
+   alone.}
+  \label{fig:multiscale}
+\end{figure}
+
 We also observed the important role of gene selection in crafting the
@@
 studies, we found that selection of 3,000 genes yields most informative
-visualizations(Fig.~\ref{fig:gene_selection}).
+visualizations (Fig.~\ref{fig:gene_selection}).
+
+\begin{figure}[htbp]
+  \includegraphics[width=\textwidth]{hrvatin_embedding_tsne_genes.pdf}
+  \caption{Gene selection plays an important role when constructing the
+  reference embedding. {\bf (a)} Using too few genes results in fragmented
+  clusters. {\bf (b)} Using an intermediate number of genes reveals clustering
+  mostly consistent with cell annotations. {\bf (c)} Including all the genes
+  may lead to under-clustering of the more specialized cell types. In our example, the neuronal subclusters are more clearly defined in (b).}
+  \label{fig:gene_selection}
+\end{figure}
 
@@
 Due to the specific nature of single-cell data, additional steps must be taken
-to properly apply t-SNE. We use a standard single-cell preprocessing pipeline,
-consisting of the selection of 3,000 representative genes (see
-Appendix~\ref{sec:gene-selection}), library size normalization,
-log-transformation, standardization, and PCA-based representation that retains
-50 principal components~\cite{Stuart2019,Wolf2018}. To obtain the reference
-embedding, we apply multi-scale t-SNE using PCA initialization~\cite{Kobak2019}.
-Due to high-dimensionality of the preprocessed input data we use cosine distance
-to estimate similarities between reference data points~\cite{Domingos2012}. When
-adding new data points from the secondary data set to the reference embedding,
-we select 1,000 genes present in both data sets and use the cosine similarity to
+to properly apply t-SNE. We use a standard single-cell preprocessing
+pipeline, consisting of the selection of 3,000 representative genes
+(see Appendix~\ref{sec:gene-selection}), library size normalization,
+log-transformation, standardization, and PCA-based representation that
+retains 50 principal components~\cite{Stuart2019,Wolf2018}. To obtain the
+reference embedding, we apply multi-scale t-SNE using PCA
+initialization~\cite{Kobak2019}. Due to high-dimensionality of the
+preprocessed input data we use cosine distance to estimate similarities
+between reference data points~\cite{Domingos2012}. When adding new data
+points from the secondary data set to the reference embedding, we select
+1,000 genes present in both data sets and use the cosine similarity to
 estimate the similarities between the secondary data item and reference data
-points. We note that similarities are computed using the raw count matrices. The
-preprocessing stages are detailed in accompanying Python notebooks
+points. We note that similarities are computed using the raw count matrices.
+The preprocessing stages are detailed in accompanying Python notebooks
 (Sec.~\ref{sec:implementation}).
@@
 
-Single-cell data sets suffer from high levels of technical noise and low capture
-efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A
-common occurrence in these data sets is ``dropout'', where an expressed gene is
-not measured, and its corresponding matrix entry is set to zero. To address this
-problem, we use a specialized feature-selection method, which exploits the
-mean-dropout relationship of expression counts as recently proposed by Kobak and
-Berens~\cite{Kobak2019}. In this context, we will refer to all genes with matrix
-entries set to zero as dropouts. Intuitively, if a gene has high mean expression
-over all cells, but is detected in only a handful of them (i.e. has high mean
-dropout rate), then this gene is likely specific to a specific cell-type and
-will serve as a good feature for any subsequent analysis where we wish to
-discriminate between cell types.
+Single-cell data sets suffer from high levels of technical noise and low
+capture efficiency, resulting in sparse and noisy expression matrices~\cite
+{umi}. A common occurrence in these data sets is ``dropout'', where an
+expressed gene is not measured, and its corresponding matrix entry is set to
+zero. To address this problem, we use a specialized feature-selection method,
+which exploits the mean-dropout relationship of expression counts as recently
+proposed by Kobak and Berens~\cite{Kobak2019}. In this context, we will refer
+to all genes with matrix entries set to zero as dropouts. Intuitively, if a
+gene has high mean expression over all cells, but is detected in only a
+handful of them (i.e. has high mean dropout rate), then this gene is likely
+specific to a specific cell-type and will serve as a good feature for any
+subsequent analysis where we wish to discriminate between cell types.
 
@@
 
-\newpage
-
-\begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
-  \caption{A comparison of standard and multi-scale t-SNE on data from the
-   mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random
-   initialization places clusters arbitrarily. The resulting clustering
-   structure is not globally consistent, as clusters of the same type of
-   cells are dispersed throughout the landscape. Non-Neuronal clusters, for
-   instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
-   {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the
-   clusters are organized in a more meaningful layout, where clusters of
-   cells of the same type appear closer together.{\bf (c)} Augmenting t-SNE
-   with multi-scale similarities and using proper initialization provides a
-   more meaningful layout of the clusters. Non-Neuronal and Endothelial cell
-   types are now placed in the same region of the embedding. There are two
-   clear sub-groups of GABAergic neurons corresponding to their developmental
-   origins, which was not as apparent when using clever initialization
-   alone.}
-  \label{fig:multiscale}
-\end{figure}
-
-\begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{hrvatin_embedding_tsne_genes.pdf}
-  \caption{Gene selection plays an important role when constructing the
-  reference embedding. {\bf (a)} Using too few genes results in fragmented
-  clusters. {\bf (b)} Using an intermediate number of genes reveals clustering
-  mostly consistent with cell annotations. {\bf (c)} Including all the genes
-  may lead to under-clustering of the more specialized cell types.}
-  \label{fig:gene_selection}
-\end{figure}
-
 \end{document}

### 2021-09-05 paper-revision/response-to-reviewers.tex
Add final paper revision

@@
 PCA fails to produce cluster-informative visualizations of any high-dimensional data, as it focuses on variance instead of clustering structure.
-Comparison of these two methods have been already and extensively published (see, for instance, XXX).
+Comparison of these two methods have been already and extensively published (see, for instance, ``viSNE enables visualization of high dimensional single-cell data and reveals phenotypic heterogeneity of leukemia'' supplementary material).
 }

### 2021-09-05 paper-revision/diffbase.tex
Add final paper revision

@@
 profiles are similar. For this task, non-linear embedding techniques such as
-t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{tsne} or
-uniform manifold approximation and projection~\cite{umap} have recently
+t\nobreakdash -distributed stochastic neighbor embedding (t-SNE)~\cite{Maaten2008} or
+uniform manifold approximation and projection~\cite{McInnes2018} have recently
 complemented traditional data transformation and embedding approaches such as
@@
 proposed procedures for batch effect removal include canonical correlation
-analysis~\cite{cca} and mutual nearest-neighbors~\cite{mnn,seurat}.  In these
+analysis~\cite{Butler2018} and mutual nearest-neighbors~\cite{Haghverdi2018,Stuart2019}.  In these
 works, batch effects are deemed removed when cells from different sources
@@
 t-SNE visualization. While the utility of such an algorithm was already hinted
-at in recent publication~\cite{art_of_using_tsne}, we here provide its
+at in recent publication~\cite{Kobak2019}, we here provide its
 practical and theoretically-grounded implementation. Considering the abundance
@@
 enables the visualization of massive data sets, possibly containing millions of
-data points~\cite{fi_tsne}.
+data points~\cite{Linderman2019}.
 
@@
 small clusters and thus obscuring local aspects of the
-data~\cite{art_of_using_tsne}.
+data~\cite{Kobak2019}.
 
@@
 achieved by replacing the Gaussian kernels in Eq.~\ref{eq:gaussian_kernel} with
-a mixture of Gaussians of varying bandwidths~\cite{multiscale_tsne}.
+a mixture of Gaussians of varying bandwidths~\cite{Lee2015}.
 Multi-scale kernels are defined as
@@
 perplexity values of 50 and 500. A similar formulation of
-multi-scale kernels was proposed in~\cite{art_of_using_tsne}, and we found the
+multi-scale kernels was proposed in~\cite{Kobak2019}, and we found the
 resulting embeddings are visually very similar to those obtained with the
@@
 has been shown to lead to slower convergence and requires more iterations to
-achieve consistent embeddings~\cite{belkina2019automated}. We follow the
+achieve consistent embeddings~\cite{Belkina2019}. We follow the
 recommendation of Belkina \etal ~and use a higher learning rate $\eta = N / 12$
@@
 differences to embeddings like PCA or MDS, which have been substantially
-covered before~\cite{tsne,umap_single_cell}.  Below, we list the data sets,
+covered before~\cite{Maaten2008,Becht2019}.  Below, we list the data sets,
 describe single-cell specific data preprocessing procedures, and display the
@@
 \begin{description}
-\item[Mouse brain.] The data set from Hrvatin \etal~\cite{hrvatin2018} contains
+\item[Mouse brain.] The data set from Hrvatin \etal~\cite{Hrvatin2018} contains
 cells from the visual cortex exploring transcriptional changes after exposure
 to light. This was used as a reference for the data from Chen
-\etal~\cite{chen2017}, containing cells from the mouse hypothalamus and
+\etal~\cite{Chen2017}, containing cells from the mouse hypothalamus and
 their reaction to food deprivation. From the secondary data, we removed cells
@@
 
-\item[Human pancreas.] Baron \etal~\cite{baron2016} created
+\item[Human pancreas.] Baron \etal~\cite{Baron2016} created
 an atlas of pancreatic cell types. We used this set as a reference
-for data from Xin \etal~\cite{xin2016}, who examined transcriptional
+for data from Xin \etal~\cite{Xin2016}, who examined transcriptional
 differences between healthy and type 2 diabetic patients.
 
-\item[Mouse retina.] Macosko \etal~\cite{macosko2015} created
+\item[Mouse retina.] Macosko \etal~\cite{Macosko2015} created
 an atlas of mouse retinal cell types. We used this as a reference
-for the data from Shekhar \etal~\cite{shekhar2016}, who built an atlas for
+for the data from Shekhar \etal~\cite{Sekhar2016}, who built an atlas for
 retinal bipolar cells.
@@
 Kobak and Linderman have shown that the global consistency of embeddings produced by
-popular visualization algorithms are largely dependent on their initialization~\cite{kobak2019umap}. By utilizing PCA-based initialization, t-SNE is able to achieve
+popular visualization algorithms are largely dependent on their initialization~\cite{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
 more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b)
@@
 clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic neurons, corresponding to their
-developmental origin, as discussed in Tasic \etal~\cite{tasic2018}, while this division
+developmental origin, as discussed in Tasic \etal~\cite{Tasic2018}, while this division
 is less apparent when using PCA-based initialization alone in
 Fig.~\ref{fig:multiscale}.b.
+
+\begin{figure}[htbp]
+  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
+  \caption{A comparison of standard and multi-scale t-SNE on data from the
+  mouse neocortex~\cite{Tasic2018}. {\bf (a)} Standard t-SNE using random 
+  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
+  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
+  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
+  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
+  {\bf (c)} Augmenting t-SNE with multi-scale
+  similarities and using proper initialization provides a more meaningful
+  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
+  the same region of the embedding. There are two clear sub-groups of GABAergic
+  neurons corresponding to their developmental origins, which was not as apparent
+  when using clever initialization alone.}
+  \label{fig:multiscale}
+\end{figure}
 
@@
 
+\begin{figure}[htbp]
+  \includegraphics[width=\textwidth]{hrvatin_embedding_tsne_genes.pdf}
+  \caption{Gene selection plays an important role when constructing the
+  reference embedding. {\bf (a)} Using too few genes results in fragmented
+  clusters. {\bf (b)} Using an intermediate number of genes reveals clustering
+  mostly consistent with cell annotations. {\bf (c)} Including all the genes
+  may lead to under-clustering of the more specialized cell types.}
+  \label{fig:gene_selection}
+\end{figure}
+
 In principle, our theoretically-grounded embedding of secondary data into the
@@
 with such cases and identify the cells whose type is new and not included in
-the reference~\cite{scmap}.
+the reference~\cite{Kiselev2018}.
 
@@
 In practice, however, newly produced {\em secondary} data would be embedded into previously-prepared reference landscapes. Large collections of data {\em e.g.} the
-Human Cell Atlas initiative~\cite{hca} make it possible to scale up our approach to wider sets of cell types.
+Human Cell Atlas initiative~\cite{Rozenblatt-Rosen2017} make it possible to scale up our approach to wider sets of cell types.
 
@@
 
-We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{scmap}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data.
+We compare our approach to two machine learning techniques, namely a $k$-nearest neighbor classifier (KNN) and a random forest ensemble, and scMap-Cluster~\cite{Kiselev2018}, a $k$-nearest neighbor based method, tailored explicitly to scRNA-seq data.
 scMap-Cluster uses three correlation-based distance measures and uses a voting scheme to perform classification.
@@
 standardization, and PCA-based representation that retains 50 principal
-components~\cite{seurat,scanpy}. To obtain the reference embedding, we apply
-multi-scale t-SNE using PCA initialization~\cite{art_of_using_tsne}. Due to
+components~\cite{Stuart2019,Wolf2018}. To obtain the reference embedding, we apply
+multi-scale t-SNE using PCA initialization~\cite{Kobak2019}. Due to
 high-dimensionality of the preprocessed input data we use cosine distance to
-estimate similarities between reference data points~\cite{Domingos2012-CACM}.
+estimate similarities between reference data points~\cite{Domingos2012}.
 When adding new data points from the secondary data set to the reference
@@
 exploits the mean-dropout relationship of expression counts as recently
-proposed by Kobak and Berens~\cite{art_of_using_tsne}. Here, genes with higher
+proposed by Kobak and Berens~\cite{Kobak2019}. Here, genes with higher
 than expected dropout rate are regarded as potential markers for cell
@@
 
-\begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{tasic_multiscale.pdf}
-  \caption{A comparison of standard and multi-scale t-SNE on data from the
-  mouse neocortex~\cite{tasic2018}. {\bf (a)} Standard t-SNE using random 
-  initialization places clusters arbitrarily. The resulting clustering structure is not globally consistent,
-  as clusters of the same type of cells are dispersed throughout the landscape. Non-Neuronal clusters, for instance, are mixed with clusters of GABAergic and Glutamaergic neurons.
-  {\bf (b)} By utilizing a globally consistent initialization for t-SNE, the clusters
-  are organized in a more meaningful layout, where clusters of cells of the same type appear closer together.
-  {\bf (c)} Augmenting t-SNE with multi-scale
-  similarities and using proper initialization provides a more meaningful
-  layout of the clusters. Non-Neuronal and Endothelial cell types are now placed in
-  the same region of the embedding. There are two clear sub-groups of GABAergic
-  neurons corresponding to their developmental origins, which was not as apparent
-  when using clever initialization alone.}
-  \label{fig:multiscale}
-\end{figure}
-
-\begin{figure}[htbp]
-  \includegraphics[width=\textwidth]{hrvatin_embedding_tsne_genes.pdf}
-  \caption{Gene selection plays an important role when constructing the
-  reference embedding. {\bf (a)} Using too few genes results in fragmented
-  clusters. {\bf (b)} Using an intermediate number of genes reveals clustering
-  mostly consistent with cell annotations. {\bf (c)} Including all the genes
-  may lead to under-clustering of the more specialized cell types.}
-  \label{fig:gene_selection}
-\end{figure}
-
 \end{document}

### 2021-09-05 paper-revision/policar2020-tsne-embedding.tex
Add final paper revision

@@
 \usepackage{graphicx}
-\usepackage{amsmath, amssymb}
+\usepackage{amsmath,amssymb}
 \usepackage{multirow}
@@
 \begin{document}
+
 %
@@
 %
-\author{Pavlin G. Poli\v{c}ar\inst{1} \and
+\author{Pavlin G. Poli\v{c}ar\inst{1,}\thanks{Corresponding author} \and
 Martin Stra\v{z}ar\inst{1} \and
@@
 \item mutual nearest neighbor-based approaches aim to identify matching
- populations of cells across the data sets, using them to either find and
- correct the data sets~\cite{Haghverdi2018} or directly construct a
- batch-corrected k-nearest neighbor graph used in downstream analyses~\cite
- {Park2018}.
+ populations of cells across the data sets, using them to either find and correct the data sets~\cite{Haghverdi2018} or directly construct a
+ batch-corrected k-nearest neighbor graph used in downstream analyses~\cite{Park2018}.
 \item embedding multiple data sets into a joint lower-dimensional space, where
@@
  combination of the two~\cite{Stuart2019,Hie2019}.
-Note that these approaches bear similarity with transfer learning~\cite{https://journalofbigdata.springeropen.com/articles/10.1186/s40537-016-0043-6}, which has also been used in domain adaptation~\cite{https://jieliu6.github.io/publication/conference-paper/jointly-embedding-multiple-single-cell/}.
+Note that these approaches bear similarity with transfer learning~\cite{weiss2016survey}, which has also been used in domain adaptation~\cite{liu2019jointly}.
 \end{enumerate}
@@
 relied on the quality control and annotations from the original publication
-and report the number of cell types after prepossessing.  The cell
+and report the number of cell types after preprocessing.  The cell
 annotations were made consistent to annotations from the Cell
@@
 that the global consistency of embeddings produced by popular visualization
-algorithms are largely dependent on their initialization~\cite
-{Kobak2019a}. By utilizing PCA-based initialization, t-SNE is able to achieve
-more meaningful layouts of the resulting clusters (Fig.~\ref
-{fig:multiscale}.b) as opposed to using randomly initialized embeddings
-(Fig.~\ref{fig:multiscale}.a). Another important extension is the use of
+algorithms are largely dependent on their initialization~\cite{Kobak2019a}.
+By utilizing PCA-based initialization, t-SNE is able to achieve more meaningful layouts of the resulting clusters (Fig.~\ref{fig:multiscale}.b) as opposed to using randomly initialized embeddings (Fig.~\ref{fig:multiscale}.a).
+Another important extension is the use of
 multi-scale similarities, which, in addition to considering short range
@@
 initialization, this produces even more meaningful visualizations where
-clusters form interpretable structures. For instance, consider Fig.~\ref
-{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
+clusters form interpretable structures. For instance, consider Fig.~\ref{fig:multiscale}.c, which reveals two meaningful subgroups of GABAergic
 neurons, corresponding to their developmental origin, as discussed in
@@
 positioning alone is insufficient. We construct a reference embedding using
-only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref
-{fig:optimization}.a) and use that to position neuronal cells from the data
+only neurons from Hrvatin \etal~\cite{Hrvatin2018} (Fig.~\ref{fig:optimization}.a) and use that to position neuronal cells from the data
 set from Campbell \etal~\cite{Campbell2017}. We utilize the weighted mean and
 median positions to initialize point positions from the secondary data set,
-as shown in Figs.~\ref{fig:optimization}.b and \ref
-{fig:optimization}.c. After initialization, we optimize point positions using
+as shown in Figs.~\ref{fig:optimization}.b and \ref{fig:optimization}.c. After initialization, we optimize point positions using
 the procedure described above for 500 iterations. The resulting
@@
    data points onto reference embeddings. {\bf (a)} We construct a reference
-   embedding using only neuronal subtypes from Hrvatin \etal~\cite
-   {Hrvatin2018}. {\bf (b)} We position neuronal cells from
+   embedding using only neuronal subtypes from Hrvatin \etal~\cite{Hrvatin2018}. {\bf (b)} We position neuronal cells from
    Campbell \etal~\cite{Campbell2017} using the median initialization scheme
@@
 \etal~\cite{Hrvatin2018}. We then select only non-neuronal cells from Campbell
-\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref
-{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
+\etal~\cite{Campbell2017} and add them to the reference embedding in Fig.~\ref{fig:incomplete_reference}.b. The non-neuronal cells from Hrvatin \etal are
 scattered somewhat arbitrarily around several clusters in the reference
@@
   \caption{A reference embedding must contain all the cell types in the secondary embedding to produce reliable results.
-{\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite {Hrvatin2018}.
+{\bf (a)} We construct a reference embedding containing only neuronal cells from Hrvatin \etal~\cite{Hrvatin2018}.
 {\bf (b)} We select only non-neuronal cells from Campbell \etal~\cite{Campbell2017} so that no overlap exists between the cell types between the data sets.
@@
 Single-cell data sets suffer from high levels of technical noise and low
-capture efficiency, resulting in sparse and noisy expression matrices~\cite
-{umi}. A common occurrence in these data sets is ``dropout'', where an
+capture efficiency, resulting in sparse and noisy expression matrices~\cite{umi}. A common occurrence in these data sets is ``dropout'', where an
 expressed gene is not measured, and its corresponding matrix entry is set to
